### 2023-06-20 paper/paper.tex
Add TeX sources, and make initial version of Nix build

\documentclass[acmsmall,review]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{cmll}
\usepackage{mathpartir}

\newcommand{\tmRec}{\mathrm{rec}}
\newcommand{\tyNat}{\mathrm{Nat}}
\newcommand{\conZero}{\mathsf{zero}}
\newcommand{\conSucc}{\mathsf{succ}}
\newcommand{\conRefl}{\mathsf{refl}}

\newcommand{\Let}{\mathrm{let}}
\newcommand{\LetPair}{\mathrm{letpair}}
\newcommand{\In}{\mathrm{in}}
\newcommand{\If}{\mathrm{if}}
\newcommand{\dupNat}{\mathrm{dupNat}}

\newcommand{\istype}{\mathrm{type}}
\newcommand{\isctxt}{\mathrm{ctxt}}

\newcommand{\natinf}{\mathbb{N}_{-\infty}}

\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\RTm}{\mathrm{RTm}}
\newcommand{\Tm}{\mathrm{Tm}}

\newcommand{\Set}{\mathrm{Set}}

\newcommand{\cat}[1]{\mathcal{#1}}
\newcommand{\op}{\mathsf{op}}

\newcommand{\LinPoset}{\mathrm{LinPoset}}

\newcommand{\bob}[1]{\textcolor{purple}{FIXME: #1}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2020}
\acmYear{2020}
\acmDOI{10.1145/1122445.1122456}

%%
%% These commands are for a JOURNAL article.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF}
\acmArticle{1}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Implicit Polynomial Time and Dependent Types}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Robert Atkey}
\email{robert.atkey@strath.ac.uk}
\orcid{1234-5678-9012}
\affiliation{%
  \institution{University of Strathclyde}
  \streetaddress{26 Richmond Street}
  \city{Glasgow}
  \country{UK}
  \postcode{G1 1XH}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types for reasoning about programs with linear
  type systems for implicit polynomial time.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{type theory, implicit computational complexity, linear types}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}

The story:
\begin{enumerate}
\item There are systems for implicit computational complexity based on
  linear logic (e.g. LFPL, SAL, LLL, BLL) that seek to capture
  complexity classes (most often polynomial time) implicitly
  (i.e. without directly talking about time bounds). This work has
  also led to followup work on explicitly resourced calculi such as
  RAML, which seek to use static analysis to determine complexity
  bounds on programs.
\item In dependent type theory, we can reason about the extensional
  behaviour of programs (within the limits of intensional type
  theory), but not about intenstional properties. The only way to
  formalise something like ``for all polynomial time functions
  $\mathbb{N} \to \mathbb{N}$'' is to develop a deep emebdding of
  polynomial time functions and to work with that, which is tedious.
\item The goal of this paper is to unify dependent types for reasoning
  about programs with implicit computational complexity type
  systems. In the first instance, we do not seek to use dependent
  types to state precise bounds on programs' time consumption, but we
  intend to use dependent types to reason about the extensional
  behaviour of programs that we know to be polynomial time. At the end
  of the paper, we show how to adapt our framework to explicit
  resource accounting, in the style of RAML.
\item The plan of the paper is as follows:
  \begin{enumerate}
  \item We first introduction the ideas behind the use of linear logic
    for capturing polynomial time, by studying two systems Lafont's
    SAL, and Hofmann's LFPL. In Hofmann and Dal Lago's presentation,
    these are two variants of Second order Multiplicative Linear
    Logic, with restricted $\oc$ modalities for duplication. We also
    recall the realisability semantics of Dal Lago and Hofmann (FIXME:
    move this later?)
  \item The most modern presentations of these systems use second
    order quantification to represent data types. This is incompatible
    with the form of dependent types we use here, so we seek a direct
    representation of natural numbers that captures
  \item We then recall QTT, a extension of Martin-Löf type theory with
    linear types, which can accomodate intensional restrictions on
    computability as well as full reasoning about programs. Since
    allowing an unrestricted $\oc$ modality would violate our
    polynomial time bounds, we instead choose the natural number
    semiring for resource accounting.
  \item We then transplant the simply typed rules into QTT, showing
    how to carefully distinguish between data that is available to the
    program, and data that is only available for type formation.
  \item We finish by showing that we can also do explicit resource
    tracking.
  \end{enumerate}
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear $\lambda$-Calculus and Implicit Computational
  Complexity}
\label{sec:linear-icc}

Introduce the idea of linear typing for complexity control.

Linear $\lambda$-calculus, as we shall be discussing it in this
section:
\begin{mathpar}
  \inferrule* [right=Var]
  { }
  {x : A \vdash x : A}

  \inferrule* [right=Lam]
  {\Gamma, x : A \vdash M : B}
  {\Gamma \vdash \lambda x.M : A \multimap B}

  \inferrule* [right=App]
  {\Gamma_1 \vdash M : A \multimap B \\
    \Gamma_2 \vdash N : A}
  {\Gamma_1, \Gamma_2 \vdash M\,N : B}
\end{mathpar}

\begin{mathpar}
  \inferrule* [right=$\otimes$-Intro]
  {\Gamma_1 \vdash M : A \\ \Gamma_2 \vdash N : B}
  {\Gamma_1, \Gamma_2 \vdash (M,N) : A \otimes B}

  \inferrule* [right=$\otimes$-Elim]
  {\Gamma_1 \vdash M : A \otimes B \\ \Gamma_2, x : A, y : B \vdash N : C}
  {\Gamma_1, \Gamma_2 \vdash \Let\,(x,y)=M\,\In\,N : C}
\end{mathpar}

\subsection{Excluding Recursion}

What do we get if we exclude recursion altogether?

What happens if we include natural number recursion naively?

\paragraph{A Binary Tape Datatype}

FIXME: how to do this?

move left, move right, update. Tape is always finite.

\subsection{Cons-free programming}

FIXME: zero case has a context, or do we $\lambda$-abstract?
\begin{displaymath}
  \inferrule*
  {\vdash M_z : A \\ x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(x) \mapsto M_s\} : A}
\end{displaymath}

\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : \tyNat}
  {\Gamma \vdash \dupNat\,M : \tyNat \otimes \tyNat}
\end{displaymath}

Intuitive reason it works: we can only iterate over the natural
numbers we have recieved as input, there is no way to construct new
natural numbers by iteration. Duplicating natural numbers is safe,
becasue we can only do it a number of times controlled by iteration of
the natural numbers we are given.

FIXME: is it possible to pin down how it relates to SAL?

\paragraph{Completeness}

\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}

\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}

\subsection{Diamond Trading with LFPL}

The cons-free style of programming allows us to express all polytime
Turing machines, but is quite awkward from the point of view of
functional programming. It allows us to iterate over natural numbers
that come from the input but does not allow us to build further values
to do iteration on. For example, if our input is a list, then we
cannot transform it into a binary search tree and then flatten it
\bob{actually write this}.

Introduce a new type $\Diamond$. Each $\Diamond$ value represents a
``chunk of iterability''.

{\bf Rough model:} programs are polynomials, data are bags of
diamonds. Applying a program to some data yields a concrete number of
steps that will bound the resulting computation.

\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M : \Diamond}
  {\Gamma \vdash \conZero(M) : \tyNat}

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}

\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}

\paragraph{Other datatypes} Trees, and so on.

\paragraph{Completeness for Polytime} We now show how every polytime
Turing machine can be simulated in LFPL with a non-iterable type of
tapes. The construction, due to Aelhig and Schwictenberg
\cite{syntactic-lfpl}, illustrates how programming with diamonds in
LFPL differs from the soft system in the previous section.

Assume a state type $\mathit{St}$ and a step function $f : \mathit{St} \multimap \mathit{St}$.

The linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}

Step:\textbf{Is the extra $f\,s$ required here?}
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}

\begin{lemma}
  (Under a plausible operational semantics) FIXME: use an equational
  theory instead, that includes $\beta$-equality.
  \begin{displaymath}
    I_k\,(\underline{n},s) \longrightarrow^* (\underline{n},v)
  \end{displaymath}
  iff
  \begin{displaymath}
    f^{\binom{n}{k}}\,s \longrightarrow^* v
  \end{displaymath}
\end{lemma}

\subsection{Explicit Resource Tracking}

\begin{enumerate}
\item We still require diamonds to build zero and successor, but the
  iterator does not give them back.
\item In the simply typed case, we can annotate types with the static
  amount of potential they have per constructor.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Complexity Control in Quantitative Type Theory}
\label{sec:qtt}

We have now seen three systems for controlling resource usage by means
of linear typing. We now look to extend these systems to include
dependent types. Our motivation is threefold:
\begin{enumerate}
\item We would like to have a means for reasoning about the behaviour
  of the programs we write in our systems. FIXME: why dependent types
  then? Why not a program logic?
\item In case of the explicitly resource accounted system, simple
  typing is limited in the reasoning the programmer can perform about
  the resources required by a particular program. While we could in
  principle
\item As well as constructing proofs \emph{about} complexity
  constrained programs, we would also be able to write constructive
  proofs that are themselves of constrained complexity.
\end{enumerate}

To integrate linear typing for controlling time complexity with
dependent types we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt}. In this section, we review the definition of QTT
and how we have adapted it to implicit and explicit control of
computational complexity.

\begin{enumerate}
\item General description of QTT
\item Non-iterable datatypes in QTT
\item Adapting QTT to include a ``soft'' natural number type
\item Diamonds
\item
\end{enumerate}

\subsection{Quantitative Type Theory}

\begin{enumerate}
\item Rules for contexts (using $\epsilon$ for empty contexts to avoid
  confusion between $\diamond$ and $\Diamond$):
  \begin{mathpar}
    \inferrule* [right=Ctxt-Emp]
    { }
    {\epsilon~\isctxt}

    \inferrule* [right=Ctxt-Ext]
    {\Gamma~\isctxt \\ 0\Gamma \vdash S~\istype}
    {\Gamma, x \stackrel\rho: S~\isctxt}
  \end{mathpar}

  \begin{mathpar}
    \inferrule* [right=Tm-Var]
    {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
    {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}

    \inferrule* [right=Tm-Conv]
    {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
    {\Gamma \vdash M \stackrel\sigma: T}
  \end{mathpar}
\item Rules for $\Pi$ and $\Sigma$ types
  \begin{mathpar}
    \inferrule* [right=$\Pi$-type]
    {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
    {0\Gamma \vdash (x \stackrel\rho: S) \to T~\istype}

    \inferrule* [right=$\Pi$-Intro]
    {\Gamma, x \stackrel{\sigma\rho}: S \vdash M \stackrel\sigma: T}
    {\Gamma \vdash \lambda x.M \stackrel\sigma: (x \stackrel\rho: S) \to T}

    \inferrule* [right=$\Pi$-Elim]
    {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
      \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
      FIXME}
    {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
  \end{mathpar}

  FIXME: $\Sigma$-types
\item Rules for equality
  \begin{mathpar}
    \inferrule* [right=Eq-type]
    {0\Gamma \vdash S~\istype \\
      0\Gamma \vdash M \stackrel0: S \\
      0\Gamma \vdash N \stackrel0: S}
    {0\Gamma \vdash M =_S N~\istype}

    \inferrule* [right=Eq-Intro]
    {\Gamma \vdash M \stackrel\sigma: S}
    {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
  \end{mathpar}
  \bob{Equality reflection?}
\item Rules for universe
\end{enumerate}

\subsection{Non-iterable Datatypes}

\subsection{Cons-free Natural Numbers}

Of course, ``cons-free'' only applies to the runtime portion of the
system. In the $\sigma = 0$ fragment, we are allowed to create new
natural numbers arbitrarily:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash}
  {\Gamma \vdash \conZero \stackrel0: \tyNat}

  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
\end{mathpar}
Creating numbers in the $\sigma = 0$ fragment ensures that we can talk
about natural numbers in our types, allowing us to state properties of
our programs.\bob{Forward ref to actually doing this.}

The eliminator:
\begin{displaymath}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, m \stackrel0: \tyNat \vdash X~\istype \\\\
    0\Gamma \vdash M_z \stackrel\sigma: X[\conZero/m] \\\\
    0\Gamma, n \stackrel0: \tyNat, x \stackrel1: X[n/m] \vdash M_s \stackrel\sigma: X[\conSucc(n)/m] \\\\
    \Gamma \vdash N \stackrel\sigma: \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(n/x) \mapsto M_s\} \stackrel\sigma: X[N/m]}
\end{displaymath}

\subsection{Diamonds}

\begin{mathpar}
  \inferrule* [right=DiaType]
  {\Gamma \vdash}
  {0\Gamma \vdash \Diamond~\istype}

  \inferrule* [right=DiaIntro]
  {\Gamma \vdash}
  {0\Gamma \vdash * \stackrel0: \Diamond}

  \inferrule* [right=DiaEta]
  {\Gamma \vdash M \stackrel0: \Diamond}
  {\Gamma \vdash M \equiv * \stackrel0: \Diamond}
\end{mathpar}

Link to Selinger et al: the $\eta$ rule for $\Diamond$s means that are
essentially depending on the ``shape'' of anything that depends on
diamonds.

\subsection{LFPL-style Natural Numbers}

\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: \Diamond}
  {\Gamma \vdash \conZero(M) \stackrel\sigma: \tyNat}

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: \Diamond \\
    \Gamma_2 \vdash N \stackrel\sigma: \tyNat \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \conSucc(M,N) \stackrel\sigma: \tyNat}
\end{mathpar}

\begin{mathpar}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, m \stackrel0: \tyNat \vdash X~\istype \\\\
    0\Gamma, d \stackrel1: \Diamond \vdash M_z \stackrel\sigma: X[\conZero(*)/m] \\\\
    0\Gamma, d \stackrel1: \Diamond, n \stackrel0: \tyNat, x \stackrel1: X[n/m] \vdash M_s \stackrel\sigma : X[\conSucc(*,n)/m] \\\\
    \Gamma \vdash N \stackrel\sigma: \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,n/x) \mapsto M_s\} \stackrel\sigma: X[N/m]}
\end{mathpar}

Note: we have used $* : \Diamond$ as the value in the types for the
zero and successor cases. By the $\eta$-rule for diamonds, we could
have equally well used the $d$ variable that is in scope in each case.

\subsection{Explicit Potential}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Soundness via Realisability}
\label{sec:soundness}

\newcommand{\cstaccess}{c_{\mathit{access}}}
\newcommand{\cstmkclo}{c_{\mathit{mkclo}}}
\newcommand{\cstapp}{c_{\mathit{app}}}
\newcommand{\cstmkpair}{c_{\mathit{mkpair}}}
\newcommand{\cstmkunit}{c_{\mathit{mkunit}}}
\newcommand{\cstTrue}{c_{\mathit{mktrue}}}
\newcommand{\cstFalse}{c_{\mathit{mkfalse}}}
\newcommand{\cstLetpair}{c_{\mathit{letpair}}}
\newcommand{\cstSeq}{c_{\mathit{seq}}}
\newcommand{\cstIf}{c_{\mathit{if}}}

\newcommand{\clo}[2]{\mathsf{clo}\langle #1 , #2 \rangle}
\newcommand{\synTrue}{\mathsf{true}}
\newcommand{\synFalse}{\mathsf{false}}

\newcommand{\ExpSet}{\mathcal{E}}
\newcommand{\ValSet}{\mathcal{V}}

The Hofmann-Dal Lago approach to proving the soundness of implicit
computational complexity systems is based on a three way coupling
between abstract mathematical elements (the \emph{what}), values from
a machine model (the \emph{how}), and resource potentials (the
\emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Following Hofmann and Dal Lago, we
will select resource monoids appropriate to the kind of system that we
want to prove complexity soundness for.

\subsection{Machine Model and Operational Semantics}
\label{sec:machine-model}

\begin{figure}
  \centering
  {\bf Syntax}
  \begin{displaymath}
    \begin{array}{lcl}
      i,j &\in&\mathbb{N} \\
      E \in \ExpSet &::=& \lambda E \mid * \mid (i, j) \mid \synTrue \mid \synFalse \mid i \mid \Let\,E_1\,\In\,E_2 \mid i \cdot j \mid \LetPair\,i\,\In\,E \mid \If\,i\,E_1\,E_2 \\
      V \in \ValSet &::=& \clo{E}{\eta} \mid * \mid (V_1, V_2) \mid \synTrue \mid \synFalse \\
      \eta &::= & [] \mid \eta :: V
    \end{array}
  \end{displaymath}

  \vspace{1em}

  {\bf Evaluation: Construction}
  \begin{mathpar}
    \inferrule* [right=MkClo]
    { }
    {\lambda E , \eta \Downarrow_{\cstmkclo} \clo{E}{\eta}}

    \inferrule* [right=MkUnit]
    { }
    {*, \eta \Downarrow_{\cstmkunit} *}

    \inferrule* [right=MkPair]
    {\eta[i] = V_1 \\ \eta[j] = V_2}
    {(i, j), \eta \Downarrow_{\cstmkpair} (V_1, V_2)}

    \inferrule* [right=MkTrue]
    { }
    {\synTrue, \eta \Downarrow_{\cstTrue} \synTrue}

    \inferrule* [right=MkFalse]
    { }
    {\synFalse, \eta \Downarrow_{\cstFalse} \synFalse}
  \end{mathpar}

  \vspace{1em}

  {\bf Evaluation: Variable access and Sequencing}
  \begin{mathpar}
    \inferrule* [right=Access]
    {\eta[i] = v}
    {i, \eta \Downarrow_{\cstaccess} v}

    \inferrule* [right=Seq]
    {E_1, \eta \Downarrow_{k_1} V \\
      E_2, (\eta :: V) \Downarrow_{k_2} V'}
    {\Let\,E_1\,\In\,E_2, \eta \Downarrow_{k_1 + \cstSeq + k_2} V'}
  \end{mathpar}

  \vspace{1em}

  {\bf Evaluation: Elimination}
  \begin{mathpar}
    \inferrule* [right=App]
    {\eta[i] = \clo{E}{\eta'} \\
      \eta[j] = V \\
      E , (\eta' :: \clo{E}{\eta'} :: V) \Downarrow_k V'}
    {(i \cdot j), \eta \Downarrow_{\cstapp + k} V'}

    \inferrule* [right=LetPair]
    {\eta[i] = (V_1, V_2) \\
      E, (\eta :: V_1 :: V_2) \Downarrow_k V}
    {\LetPair\,i\,\In\,E, \eta \Downarrow_{\cstLetpair + k} V}

    \inferrule* [right=IfTrue]
    {\eta[i] = \synTrue \\ E_1, \eta \Downarrow_k V}
    {\If\,i\,E_1\,E_2, \eta \Downarrow_{\cstIf + k} V}

    \inferrule* [right=IfFalse]
    {\eta[i] = \synFalse \\ E_2, \eta \Downarrow_k V}
    {\If\,i\,E_1\,E_2, \eta \Downarrow_{\cstIf + k} V}
\end{mathpar}
  \caption{Language with Big-step Costed Evaluation Semantics}
  \label{fig:opsem}
\end{figure}

Note: the Agda formalisation is intrinsically well-scoped, which
makes expressing things a bit easier.

The syntax and rules of our target language are given in
\autoref{fig:opsem}.

\begin{enumerate}
\item Justification that looking up variables is constant: Due to
  lexical scoping, the size of the maximum variable lookup is bounded
  for a fixed program.
\item An advantage of our more ``elementary'' operational semantics
  (rather than encoding everything in the $\lambda$-calculus)
\end{enumerate}

\subsection{Resource Monoids}
\label{sec:resource-monoids}

Resource potentials are attached to values to represent the amount of
intrinsic potential they have to fuel computation. Resource potentials
are organised into resource monoids. To motivate the definition, we
first enumerate the structure we will. To be able to account for the
combined potential attached to composite data and programs (e.g.,
pairs, or functions applied to arguments) we will require monoid
structure on potentials. The action of turning potential difference
into fuel for computation will be modelled by a difference
function. Finally, we require that our resource monoid contains
sufficient elements to fuel constant time operations. We gather these
requirements into a formal definition as follows:

\begin{definition}
  A \emph{resource monoid} $M$ consists of:
  \begin{enumerate}
  \item A carrier set $|M|$, whose elements represent amounts of
    potential. We use Greek letters $\alpha$, $\beta$, $\gamma$ to
    denote elements of a resource monoid.
  \item Commutative monoid structure $(\oplus,\emptyset)$ on $|M|$, so
    we can add potentials.
  \item A \emph{difference function} $M : |M| \times |M| \to \natinf$,
    where $\natinf$ is the natural numbers extended with a negative
    infinity $- \infty$ and $- \infty + k = -\infty$. A difference
    $M(\alpha, \beta) = k \in \mathbb{N}$ means that starting with
    potential $\alpha$ and ending with potential $\beta$ yields $k$
    units of fuel. A difference of $- \infty$ means that $\alpha$
    contains insufficient potential to reach $\beta$. Differencing
    must satisfy:
    \begin{enumerate}
    \item for all $\alpha$, $M(\alpha,\alpha) = 0$; and
    \item for all $\alpha, \beta, \gamma$,
      $M(\alpha, \beta) + M(\beta, \gamma) \leq M(\alpha, \gamma)$.
    \end{enumerate}
    The latter is a ``reverse triangle inequality'': the fuel
    recoverable by moving between potential levels $\alpha$ and
    $\gamma$ via $\beta$ may be less than the fuel recoverable
    moving from $\alpha$ to $\gamma$ directly.
  \item An \emph{accounting function}
    $\mathit{acct} : \mathbb{N} \to |M|$ such that for all $k$,
    $k \leq M(\mathit{acct}(k),\emptyset)$.
  \end{enumerate}
\end{definition}

\subsubsection{Specific Resource Monoids}

The simplest example of a resource monoid is given by the natural
numbers $\mathbb{N}$, where each number stands directly an amount of
stored fuel.

\begin{definition}[Natural Number Resource Monoid]
  Monoid structure is given by normal addition. Differencing is
  defined as
  \begin{displaymath}
    \mathbb{N}(m,n) = \left\{
      \begin{array}{ll}
        m - n & m \geq n \\
        - \infty & \textrm{otherwise}
      \end{array}\right.
  \end{displaymath}
  and $\mathit{acct}(k) = k$. Note that this is the simplest possible
  resource monoid due to the requirement that the $\mathit{acct}$
  function exists. \bob{Forward ref to where this is used.}
\end{definition}

Every resource monoid induces a partial ordering on its carrier set by
$\alpha \leq \beta$ iff $0 \leq M(\alpha, \beta)$. In the case of the
natural number resource monoid, this yields the usual ordering on the
naturals.

\newcommand{\MaxPoly}{\mathrm{MaxPoly}}
\newcommand{\PlusPoly}{\mathrm{PlusPoly}}

The differencing operator of the natural number resource monoid can
only supply as much fuel as is contained in the potential. This makes
it suitable for amortised analysis, where we directly store potential
in data structures. For the two polynomial time systems, we need more
sophisticated structures originally presented by Dal Lago and
Hofmann. The fundamental idea with both is to represent potentials as
pairs $(m,p)$, where $m$ is a natural number and $p$ is a
polynomial. The $m$ tracks the ``size'' of data as it pertains to the
number of times an operation will be repeated by iterating over it ---
for example, an iterable natural number will have size equal to
itself, but a non-iterable natural number may be assigned zero
size. The polynomial $p$ tracks the complexity of a program as a
function of the size of the input. This leads to a differencing
operator that evaluates the polynomial with the size of the data:

\begin{definition}[Polynomial Resource Monoids]
  The \emph{Max-Polynomial} resource monoid $\MaxPoly$ has carrier
  set consisting of pairs $(m,p)$ where $m$ is natural number and $p$
  is a polynomial with natural number coefficients. Addition of
  elements is defined as $(m,p) \oplus (n,q) = (m \sqcup n, p+q)$ with
  $\emptyset = (0,0)$. Difference is defined as:
  \begin{displaymath}
    \MaxPoly((m,p),(n,q)) = \left\{
      \begin{array}{ll}
        p(m) - q(m)&m \geq n \textrm{ and } \forall k \geq m. p(k) \geq q(k) \\
        - \infty & \textrm{otherwise}
      \end{array}
    \right.
  \end{displaymath}
  $\MaxPoly$ accounts for constant time with constant polynomials:
  $\mathit{acct}(k) = (0,\lambda x.k)$.

  The \emph{Plus-Polynomial} resource monoid $\PlusPoly$ is defined
  the same way as $\MaxPoly$ except that the monoid addition adds the
  natural number components instead of taking their maximum:
  $(m,p) \oplus (n,q) = (m + n, p+q)$.
\end{definition}

It is perhaps easier to see how the differencing operator works in the
special case of the difference $\MaxPoly((m,p),(0,0)) = p(m)$. I.e.,
if we have code that contains data of size $m$ and a program with
complexity $p$, then running the combination with no expectation of
remaining potential yields $p(m)$ available steps.

The $\MaxPoly$ and $\PlusPoly$ resource monoids will be used for the
cons-free and LFPL-style (i.e., with diamonds) systems
respectively. We will elaborate on how in
\autoref{sec:realisability-model} and how they support the two kinds
of natural number iteration in \autoref{sec:realising-iteration}.

FIXME: difference between this definition and Hofmann-Dal Lago's: we
use polynomials with natural number coefficients instead of a function
that is bounded by a polynomial.

\subsubsection{Resource sub-monoids}

The separation between sizes of data and complexity of code in the
polynomial resource monoids motivates the use of resource sub-monoids
to ensure that programs themselves (as opposed to higher order code
which may contain closed over data) do not contain data that can be
iterated. We do this by requiring that programs' potential must come
from a specified resource sub-monoid:

\begin{definition}[Resource Sub-Monoids]
  A \emph{resource sub-monoid} $M_0 \subseteq M$ of a resource monoid
  $M$ consists of a subset $|M_0| \subseteq |M|$ that is closed under
  the monoid operations and $\mathit{acct}$.
\end{definition}

For both $\MaxPoly$ and $\PlusPoly$, the elements with zero size
component, i.e., of the form $(0,p)$, form a resource sub-monoid that
we will use for interpreting programs. We will call these sub-monoids
$\MaxPoly_0$ and $\PlusPoly_0$. We make use of resource sub-monoids
in our definitions of the interpretations of simultaneous
substitutions (\autoref{defn:rl-morphism}) and terms
(\autoref{defn:rl-terms}).

\subsubsection{Resource Monoids as Enriched Categories}

An alternative perspective on resource monoids is to see them as a
kind of entriched category \cite{kelly}. Enriched categories replace
the sets of morphisms between objects with objects from some category
other than $\Set$. Define the posetal category $\natinf$ with objects $\mathbb{N} \cup \{\infty\}$ and

$\natinf$ as a category.

\bob{remember to look at the Marsden and thingy paper about
  quantitative resources and enriched categories.}  A concise
definition of Hofmann and Dal Lago's resource monoids can be given as:
\begin{definition}
  A \emph{resource monoid} is a strict symmetric monoidal
  $\natinf$-enriched category.
\end{definition}

FIXME: Relational definition, used for the formalisation.

\subsection{Models of Quantitative Type Theory}
\label{sec:qtt-models}

\cite{atkey18} describes a general class of QTT models termed
\emph{Quantitative Categories with Families} (QCwFs). This class of
models fits will with the syntax of QTT, but to actually build models
it is revealing to construct them from certain indexed partial orders,
as we explain now. In the next section, we will construct a class of
specific models that prove the soundness of our complexity constrained
systems.

\begin{definition}[Category with Families]
  BLAH
\end{definition}

\begin{definition}
  A \emph{linear poset} is a partially ordered set $(A, \leq)$ with
  \begin{enumerate}
  \item A commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order; and
  \item Is closed: there is an operation
    $\multimap : A \times A \to A$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$.
  \end{enumerate}
  The collection of all linear posets and functions that preserve the
  order and the operations forms a category $\LinPoset$.
\end{definition}

\begin{definition}
  \begin{enumerate}
  \item A CwF $(\cat{C}, \top, \langle-,-\rangle, ...)$ that supports
    $\Pi$ and $\Sigma$ types.
  \item A functor $L : \cat{C}^\op \to \LinPoset$, such that
    reindexing along projections has a right adjoint:
    \begin{displaymath}
      L_{\Gamma.A}(\pi^*X, Y) \cong L_{\Gamma}(X, \forall_A Y)
    \end{displaymath}
  \end{enumerate}
\end{definition}

\subsection{Realisability Model}
\label{sec:realisability-model}

\bob{Rearrange this as an indexed linear poset}

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we now begin to construct
models of QTT that witness the resource (and type) soundness of our
complexity constrained systems.

\bigskip

\cite{atkey18qtt} describes a general class of QTT models termed
\emph{Quantitative Categories with Families} (QCwFs), an extension of
the \cite{dyber}'s \emph{Categories with Families} formulation of
models of dependent type theory. We will not regurgitate the full
definition here. Instead, we will define each of the parts of our
specialised QCwF for proving complexity soundness and explain
informally how to use them to model the syntax of QTT. Please refer to
\cite{atkey18qtt} for details of the interpretation.

Fix a resource monoid $M$ with a submonoid $M_0 \subseteq M$.

We define a category $\mathbf{Asm}(M,M_0)$.\bob{``Asm'' is the wrong
  name here.}

\begin{definition}[Resourced Sets, Semantic Contexts and Types]
  \label{defn:rl-object}
  \begin{enumerate}
  \item A \emph{resourced set} $X$ is a pair $(|X|,\models_\Gamma)$ of
    a set $|X|$ and a ternary relation
    $\mathord{\models_\Gamma} \subseteq |M| \times \ValSet \times |X|$. We write
    the first two elements to the left of the $\models$ and the last
    to the right: e.g. $\alpha, v \models_X x$.
  \item \emph{Semantic contexts} $\Gamma$ are resourced sets.
  \item Given a set $\Delta$ (representing a context bereft of any
    resource information), a family $S$ of semantics types over
    $\Delta$ is a collection $\{S_\delta\}_{\delta \in \Delta}$ of
    resourced sets. We write $\Ty(\Delta)$ for the large set of all
    semantic types over $\Delta$.
  \end{enumerate}
\end{definition}
Note how the intepretation of types only depends on a set of values,
not any resource information. This corresponds to

We define scaling and addition of resourced sets (and also semantic
contexts) to interpret the scaling and addition of syntactic contexts
\autoref{defn:FIXME}.

\begin{definition}[Scaling and Addition of Resourced Sets]
  \begin{enumerate}
  \item \emph{(Scaling)} If $X$ is a resourced set and
    $n \in \mathbb{N}$, then $nX$ is the resourced set with
    $|nX| = |X|$ and $\alpha, v \models_{nX} x$ iff there exist
    $\alpha_1, \dots, \alpha_n$ such that
    $\alpha \leq \alpha_1 \oplus \dots \oplus \alpha_n$ and for all
    $1 \leq i \leq n$, $\alpha_i, v \models_X x$.
  \item \emph{(Addition)} If $X$ and $Y$ are resourced sets such that
    $|X| = |Y|$, then define their addition $X + Y$ with
    $|X + Y| = |X|$ and $\alpha, v \models_{X + Y} x$ iff there exist
    $\alpha_1, \alpha_2$ such that
    $\alpha \leq \alpha_1 \oplus \alpha_2$ and
    $\alpha_1, v \models_X x$ and $\alpha_2, v \models_Y x$.
  \end{enumerate}
\end{definition}

Morphisms are where the connection between abstract elements, machine
values, and potential is given force:
\begin{definition}[Morphisms, Semantic Substitutions and Terms]
  \label{defn:rl-morphism}
  \begin{enumerate}
  \item Given resourced sets $X$ and $Y$, a morphism $f : X \to Y$ is
    a function $f : |X| \to |Y|$ for which there exists a closed
    expression $e \in \ExpSet$ (the \emph{realiser}) and a potential
    $\gamma \in |M_0|$ such that for all $\alpha, v, x$ with
    $\alpha, v \models_X x$, there exists a $\beta$, $v'$ and $k$ such
    that:
    \begin{enumerate}
    \item $e, [v] \Downarrow_k v'$ (\emph{the realiser evaluates to a result value}) and
    \item $k \leq M(\alpha + \gamma, \beta)$ (\emph{the potential difference yields adequate fuel}) and
    \item $\beta, v' \models_Y f(x)$ (\emph{the result is correct}).
    \end{enumerate}
  \item A \emph{semantic substitution} $\Gamma \Rightarrow \Delta$ is
    a morphism between the resourced sets.
  \item Given a semantic context $\Gamma$ and a semantic type
    $S \in \Ty(|\Gamma|)$, a \emph{semantic resourced term} $m$ is a
    function $m : (\gamma \in |\Gamma|) \to |S(\gamma)|$ such that there
    exists a closed expression $e \in \ExpSet$ and potential
    $\gamma \in |M_0|$ satisfying a similar property to that for
    morphisms. We write $\RTm(\Gamma, S)$ for the set of all semantic
    resourced terms between $\Gamma$ and $S$.
  \item Given a set $\Delta$ and a semantic type $S \in \Ty(\Delta)$,
    a \emph{semantic unresourced term} $m$ is a function
    $m : (\delta \in \Delta) \to |S(\delta)|$. We write
    $\Tm(\Gamma, S)$ for the set of all semantic unresourced terms
    between $\Delta$ and $S$.
  \end{enumerate}
\end{definition}
The definition of \emph{semantic resourced term} is the key to our
claim of soundness for any applications of QTT. By choosing the
correct resource monoid, and interpreting QTT with extensions
supported by that monoid into the resulting structure, we can deduce
our desired resource soundness theorems. FIXME: forward references.

Notice that the resource potential $\gamma$ for morphisms and
resourced terms must come from the sub-monoid $M_0$. For the
polynomial time systems, this will ensure that programs cannot
generate iterability from scratch. This is benign in the case of the
cons-free system (though it does complicate the statement of
soundness), but would be catastrophic in the LFPL-style system because
it would mean that we could generate $\Diamond$s from nothing.

To be able to interpret

\begin{definition}
  Context extension
\end{definition}

\subsection{Realising Finite Type Formers}

\begin{definition}[$\Pi$-types]
  FIXME: taken from
\end{definition}

\begin{definition}[$\Sigma$-types]

\end{definition}

\begin{definition}[Equality types]

\end{definition}

\begin{definition}[Non-iterable types]

\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Realising Iteration for Implicit Polynomial Time}
\label{sec:realising-iteration}

Explain why the polytime iterators work.

\begin{theorem}[Soundness for the Cons-free System]
  \label{thm:cons-free-soundness}
\end{theorem}

\begin{theorem}[Soundness for the LFPL-style System]
  \label{thm:lfpl-soundness}
\end{theorem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Realising Explicit Amortised Complexity}
\label{sec:explicit-amortised-complexity}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Relationship to Second-order MALL with Modalities}
\label{sec:poly-mall}

\begin{enumerate}
\item Compare the SAL and LFPL modalities to the natural number
  constructions
  \begin{enumerate}
  \item In the SAL case, we have $\oc A \multimap A^n$, which allows
    us to
  \end{enumerate}
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programming in Polytime}
\label{sec:programming-polytime}

\begin{enumerate}
\item Non-deterministic polytime via a bit oracle -- needs a statement
  of success
\item Previous could also be interpreted as probabilistic polytime
  (access to coin flips); can we formalise some kind of cryptographic
  property this way? Have a look at hash functions in the
  Authenticated Data Structures paper.
\item Insertion sort, quicksort as polytime functions.
\item A type of polytime functions; via a reflection operator
\item Programming with explicit resource annotations, including
  reasoning about the resource requirements directly.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conclusion}

\begin{enumerate}
\item Brunel's classical quantitative realisability
\item Works on polytime realisability for proof extraction (Schwictenberg, Soft Set Theory)
\item Lower complexity classes, like LOGSPACE
\item Higher complexity classes, like EXPTIME, ELEMENTARY
\item More fine-granied classes, like PTIME($k$)
\item Automation, in the style of Jan's stuff
\item Circuits and shit.
\end{enumerate}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
  FIXME
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

\end{document}
\endinput

### 2023-06-20 talks/graded-types-kent/slides.tex
Add TeX sources, and make initial version of Nix build

% -*- TeX-engine: xetex -*-

\documentclass[xetex,serif,mathserif,aspectratio=169]{beamer}

\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{attrib}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{mathpartir}
\usepackage{tikz}
\usepackage{cmll}

\usepackage{libertine}
\usepackage{newtxmath}
\usepackage{zi4}

\newcommand{\TyBool}{\mathsf{Bool}}
\newcommand{\TySet}{\mathsf{Set}}
\newcommand{\TyEl}{\mathsf{El}}

\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\RTm}{\mathrm{RTm}}
\newcommand{\wk}{\mathsf{wk}}
\newcommand{\proj}{\mathsf{p}}
\newcommand{\vartm}{\mathsf{v}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\cat}[1]{\mathcal{#1}}
\newcommand{\id}{\mathrm{id}}

\newcommand{\op}{\mathsf{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\skw}[1]{\mathit{#1}}

\usepackage{mathtools}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

\newcommand{\tyPrim}[2]{\textup{\texttt{#1}}\langle #2 \rangle}

\newcommand{\sechead}[1]{\textcolor{titlered}{\emph{#1}}}

\newcommand{\typeOfCartSp}[1]{\lbag #1 \rbag}

\def\greyuntil<#1>#2{{\temporal<#1>{\color{black!40}}{\color{black}}{\color{black}} #2}}
\def\greyfrom<#1>#2{{\temporal<#1>{\color{black}}{\color{black!40}}{\color{black!40}} #2}}

\newcommand{\superscript}[1]{\ensuremath{^{\textrm{#1}}}}
\newcommand{\highlight}{\textbf}

\newcommand{\atomprop}{\mathrm}
\newcommand{\true}{\mathbf{T}}
\newcommand{\false}{\mathbf{F}}

\newcommand{\append}{\mathop{+\kern-3pt+}}

\newcommand{\citationgrey}[1]{\textcolor{black!60}{#1}}

\definecolor{titlered}{rgb}{0.8,0.0,0.0}

\newcommand{\hlchange}[1]{\setlength{\fboxsep}{1pt}\colorbox{black!20}{$#1$}}
\newcommand{\altdiff}[3]{\alt<-#1>{#2}{\hlchange{#3}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% from http://tex.stackexchange.com/questions/118410/highlight-terms-in-equation-mode
\newlength{\overwritelength}
\newlength{\minimumoverwritelength}
\setlength{\minimumoverwritelength}{0.1cm}
\def\overwrite<#1>#2#3{%
  \settowidth{\overwritelength}{$#2$}%
  \ifdim\overwritelength<\minimumoverwritelength%
    \setlength{\overwritelength}{\minimumoverwritelength}\fi%
  \temporal<#1>{#2}%
    {\stackrel%
      {\begin{minipage}{\overwritelength}%
          \color{red}\centering\small #3\\%
          \rule{1pt}{9pt}%
        \end{minipage}}%
      {\colorbox{red!50}{\color{black}$\displaystyle#2$}}}%
    {\stackrel%
      {\begin{minipage}{\overwritelength}%
          \color{red}\centering\small #3\\%
          \rule{1pt}{9pt}%
        \end{minipage}}%
      {\colorbox{red!50}{\color{black}$\displaystyle#2$}}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\setbeamertemplate{navigation symbols}{}
\usecolortheme[rgb={0.8,0,0}]{structure}
\usefonttheme{serif}
\usefonttheme{structurebold}
\setbeamercolor{description item}{fg=black}

\title{Unifying Models of Linear/Graded Dependent Types}
\author{Robert Atkey \\
  \emph{University of Strathclyde} \\
  robert.atkey@strath.ac.uk}
\date{Graded Types Meeting, U. of Kent, 17th June 2022}

\newcommand{\youtem}{\quad \textcolor{titlered!80}{---} \quad}

\newcommand{\titlecard}[1]{\begin{frame}%
  \begin{center}%
    \Large \textcolor{titlered}{#1}%
  \end{center}%
\end{frame}}

\newcommand{\HEAD}[1]{\textcolor{titlered}{#1}}

\begin{document}

\frame{\titlepage}

\begin{frame}
  Lots of proposals for combining type dependency with
  gradings/linearity/intensional information.

  \bigskip

  \begin{itemize}
  \item Linear Dependent Types \citationgrey{(V{\'a}k{\'a}r, 2014)}
  \item Integrating Linear and Dependent Types \citationgrey{(Krishnaswami, Pradic, Benton, 2015)}
  \item Quantitative Type Theory \citationgrey{(McBride, 2016; Atkey, 2018)}
  \item Linear DTT for Quantum Programming Languages \citationgrey{(Fu, Kishida, Selinger, 2020)}
  \item Graded Modal Dependent Type Theory \citationgrey{(Moon, Eades, Orchard, 2021)}
  \item Graded Dependent Type System \citationgrey{(Choudhury, Eades, Eisenberg, Weirich, 2021)}
  \item Parametric Quantifiers \citationgrey{(Nuyts, Vezzosi, Devriese, 2017)}
  \item Multi-modal type theory \citationgrey{(Gratzer, Kaavos, Nuyts, Birkedal, 2021)}
  \item Dependent Dependency Calculus \citationgrey{(Choudhury, Eades, Weirich, 2022)}
  \item ...
  \end{itemize}
\end{frame}

\titlecard{But what is \emph{really} going on?}

\begin{frame}

  \begin{enumerate}
  \item<1-> Linearity vs. dependency. \\
    {\small Does mentioning a variable in a type count as a use? Can
      specifications consume resources?}

    \bigskip
  \item<2-> Intensional information used for specification \\
    {\small productivity; information flow; parametricity; ...}

    \bigskip
  \item<3-> Intensional information used for runtime properties \\
    {\small resource usage; “sub-Turing” implementability}
  \end{enumerate}

\end{frame}

\begin{frame}
  \frametitle{Linearity vs Dependency}

  QTT pretends that you can interleave linearity and dependency
  \begin{displaymath}
    x_1 \stackrel{\rho_1}: A_1, ..., x_n \stackrel{\rho_n}: A_n \vdash M \stackrel\sigma: B
  \end{displaymath}
  \begin{itemize}
  \item But really there are two systems; $\sigma = 0$ and $\sigma = 1$
  \item Every type has two parts:
    \begin{itemize}
    \item a specification part (normal type theory)
    \item an implementation part (linear type theory)
    \end{itemize}
  \item In the $\sigma = 0$ fragment, only the specification part
    matters; in particular types can only depend on the specification
    part.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Linearity vs Dependency}

  Split context systems \citationgrey{(Krishnaswami et al., 2015; V{\'a}k{\'a}r, 2014)}
  \begin{displaymath}
    x_1 : A_1, ..., x_n : A_n \mid y_1 : X_1, ..., y_m : X_m \vdash M : Y
  \end{displaymath}

  \bigskip
  Explicit separation of specification types $A_i$ and implementation types $X_j$.

  \bigskip
  Conjectures:
  \begin{enumerate}
  \item QTT can be translated into a split context system
  \item Split context system can be translated into QTT \\
    \quad {\small (probably; terms and conditions apply)}
  \item GMDTT and GraD can be translated into split context system
    (assuming a universe)
  \end{enumerate}

  \pause
  \begin{center}
    {\Large All Linear Dependent Type Theories are a mode of use
      of split contexts}
  \end{center}
\end{frame}

\begin{frame}
  \frametitle{General Semantic Picture}

  Looking at models helps us (me) understand how these interact.

  \bigskip

  \citationgrey{(V{\'a}k{\'a}r, 2015)}
  \begin{displaymath}
    L : \mathcal{C}^{op} \to \textbf{SMCCat}
  \end{displaymath}

  \emph{i.e.}, $\mathcal{C}$-indexed symmetric monoidal categories,
  plus $\mathcal{C}$ is a \emph{Category with Families} and:
  \begin{displaymath}
    L_{\Gamma.A}(\mathsf{p}^*X,Y) \cong L_\Gamma(X, \forall_AY)
  \end{displaymath}

  \medskip
  \begin{itemize}
  \item QTT: Grothendieck construction applied to $L$ \\
    {\small $\cdot$ \quad but each $L_\Gamma$ must be thin! (can this be relaxed?)} \\
    {\small $\cdot$ \quad need a theory of graded multicategories}
  \item QTT is essentially a form of abstract realisability
  \item The $\mathsf{G}$ constructor of Krishnaswami et al.~yields a
    notion of ``reflection'' for QTT
  \item Could also be expressed in terms of \emph{monoidal fibrations}
    or \emph{internal categories}.
  \end{itemize}

\end{frame}

% \begin{frame}
%   \frametitle{Reflection}

%   Starting on making the implementation and specification parts interact:

%   \begin{displaymath}
%     \mathrm{Tm}(\Gamma, GX) \cong L_\Gamma(I, X)
%   \end{displaymath}

%   \begin{mathpar}
%     \inferrule*
%     {0\Gamma \vdash M \stackrel{1}: X}
%     {0\Gamma \vdash \mathsf{G}M \stackrel{0}: \mathsf{G}X}

%     \inferrule*
%     {\Gamma \vdash M \stackrel{\sigma}: \mathsf{G}X}
%     {\Gamma \vdash \mathsf{G}^{-1}M \stackrel{1}: X}
%   \end{mathpar}

%   \begin{itemize}
%   \item
%   \end{itemize}
% \end{frame}

\begin{frame}

  \begin{enumerate}
  \item Do these translations or this general model actually work?

    \bigskip
  \item Can we reason about the intensional information in
    implementations?

    \medskip {\small e.g., if we know a term is polytime computable,
      then what can we do with this?}

    \bigskip
  \item How can we systematically create systems that combine
    intensional information at the type level with runtime
    information?

    \medskip {\small e.g., productivity information, }

    \medskip {\small Conjecture: multi-modal type theory will work
      well for this (in place of $\mathcal{C}$)}
  \end{enumerate}
\end{frame}

\end{document}

### 2023-07-05 paper/paper.tex
[ paper ] Some work on describing the realisability model

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Implicit Polynomial Time and Dependent Types}
+\title{Polynomial Time and Full Dependent Types}
 
@@
 
-The story:
+{\bf New story:}
+\begin{enumerate}
+\item Introduction: The goal of this paper is to unify implicit
+  computational complexity with full dependent types. ICC enables us
+  to ensure that \emph{programs} are within a certain complexity
+  class, while dependent types allow us to state and prove that these
+  programs solve the problems that we intend them to, and to construct
+  reductions between problems. We can marry complexity-controlled
+  implementations with problems, paving the way for a \emph{synthetic
+    computational complexity theory}.
+\item Section 1: Linear types and Implicit Polynomial Time.
+  \begin{enumerate}
+  \item Total typed functional programming, as we see in type theory,
+    is restricted to the use of higher order functions and iteration
+    over inductive data types.
+  \item Polynomial time is the smallest class closed under nested
+    iteration over the input -- i.e. we consider iteration over the
+    input as feasible, and compositions thereof. The key thing is to
+    avoid iteration over data constructed from the input.
+  \item Key is to control iteration. So we split inductive datatypes
+    into iterable and non-iterable versions. We can
+  \item One way to capture polytime is to ensure that all iteration is
+    only done over the input. So we can duplicate iterable things, but
+    we can't construct new iterable things (at least not from an
+    iteration). We also can't duplicate higher order things, because
+    that effectively acts as a way to duplicate iterations (half
+    instantiated iterators). Compare to Jones' CONS-FREE programming.
+  \item Another approach is to allow ``borrowing'' of iterability by
+    means of diamonds. Iterability becomes a first-class object, so we
+    can use it to construct new data structures from the input. This
+    allows the construction of new iterable data structures. LFPL.
+  \end{enumerate}
+\item Section 2: Extending to Dependent Types
+  \begin{enumerate}
+  \item Combining linear types and dependent types is subtle, because
+    there is a fundamental incompatibility in the syntax of the two
+    systems.
+  \item There have been several approaches, though they are all
+    somewhat similar (list them all here). We will use QTT, because it
+    offers a firm semantic foundation as well as a relatively
+    straightforward syntax.
+  \item Recap the syntax of QTT, with extensional equality. Key point
+    is that there is a normal dependently typed language with a linear
+    sub-language. We can have special constructs in the linear
+    sub-language that ``erase'' to normal type theory -- products are
+    the prime example.
+  \item Non iterable datatypes -- have normal eliminators in the DTT
+    fragment, but
+  \end{enumerate}
+\item Section 3: Programming and Proving with Poly-time
+  \begin{enumerate}
+  \item Internalised completeness proofs.
+  \item Construct the type of problems that are solvable in polynomial
+    time.
+  \item Construct reductions between problems.
+  \item NP. Doesn't make sense to talk about this without dependent
+    types, because we can't say when we've solved the problem being
+    stated.
+  \item PP (Probabilistic), using a monad. Again, need to be able to
+    state connection with the problem to know that we are solving it.
+  \end{enumerate}
+\item Section 4: Soundness via Realisability
+  \begin{enumerate}
+  \item The underlying CBV language and its
+  \end{enumerate}
+\item Section 5: Extensions
+  \begin{enumerate}
+  \item Explicit Complexity via Diamond Counting
+  \item Poly-time univalence
+  \item Towards P- and NP-completeness proofs. Issue here is the
+    extensionality. Towards a synthetic (poly-time based)
+    computational complexity theory.
+  \end{enumerate}
+\item Section 6: Related Work
+  \begin{enumerate}
+  \item ICC stuff
+    \begin{enumerate}
+    \item LFPL
+    \item CONS-free
+    \item Soft Affine Logic; which has the same underlying idea as
+      CONS-free (is this a novel observation?)
+    \item Other systems based on information flow.
+    \end{enumerate}
+  \item Explicit complexity: lambda-amor, and things by Deepak Garg
+    and Marco Gaboardi.
+  \item Recent CMU stuff on calf
+  \end{enumerate}
+
+\end{enumerate}
+
+{\bf Old story:}
 \begin{enumerate}
@@
 
+\newcommand{\rplus}{\oplus}
+\newcommand{\rzero}{\emptyset}
+
+FIXME: introductory paragraph.
+
 The Hofmann-Dal Lago approach to proving the soundness of implicit
@@
     denote elements of a resource monoid.
-  \item Commutative monoid structure $(\oplus,\emptyset)$ on $|M|$, so
+  \item Commutative monoid structure $(\rplus,\rzero)$ on $|M|$, so
     we can add potentials.
@@
     \item for all $\alpha, \beta, \gamma$,
-      $M(\alpha, \beta) + M(\beta, \gamma) \leq M(\alpha, \gamma)$.
+      $M(\alpha, \beta) \rplus M(\beta, \gamma) \leq M(\alpha, \gamma)$.
     \end{enumerate}
@@
     moving from $\alpha$ to $\gamma$ directly.
+  \item \bob{Interaction between $(\rplus,\rzero)$ and $M(\alpha,\beta)$.}
   \item An \emph{accounting function}
     $\mathit{acct} : \mathbb{N} \to |M|$ such that for all $k$,
-    $k \leq M(\mathit{acct}(k),\emptyset)$.
+    $k \leq M(\mathit{acct}(k),\rzero)$.
   \end{enumerate}
@@
 
+\bob{PLAN: RCwFs, and state that they can be constructed from Indexed Linear Posets}
+
 \cite{atkey18} describes a general class of QTT models termed
@@
 
-\subsection{Realisability Model}
+\bob{define the construction}
+\begin{enumerate}
+\item Start with a CwF $\cat{C}$ that supports $\Pi$, $\Sigma$ and
+  natural numbers.
+\item Assume a functor $L : \cat{C}^\top \to \LinPoset$ with
+  reindexing along projections having a right adjoint.
+\item Construct a new CwF with the same base category as $\cat{C}$,
+  but with $\Ty'(\Delta) = \Sigma A : \Ty(\Delta).~L(\Delta.A)$ and
+  $\Tm'(\Delta, (T, X)) = \Tm(\Delta, T)$.
+\item
+\end{enumerate}
+
+This construction is implicit in \cite{atkey2018qtt}, where the
+indexed linear poset is defined from a $\mathcal{R}$-LCA.
+
+\subsection{Amortised Complexity Realisability Model}
 \label{sec:realisability-model}
-
-\bob{Rearrange this as an indexed linear poset}
 
@@
 (\autoref{sec:machine-model}) and a compositional notion of resource
-potential (\autoref{sec:resource-monoids}), we now begin to construct
-models of QTT that witness the resource (and type) soundness of our
-complexity constrained systems.
-
-\bigskip
-
-\cite{atkey18qtt} describes a general class of QTT models termed
-\emph{Quantitative Categories with Families} (QCwFs), an extension of
-the \cite{dyber}'s \emph{Categories with Families} formulation of
-models of dependent type theory. We will not regurgitate the full
-definition here. Instead, we will define each of the parts of our
-specialised QCwF for proving complexity soundness and explain
-informally how to use them to model the syntax of QTT. Please refer to
-\cite{atkey18qtt} for details of the interpretation.
-
-Fix a resource monoid $M$ with a submonoid $M_0 \subseteq M$.
-
-We define a category $\mathbf{Asm}(M,M_0)$.\bob{``Asm'' is the wrong
-  name here.}
-
-\begin{definition}[Resourced Sets, Semantic Contexts and Types]
-  \label{defn:rl-object}
-  \begin{enumerate}
-  \item A \emph{resourced set} $X$ is a pair $(|X|,\models_\Gamma)$ of
-    a set $|X|$ and a ternary relation
-    $\mathord{\models_\Gamma} \subseteq |M| \times \ValSet \times |X|$. We write
-    the first two elements to the left of the $\models$ and the last
-    to the right: e.g. $\alpha, v \models_X x$.
-  \item \emph{Semantic contexts} $\Gamma$ are resourced sets.
-  \item Given a set $\Delta$ (representing a context bereft of any
-    resource information), a family $S$ of semantics types over
-    $\Delta$ is a collection $\{S_\delta\}_{\delta \in \Delta}$ of
-    resourced sets. We write $\Ty(\Delta)$ for the large set of all
-    semantic types over $\Delta$.
-  \end{enumerate}
-\end{definition}
-Note how the intepretation of types only depends on a set of values,
-not any resource information. This corresponds to
-
-We define scaling and addition of resourced sets (and also semantic
-contexts) to interpret the scaling and addition of syntactic contexts
-\autoref{defn:FIXME}.
-
-\begin{definition}[Scaling and Addition of Resourced Sets]
-  \begin{enumerate}
-  \item \emph{(Scaling)} If $X$ is a resourced set and
-    $n \in \mathbb{N}$, then $nX$ is the resourced set with
-    $|nX| = |X|$ and $\alpha, v \models_{nX} x$ iff there exist
-    $\alpha_1, \dots, \alpha_n$ such that
-    $\alpha \leq \alpha_1 \oplus \dots \oplus \alpha_n$ and for all
-    $1 \leq i \leq n$, $\alpha_i, v \models_X x$.
-  \item \emph{(Addition)} If $X$ and $Y$ are resourced sets such that
-    $|X| = |Y|$, then define their addition $X + Y$ with
-    $|X + Y| = |X|$ and $\alpha, v \models_{X + Y} x$ iff there exist
-    $\alpha_1, \alpha_2$ such that
-    $\alpha \leq \alpha_1 \oplus \alpha_2$ and
-    $\alpha_1, v \models_X x$ and $\alpha_2, v \models_Y x$.
-  \end{enumerate}
-\end{definition}
-
-Morphisms are where the connection between abstract elements, machine
-values, and potential is given force:
-\begin{definition}[Morphisms, Semantic Substitutions and Terms]
-  \label{defn:rl-morphism}
-  \begin{enumerate}
-  \item Given resourced sets $X$ and $Y$, a morphism $f : X \to Y$ is
-    a function $f : |X| \to |Y|$ for which there exists a closed
-    expression $e \in \ExpSet$ (the \emph{realiser}) and a potential
-    $\gamma \in |M_0|$ such that for all $\alpha, v, x$ with
-    $\alpha, v \models_X x$, there exists a $\beta$, $v'$ and $k$ such
-    that:
-    \begin{enumerate}
-    \item $e, [v] \Downarrow_k v'$ (\emph{the realiser evaluates to a result value}) and
-    \item $k \leq M(\alpha + \gamma, \beta)$ (\emph{the potential difference yields adequate fuel}) and
-    \item $\beta, v' \models_Y f(x)$ (\emph{the result is correct}).
-    \end{enumerate}
-  \item A \emph{semantic substitution} $\Gamma \Rightarrow \Delta$ is
-    a morphism between the resourced sets.
-  \item Given a semantic context $\Gamma$ and a semantic type
-    $S \in \Ty(|\Gamma|)$, a \emph{semantic resourced term} $m$ is a
-    function $m : (\gamma \in |\Gamma|) \to |S(\gamma)|$ such that there
-    exists a closed expression $e \in \ExpSet$ and potential
-    $\gamma \in |M_0|$ satisfying a similar property to that for
-    morphisms. We write $\RTm(\Gamma, S)$ for the set of all semantic
-    resourced terms between $\Gamma$ and $S$.
-  \item Given a set $\Delta$ and a semantic type $S \in \Ty(\Delta)$,
-    a \emph{semantic unresourced term} $m$ is a function
-    $m : (\delta \in \Delta) \to |S(\delta)|$. We write
-    $\Tm(\Gamma, S)$ for the set of all semantic unresourced terms
-    between $\Delta$ and $S$.
-  \end{enumerate}
-\end{definition}
-The definition of \emph{semantic resourced term} is the key to our
-claim of soundness for any applications of QTT. By choosing the
-correct resource monoid, and interpreting QTT with extensions
-supported by that monoid into the resulting structure, we can deduce
-our desired resource soundness theorems. FIXME: forward references.
-
-Notice that the resource potential $\gamma$ for morphisms and
-resourced terms must come from the sub-monoid $M_0$. For the
-polynomial time systems, this will ensure that programs cannot
-generate iterability from scratch. This is benign in the case of the
-cons-free system (though it does complicate the statement of
-soundness), but would be catastrophic in the LFPL-style system because
-it would mean that we could generate $\Diamond$s from nothing.
-
-To be able to interpret
-
-\begin{definition}
-  Context extension
-\end{definition}
-
-\subsection{Realising Finite Type Formers}
-
-\begin{definition}[$\Pi$-types]
-  FIXME: taken from
-\end{definition}
-
-\begin{definition}[$\Sigma$-types]
-
-\end{definition}
-
-\begin{definition}[Equality types]
-
-\end{definition}
-
-\begin{definition}[Non-iterable types]
-
-\end{definition}
+potential (\autoref{sec:resource-monoids}), we construct models of QTT
+that witness the resource and type soundness of our complexity
+constrained systems.
+
+\bob{Tie back to previous section}
+
+We fix a resource monoid $M$ with distinguished sub-monoid $M_0$.
+
+\subsubsection{Base CwF}
+\label{sec:realisability-base-cwf}
+
+As our base CwF, we use the category $\Set$ of sets and functions with
+the standard interpretation of $\Pi$, $\Sigma$ types and the
+$\mathbb{N}$ type.
+
+\subsubsection{Indexed Linear Preorder}
+\label{sec:realisability-indexed-linear-preorder}
+
+We now define an indexed linear poset $L$ of realisers over $\Set$
+that ties together our ``mathematical'' model of types in $\Set$ with
+our machine model and resource monoid. For a set $A$, the carrier of
+$L(A)$ is the set of ternary relations
+$X \subseteq A \times M \times \ValSet$ and we define the ordering
+$X \leq Y$ to hold iff there exists realising expression
+$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
+$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
+$(a,\alpha,v) \in X$, we have that there exists a result
+$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
+$\beta \in M$ with:
+\begin{enumerate}
+\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
+\item $(a, \beta, v') \in Y$ (the result is well-resourced and
+  satisfies $Y$); and
+\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
+  difference between the initial potential and the result potential).
+\end{enumerate}
+Note that the definition of realisablity is uniform in the element $a$
+-- the realising expression $e$ and the potential $\gamma$ must work
+for all $a$ -- thus the implementation and complexity measure of the
+transition being modelled cannot depend on what the input is. Put in
+implementation terms, the input $a$ is not present at
+runtime. Moreover note that the potential $\gamma$ attached to the
+expression $e$ must come from the sub-monoid $M_0$, indicating that is
+intended to be data-free, while the potential $\alpha$ for the input
+is from the full monoid $M$, so it can contain data and functions.
+
+For $X, Y \in L(A)$, the required elements for symmetric monoidal
+closed structure are defined as follows. For the tensor product
+$X \otimes Y \in L(A)$, the realising value must be a pair $(v_1,v_2)$
+and the potential of the pair must split into suitable potentials
+$\alpha_1$, $\alpha_2$ for the components. For the residual
+$X \multimap Y$, the realising value must be a closure with potential
+to, when added to the potential of an input, compute the output with
+enough remaining. Note that the potential attached to a closure
+($\alpha$, here) need not be from the sub-monoid $M_0$. Unlike
+top-level term interpretations, closures may contain data.
+\begin{displaymath}
+  \begin{array}{lcl}
+    X \otimes Y &=& \{ (a, \alpha, (v_1, v_2)) \mid \exists \alpha_1, \alpha_2.~0 \leq M(\alpha, \alpha_1 \rplus \alpha_2) \land X(a,\alpha_1,v_1) \land Y(a,\alpha_2,v_2) \}\\
+    X \multimap Y &=& \{ (a, \alpha, \clo{E}{\eta}) \mid
+                      \begin{array}[t]{@{}l}
+                        \forall \alpha' \in M, v,w \in \ValSet.\,X(a,\alpha',v) \Rightarrow\\
+                        \quad \exists v', k, \beta.\,
+                        E, (\eta :: w :: v) \Downarrow_k v' \land Y(a,\beta,v') \land k \leq M(\alpha \rplus \alpha', \beta) \} \end{array}
+  \end{array}
+\end{displaymath}
+The seemingly useless $w \in \ValSet$ in the formula for
+$X \multimap Y$ is a dummy argument standing for the self-referential
+reference to the closure used for defining recursive programs.
+
+Each $L(A)$ has a terminal (i.e. top) element, which is also the unit
+for $\otimes$, defined as
+$I_A = \{(a, \alpha, *) \mid a \in A, \alpha \in M\}$. The potential
+$\alpha$ here is unrestricted, so $I_A$ can consume an arbitrary
+resource.
+
+$\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
+action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above \bob{do
+  this}. At the level of realising values, the modality $\oc_n$ has no
+effect; it only serves to alter the required potentials.
+\begin{displaymath}
+  \oc_n\,X = \{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
+\end{displaymath}
+
+$L$ also has arbitrary $\Set$-indexed products, realised ``lazily'' as
+functions that take dummy arguments. For $A \in \Set$ and
+$B \in A \to \Set$ and $X \in L(\Sigma A.\,B)$, we define
+$\forall_B X \in L(A)$ similarly to $\multimap$ above, but with
+different resource and indexing requirements:
+\begin{displaymath}
+  \forall_B X = \{ (a,\alpha,\clo{E}{\eta}) \mid \forall b, v.~\exists v', \beta, k. E, (\eta :: v :: *) \Downarrow_k v' \land X((a,b),\beta,v') \land k \leq M(\alpha,\beta) \}
+\end{displaymath}
+Note, as with the definition of $X \leq Y$ above, the realiser closure
+$\clo{E}{\eta}$ must be chosen uniformly for all $b$. This definition
+also appears to allow arbitrary computation (paid for by $\alpha$) to
+happen when the realising closure is applied, but the potential
+$\alpha$ will only ever be greater than $\beta$ by enough to handle
+the administrative costs of applying the function.
+
+To complete the construction of $L$ as an indexed linear preorder, we
+need to give realisers for each of the required inequalities in
+\autoref{def:indexed-linear-preorder}. In each case, this is a matter
+of programming in the language of \autoref{sec:machine-model}. The
+corresponding potentials are calculated by counting the steps in the
+ensuing programs. Note that, so far, all realisers are constant time
+(relative to their input), so the constructions so far work for any
+resource monoid. For systems that require iteration, we will need more
+structure on our resource monoids \bob{fwd ref}.
+
+\begin{proposition}
+  $L$, with $I$, $\otimes$, $\multimap$, $\oc_n$, and $\forall_{B}$ defined above, is
+  an indexed linear preorder.
+\end{proposition}
+
+\begin{proof}
+  See the Agda development.
+\end{proof}
+
+\begin{proposition}\label{prop:basic-qtt-realisability-model}
+  Get a model QTT with $\Pi$, $\Sigma$, and a universe.
+\end{proposition}
+
+\subsubsection{Non-iterable Data Types}
+
+The model of QTT constructed in
+\autoref{prop:basic-qtt-realisability-model} does not yet include any
+useful base types. Handling iterable types, which are the ones that
+induce non-constant-time complexities, requires specific properties of
+resource monoids that we introduce in
+\autoref{sec:realising-iteration}.
+
+Rather than perform a complex generic construction of datatypes to
+show how non-iterable versions may be realised, we focus on two
+representative examples. Booleans
+
+\paragraph{Booleans} Fix $\mathbb{B} = \{ \mathit{tt}, \mathit{ff} \}$
+as our set of boolean elements. We define an element of $L(\mathbb{B}$
+to represent boolean values:
+\begin{displaymath}
+  \mathrm{bool} = \{ (\mathit{tt}, \alpha, \synTrue) \mid \alpha \in M \} \cup \{ (\mathit{ff}, \alpha, \synFalse) \mid \alpha \in M \}
+\end{displaymath}
+Thus, the boolean $\mathit{tt}$ is represented by the value $\synTrue$
+and $\mathit{ff}$ is represented by $\synFalse$. In both case, we
+allow arbitrary potentials to be attached.
+
+\paragraph{Lists}
 

### 2023-07-09 paper/paper.tex
[ paper ] Lots of writing

@@
 \usepackage{mathpartir}
+
+\newcommand{\TapeType}{\mathrm{Tape}}
 
@@
 \newcommand{\If}{\mathrm{if}}
+\newcommand{\Then}{\mathrm{then}}
+\newcommand{\Else}{\mathrm{else}}
 \newcommand{\dupNat}{\mathrm{dupNat}}
-
+\newcommand{\Match}{\mathrm{match}}
+
+% QTT stuff
 \newcommand{\istype}{\mathrm{type}}
 \newcommand{\isctxt}{\mathrm{ctxt}}
+
+\newcommand{\BoolTy}{\mathrm{Bool}}
+\newcommand{\cTrue}{\mathsf{true}}
+\newcommand{\cFalse}{\mathsf{false}}
+\newcommand{\ListTy}{\mathrm{List}}
+\newcommand{\cNil}{\mathsf{nil}}
+\newcommand{\cCons}{\mathsf{cons}}
+
+\newcommand{\Rtype}{\mathbf{R}}
+\newcommand{\rIntro}{\mathbf{R}}
+\newcommand{\rElim}{\mathbf{R}^{-1}}
 
@@
 
-\newcommand{\LinPoset}{\mathrm{LinPoset}}
+\newcommand{\LinPreorder}{\mathrm{LinPreorder}}
 
 \newcommand{\bob}[1]{\textcolor{purple}{FIXME: #1}}
+
+\renewcommand{\sectionautorefname}{Section}
+\renewcommand{\subsectionautorefname}{Section} % PS
+\renewcommand{\subsubsectionautorefname}{Section}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Linear $\lambda$-Calculus and Implicit Computational
-  Complexity}
-\label{sec:linear-icc}
-
-Introduce the idea of linear typing for complexity control.
-
-Linear $\lambda$-calculus, as we shall be discussing it in this
-section:
-\begin{mathpar}
-  \inferrule* [right=Var]
+Related work: only applications of Linear TT so far are erasure (Abel
+et al.) and quantum stuff (Selinger et al.), and (sort of) dependency
+analysis. This is another one.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Linear $\lambda$-Calculus and Polytime}
+\label{sec:linear-intro}
+
+Not long after Girard introduced Linear Logic \cite{girard87}, it was
+observed that its resource sensitivity could be turned to describing
+computational complexity classes by purely logical means. Typically, a
+logical system is described for which the process of reducing a proof
+to a normal form (often by cut elimination) is guaranteed to always be
+accomplished within a certain complexity bound. Moreover, the system
+is usually proven to be complete for the relevant complexity class, by
+constructing a simulation of some known representation. Such systems
+that characterise polytime include Bounded Linear Logic (BLL), which
+uses explicit polynomials in the formulas \cite{bll} and Soft Affine
+Logic (SAL) \cite{lafont}, which does not explicitly represent
+time information in formulas, but uses a restricted form of Linear
+Logic's $\oc$ modality instead. Light Affine Logic (LAL) \cite{lal} is
+another ``counting-free'' system for polytime.
+
+Viewing logical systems though the lens of the Curry-Howard
+correspondence, the idea arises that one could define functional
+programming languages that characterise complexity classes such as
+polytime. SAL has been transformed into a programming language for
+polytime by \cite{gaboardi}. Hofmann proposed a new programming
+language, Linear Functional Programming Language (LFPL), that uses a
+novel ``payment'' system to track iteration. In this section, we
+review the use of linear types to capture polytime by presenting two
+systems, one based on ideas from SALL and the second more explicitly
+based on LFPL. With a view to adapting these systems to dependent
+types in \autoref{sec:qtt}, we take an approach slightly different to
+much of the polytime linear logic literature. We emphasise the use of
+explicit datatypes and eliminators, rather than using impredicative
+encodings via universal types. Thus, we are closer to Hofmann's
+original LFPL (though not later presentations of it
+\cite{dallago-hofmann}) than BLL, SALL, or LAL.
+
+For this section, the linear $\lambda$-calculus we will use will have
+linear functions and $\otimes$-products:
+\begin{mathpar}
+  \inferrule*
   { }
@@
 
-  \inferrule* [right=Lam]
+  \inferrule*
   {\Gamma, x : A \vdash M : B}
@@
 
-  \inferrule* [right=App]
+  \inferrule*
   {\Gamma_1 \vdash M : A \multimap B \\
@@
   {\Gamma_1, \Gamma_2 \vdash M\,N : B}
-\end{mathpar}
-
-\begin{mathpar}
-  \inferrule* [right=$\otimes$-Intro]
+
+  \inferrule*
   {\Gamma_1 \vdash M : A \\ \Gamma_2 \vdash N : B}
@@
 
-  \inferrule* [right=$\otimes$-Elim]
+  \inferrule*
   {\Gamma_1 \vdash M : A \otimes B \\ \Gamma_2, x : A, y : B \vdash N : C}
@@
 \end{mathpar}
-
-\subsection{Excluding Recursion}
-
-What do we get if we exclude recursion altogether?
-
-What happens if we include natural number recursion naively?
-
-\paragraph{A Binary Tape Datatype}
-
-FIXME: how to do this?
-
-move left, move right, update. Tape is always finite.
-
-\subsection{Cons-free programming}
-
-FIXME: zero case has a context, or do we $\lambda$-abstract?
+These rules are standard, so we do not describe them further except to
+note how linear typing uses presence or absences in a context to
+control resource usage -- if a variable is in the context it must be
+used exactly once. The fact that this discipline interferes with
+dependent types is one of the reasons we turn to QTT when we wish to
+add dependent types in \autoref{sec:qtt}.
+
+\subsection{No Recursion, Only Case Analysis}
+
+It is not too difficult to see that reduction of linear
+$\lambda$-terms always takes a number of steps linearly proprotional
+to the size of the term. This is because every $\beta$-redex
+substitutes into exactly one variable, reducing the size of the term
+by one each time.
+
+We can increase the expressivity, but not the computational
+complexity, of the system by adding datatypes that do not allow
+iteration. These can be used for representation, but not for driving
+computation. We include the rules here, to show how linearity must be
+preserved in these rules, and to foreshadow their dependently typed
+counterparts in \autoref{sec:qtt-noniter}. The first type is the
+booleans, which are anyway non-recursive and so wouldn't allow
+iteration anyway:
+\begin{mathpar}
+  \inferrule*
+  { }
+  {\vdash \cTrue, \cFalse : \BoolTy}
+
+  \inferrule*
+  {\Gamma_1 \vdash M : \BoolTy \\ \Gamma_2 \vdash N_1 : A \\ \Gamma_2 \vdash N_2 : A}
+  {\Gamma_1, \Gamma_2 \vdash \If\,M\,\Then\,N_1\,\Else\,N_2 : A}
+\end{mathpar}
+The if-then-else rule is careful to ensure that the resources used by
+the eliminated $\BoolTy$ and the resources used by the chosen branch
+are accounted for separately. The two branches must have the same
+resource usage.
+
+Construction and case analysis of lists are given by the following
+rules:
+\begin{mathpar}
+  \inferrule*
+  { }
+  {\vdash \cNil : \ListTy(A)}
+
+  \inferrule*
+  {\Gamma_1 \vdash M : A \\ \Gamma_2 \vdash N : \ListTy(A)}
+  {\Gamma_1, \Gamma_2 \vdash \cCons(M,N) : \ListTy(A)}
+
+  \inferrule*
+  {\Gamma_1 \vdash M : \ListTy(A) \\
+    \Gamma_2 \vdash N_1 : B \\
+    \Gamma_2, h : A, t : \ListTy(A) \vdash N_2 : B}
+  {\Gamma_1, \Gamma_2 \vdash \Match\,M\,\{\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\} : B}
+\end{mathpar}
+Hence, we can construct lists arbitrarily, but we can only do case
+analysis on them. If we wish to explore a list to a arbitrary depth it
+must be driven by an iterable type of the form that we will see in the
+next two sections.
+
+With booleans and lists, we can construct several other useful
+types. For example, to simulate Turing machines, one can construct a
+$\mathit{Tape}$ type as a Zipper \cite{huet}
+$\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
+representing a position on the tape with the items before, under, and
+after the head.
+
+\subsection{The Cons-free system}
+\label{sec:cons-free-intro}
+
+Polynomial time is usually seen as a proxy for ``feasible''
+computation. On the face of it, there does not seem to be any
+particular reason why polynomials have anything to do with
+feasibility. However, one can arrive at the definition of polynomial
+time in three steps, by assuming that (i) iterating over the whole
+input is feasible; (ii) if two computations are feasible, then so is
+their composition; and (iii) performing a feasible computation for
+every element of the input is also feasible. It is the last point that
+allows complexities of arbitrary polynomial degree to be constructed
+(we will see this in action in the completeness construction below and
+soundness proofs in \autoref{sec:iterator-soundness}).
+
+Following these ideas, let us assume that the input is a natural
+number, so we assume that there is some type of natural numbers
+$\tyNat$. For point (i), we must be able to iterate over these natural
+numbers, so we use a linear iterator defined by this typing rule:
 \begin{displaymath}
@@
 \end{displaymath}
-
+Note that in the zero, $M_z$, and sucessor, $M_s$, cases, the context
+is empty to ensure that these cases may be invoked as many times as
+required. Point (ii) above is automatically satisfied by being in a
+typed $\lambda$-calculus, where it is difficult to stop functions from
+being composable. For point (iii), the iterator as given does not
+allow us to nest iterations. Once the natural number input $n$ has
+been used for an iteration, the linear typing discipine prevents us
+from using it again (note the two separate contexts $\Gamma_1$,
+$\Gamma_2$ in the rule for application). In order to allow nested
+iterations, we add an operator to duplicate numbers:
 \begin{displaymath}
@@
 \end{displaymath}
-
-Intuitive reason it works: we can only iterate over the natural
-numbers we have recieved as input, there is no way to construct new
-natural numbers by iteration. Duplicating natural numbers is safe,
-becasue we can only do it a number of times controlled by iteration of
-the natural numbers we are given.
-
-FIXME: is it possible to pin down how it relates to SAL?
-
-\paragraph{Completeness}
-
+Somewhat surprisingly, this system is now sound and complete for
+polynomial time. Crucially, this depends on the two things we have
+\emph{not} allowed. First, we have disallowed the construction of new
+natural numbers via the $\conZero$ and $\conSucc$
+constructors\footnote{Actually, $\conZero$ would be acceptable, as
+  well as any constant natural number. It is only unrestricted use of
+  $\conSucc$ that is dangerous.}. If we were to permit this, then we
+could use iteration over the input to construct addition,
+multiplication (by repeated addition), and then exponentials (by
+repeated multiplication). We therefore refer to this system as the
+\emph{Cons-Free} system.  The second prohibited feature is the ability
+to duplicate values of function type, even though we have allowed
+duplication of iterable naturals. If we were to allow this, then we
+would be able to sneak in a form of constructors for natural numbers
+by encoding them as eliminators that duplicate a function for every
+$\conSucc$ step.
+
+We will see in \autoref{sec:consfree-sound} that this system is sound
+for polytime by a realisability argument. Completeness can be seen
+more directly by constructing a function that iterates a function for
+a statically known polynomial number of times in the size of the
+input. Assume that we have a known polynomial
+$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
+coefficients and some single step function
+$f : \mathit{St} \multimap \mathit{St}$ over a state type
+$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
+steps. Then, using the iterator above we can iterate $f$ over a
+$\tyNat$ representing the size of the input:
 \begin{displaymath}
@@
 \end{displaymath}
-
+To achieve higher degrees, we can use $\dupNat$ to nest iterations:
 \begin{displaymath}
@@
 \end{displaymath}
+By further use of $\dupNat$ and composition to handle addition of
+polynomials, the function $f$ can now be iterated $p(n)$ many times,
+where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
+represent all polytime computations.
 
@@
 
-The cons-free style of programming allows us to express all polytime
-Turing machines, but is quite awkward from the point of view of
-functional programming. It allows us to iterate over natural numbers
-that come from the input but does not allow us to build further values
-to do iteration on. For example, if our input is a list, then we
-cannot transform it into a binary search tree and then flatten it
-\bob{actually write this}.
-
-Introduce a new type $\Diamond$. Each $\Diamond$ value represents a
-``chunk of iterability''.
-
-{\bf Rough model:} programs are polynomials, data are bags of
-diamonds. Applying a program to some data yields a concrete number of
-steps that will bound the resulting computation.
-
+The Cons-free system is sound and complete for polytime, but is quite
+awkward from the point of view of functional programming. It allows us
+to iterate over natural numbers that come from the input but does not
+allow us to build further values to do iteration on. For example, if
+our input is a list, then we cannot transform it into a binary search
+tree and then flatten it, we must always refer back to the original
+natural number input. Even dividing the input into two halves to be
+treated separately is difficult.
+
+A more flexible system was proposed by \cite{hofmann99}. The
+\emph{Linear Functional Programming Language} (LFPL) takes a different
+approach to the \emph{Cons-free} system. Instead of completely
+prohibiting construction of new data, constructions is allowed if it
+is paid for by values of new type $\Diamond$ (``diamonds''):
 \begin{mathpar}
@@
 \end{mathpar}
-
+Thus, to construct a $\conZero$, we must have a $\Diamond$ to pay for
+it, and likewise, to construct a $\conSucc$ we must have a diamond. We
+can think of $\Diamond$s as an unit of iterable data. Iterability is
+``saved up'' in data during construction, and released during
+iteration. Diamonds cannot be created by a program itself, for the
+same reason that constructors were prohibited in the \emph{Cons-free}
+system, but they are released from iterable data during iteration. The
+LFPL natural number iterator has the following typing rule:
 \begin{displaymath}
@@
 \end{displaymath}
-
-\paragraph{Other datatypes} Trees, and so on.
-
-\paragraph{Completeness for Polytime} We now show how every polytime
-Turing machine can be simulated in LFPL with a non-iterable type of
-tapes. The construction, due to Aelhig and Schwictenberg
-\cite{syntactic-lfpl}, illustrates how programming with diamonds in
-LFPL differs from the soft system in the previous section.
-
-Assume a state type $\mathit{St}$ and a step function $f : \mathit{St} \multimap \mathit{St}$.
-
-The linear iterator:
+The difference with the \emph{Cons-free} iterator above is that the
+$\conZero$ and $\conSucc$ cases now both have an addition binding of
+type $\Diamond$. This allows some form nesting of iterations: during
+an iteration over the input, the program can accumulate $\Diamond$s to
+use for iteration over substructures that are smaller than the current
+point in the iteration. A construction, due to Aelhig and
+Schwictenberg \cite{syntactic-lfpl}, illustrates how this leads to all
+polytime computations. As above, we assume a polynomial $p(n)$ and a
+step function $f : \mathit{St} \multimap \mathit{St}$ that needs to be
+iterated $p(n)$ times. We construct a linear iterator:
 \begin{displaymath}
@@
 \end{displaymath}
-
-Step:\textbf{Is the extra $f\,s$ required here?}
+Note that this iterator returns the natural number input as well as
+the new state. LFPL does not allow duplication of iterable inputs, so
+we must always reconstruct it if we want to do further
+iteration. Addition of polynomials is accomplished by composition of
+iterators. To raise the degree, we again use a nesting iterator:
 \begin{displaymath}
@@
 \end{displaymath}
-
-\begin{lemma}
-  (Under a plausible operational semantics) FIXME: use an equational
-  theory instead, that includes $\beta$-equality.
-  \begin{displaymath}
-    I_k\,(\underline{n},s) \longrightarrow^* (\underline{n},v)
-  \end{displaymath}
-  iff
-  \begin{displaymath}
-    f^{\binom{n}{k}}\,s \longrightarrow^* v
-  \end{displaymath}
-\end{lemma}
-
-\subsection{Explicit Resource Tracking}
-
-\begin{enumerate}
-\item We still require diamonds to build zero and successor, but the
-  iterator does not give them back.
-\item In the simply typed case, we can annotate types with the static
-  amount of potential they have per constructor.
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Complexity Control in Quantitative Type Theory}
+Unlike in the \emph{Cons-free} system, this iterator does not raise
+the degree of the nested iterator directly. Rather the iterator $I_k$
+performs $\binom{n}{k}$ iterations. As observed by Aelhig and
+Schwictenberg, this is sufficient because the binomials form a basis
+for the vector space of all polynomials.
+
+Despite this slightly more involved completeness construction, the
+advantage of LFPL is that it is now easy to have arbitrary iterable
+datatypes and to transform between them. We need only take the
+introduction and elimination rules for any inductive datatype and add
+$\Diamond$ premises to the introduction rules and $\Diamond$ bindings
+to the eliminators.
+
+% \subsection{Explicit Resource Tracking}
+
+% \begin{enumerate}
+% \item We still require diamonds to build zero and successor, but the
+%   iterator does not give them back.
+% \item In the simply typed case, we can annotate types with the static
+%   amount of potential they have per constructor.
+% \end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Quantitative Type Theory}
 \label{sec:qtt}
 
-We have now seen three systems for controlling resource usage by means
+We have now seen two systems for controlling resource usage by means
 of linear typing. We now look to extend these systems to include
@@
 \cite{atkey18qtt}. In this section, we review the definition of QTT
-and how we have adapted it to implicit and explicit control of
-computational complexity.
-
-\begin{enumerate}
-\item General description of QTT
-\item Non-iterable datatypes in QTT
-\item Adapting QTT to include a ``soft'' natural number type
-\item Diamonds
-\item
-\end{enumerate}
+and how we have adapted it to the polytime systems we saw in the
+previous section.
 
@@
 
-\begin{enumerate}
-\item Rules for contexts (using $\epsilon$ for empty contexts to avoid
-  confusion between $\diamond$ and $\Diamond$):
+\bob{Explain the judgements and general idea of QTT, and the admissibility of $0$-ing}
+
+\subsubsection{Contexts, variables, and conversion}
 \begin{mathpar}
@@
 \end{mathpar}
-\item Rules for $\Pi$ and $\Sigma$ types
+
+\subsubsection{$\Pi$- and $\Sigma$-types}
 \begin{mathpar}
@@
 
-  FIXME: $\Sigma$-types
-\item Rules for equality
+\bob{$\Sigma$-types, from the LICS paper}
+
+\subsubsection{The Identity Type}
 \begin{mathpar}
@@
 \end{mathpar}
-  \bob{Equality reflection?}
-\item Rules for universe
-\end{enumerate}
+\bob{Equality reflection}
+
+\subsubsection{Universe}
+\bob{Copy from LICS paper}
+
+\subsubsection{Datatypes} QTT, as we have presented it so far, has no
+interesting base types to perform computation on. Following our
+presentation of the simply typed linear systems in
+\autoref{sec:lin-intro}, we add two kinds of datatype to QTT. First,
+\autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
+construction and case analysis, but no recursion. In
+\autoref{sec:cons-free-qtt} we describe how to extend QTT to be a
+dependently typed adaptation of the Cons-free system of
+\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
+same treatment to the LFPL-style system.
 
 \subsection{Non-iterable Datatypes}
-
-\subsection{Cons-free Natural Numbers}
-
-Of course, ``cons-free'' only applies to the runtime portion of the
-system. In the $\sigma = 0$ fragment, we are allowed to create new
-natural numbers arbitrarily:
+\label{sec:noniter-qtt}
+
+\subsubsection{Booleans} The boolean type was already described in
+\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
+offer no possibility for iteration in any case, but it is useful to
+see how their typing rules handle resource distribution before moving
+to more complex types.
+\begin{mathpar}
+  \inferrule*
+  {\Gamma~\isctxt}
+  {\Gamma \vdash \BoolTy~\istype}
+
+  \inferrule*
+  {\Gamma~\isctxt}
+  {0\Gamma \vdash \cTrue, \cFalse \stackrel\sigma: \BoolTy}
+
+  \inferrule*
+  {0\Gamma_1, x \stackrel0: \BoolTy \vdash P~\istype \\
+    \Gamma_1 \vdash M \stackrel\sigma: \BoolTy \\
+    \Gamma_2 \vdash N_t \stackrel\sigma: P[\cTrue/x] \\
+    \Gamma_2 \vdash N_f \stackrel\sigma: P[\cFalse/x] \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \If_{x.P}\: M \: \Then \: N_t \: \Else \: N_f \stackrel\sigma: P[M/x]}
+\end{mathpar}
+The introduction rules for booleans both use a $0$-d context,
+indicating that construction of boolean values is free. Elimination of
+booleans via a dependently typed if-then-else is more subtle with its
+resource usage. The boolean to be eliminated must be constructed in a
+context $\Gamma_1$, while the two branches are constructed in context
+$\Gamma_2$. Since only one of the branches will be used, sharing
+resources between the branches is expected.
+
+One might wonder how, since constructing booleans is $0$-cost by their
+introduction rules, the $\Gamma_1$ context will ever be non-$0$. This
+is resolved by observing that booleans may be the output of processes
+that consume time (e.g., the iteration constructs defined below), and
+so $\Gamma_1$ will represent a requirement that the necessary resource
+is provided.
+
+\subsubsection{Lists} Lists are a little more complex than booleans,
+because the $\cCons$ constructor takes two arguments, so their
+resource usage must be combined. The type formation and introduction
+rules are as follows:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma \vdash T~\istype}
+  {0\Gamma \vdash \ListTy(T)~\istype}
+
+  \inferrule*
+  {\Gamma \vdash T~\istype}
+  {0\Gamma \vdash \cNil \stackrel\sigma: \ListTy(T)}
+
+  \inferrule*
+  {\Gamma_1 \vdash M \stackrel\sigma: T \\
+    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
+\end{mathpar}
+Lists do have the potential for iteration by their recursive nature,
+but in order to ensure the complexity guarantees we only permit
+matching without recursion in the full calculus. Here is the rule for
+dependently typed case analysis on lists:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
+    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
+    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
+    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} : P[M/x]}
+\end{mathpar}
+In the $\sigma = 0$ fragment, however, we are free to iterate on lists
+because computations in this fragment are only meant for type-level
+computation, not for the program itself. Put another way, the type
+checker may perform arbitary recursion on lists to type check the
+program, but the program itself may not do so without correctly
+accounting its costs as described in the following sections. The
+$\sigma = 0$ recursor for lists has the following typing rule, which
+is the standard dependent eliminator for lists, with everything
+annotated as $0$ usage.
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
+    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
+    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
+    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
+  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
+\end{mathpar}
+
+\subsection{Cons-free Natural Numbers and their Recursor}
+\label{sec:cons-free-qtt}
+
+\bob{Reword this paragraph in light of whatever is said in the
+  previous section.} The datatypes of the previous section still only
+allow us to write programs in the $\sigma = 1$ fragment that are
+constant time in the size of their input. If we are handed a list of
+an unknown length, we can only explore it to a fixed depth determined
+statically. As we saw above, to write programs that perform work
+proportional to the length of their input, we need some form of
+iterable datatype. In both our cons-free and LFPL-style systems, we
+use a natural number datatype.
+
+The cons-free system cannot allow the programmer to construct natural
+numbers in the $\sigma = 1$ fragment, as this would violate the
+complexity guarantees. However, we can use the flexibility of QTT to
+allow free construction of naturals in the $\sigma = 0$ fragment,
+which allows us to use natural numbers freely in types. Therefore, we
+have the following introduction rules, only usable in the $\sigma = 0$
+fragment:
 \begin{mathpar}
@@
 \end{mathpar}
-Creating numbers in the $\sigma = 0$ fragment ensures that we can talk
-about natural numbers in our types, allowing us to state properties of
-our programs.\bob{Forward ref to actually doing this.}
-
-The eliminator:
+The cons-free system allows free duplication of complete natural
+numbers. This is accomplished by a special construct copying the
+simply linear typed rule we gave above:
+\begin{mathpar}
+  \inferrule*
+  {\Gamma \vdash M \stackrel\sigma: \tyNat}
+  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
+\end{mathpar}
+Anyone who has reasoned about the metatheory of or implemented a type
+checker for dependent types will view this rule with unease as it
+appears to grant the ability to construct non-canonical values of pair
+type, and consequently generate non-canonical naturals. We fix this by
+adding an equational rule to the $\sigma = 0$ fragment, ensuring
+\emph{definitionally} that $\mathrm{dupNat}$ acts as its name implies:
+\begin{mathpar}
+  \inferrule*
+  {\Gamma \vdash M \stackrel0: \tyNat}
+  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
+\end{mathpar}
+Note that this rule is well-typed by the $0$-needs-$0$ property of
+QTT, and the fact that $0+0=0$.
+
+The eliminator for these natural numbers takes the following
+form. Disregarding the usage annotations, this has the same type
+structure as the normal dependently typed recursor for natural
+numbers:
 \begin{displaymath}
@@
   \inferrule*
-  {0\Gamma, m \stackrel0: \tyNat \vdash X~\istype \\\\
-    0\Gamma \vdash M_z \stackrel\sigma: X[\conZero/m] \\\\
-    0\Gamma, n \stackrel0: \tyNat, x \stackrel1: X[n/m] \vdash M_s \stackrel\sigma: X[\conSucc(n)/m] \\\\
-    \Gamma \vdash N \stackrel\sigma: \tyNat}
-  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(n/x) \mapsto M_s\} \stackrel\sigma: X[N/m]}
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
+    0\Gamma \vdash N_z \stackrel\sigma: P[\conZero/x] \\\\
+    0\Gamma, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma: P[\conSucc(n)/x]}
+  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
 \end{displaymath}
-
-\subsection{Diamonds}
-
-\begin{mathpar}
-  \inferrule* [right=DiaType]
+In the successor case, $N_s$, there are two bound variables: $n$ for
+the natural number and $p$ for its induction hypothesis. Note that $n$
+is required to be usage $0$ no matter what $\sigma$ is. We need the
+variable $n$ to be present in order to correctly type the induction
+hypothesis and the conclusion, but it must be marked as usage $0$ to
+ensure that the resources captured by the number are not duplicated.
+
+The reader is invited to compare this dependently typed rule with the
+simply typed linear version in \autoref{sec:cons-free-intro}. Removing
+the $0$-annotated parts of the rule, and the type dependency, yield
+the exact same rule. Conversely, when $\sigma = 0$, this rule is
+identical (up to $0$-annotations) to the usual dependently typed
+recursor for natural numbers, and so we can use it in the types to
+prove properties of programs just as we do in normal type theory. We
+will see in \autoref{sec:cons-free-sound} that this rule is realisable
+by polynomial-time amortised computation, and so is sound for
+polynomial time. \bob{And we do some examples?}
+
+\subsection{LFPL-style Diamonds and Natural Numbers}
+\label{sec:lfpl-qtt}
+
+As explained in \autoref{sec:lfpl-intro}, the LFPL system differs from
+the Cons-free system in that it is possible to construct natural
+numbers (and other iterable datatypes), provided one has the necessary
+diamonds to pay for the construction. As with the natural number type
+in the Cons-free system, it ought not be possible to construct
+diamonds in the $\sigma = 1$ fragment, as this would amount to the
+free distribution of diamonds to all which would lead to a collapse in
+the complexity guarantees of the system. It is possible construct
+diamonds in the $\sigma = 0$, though:
+\begin{mathpar}
+  \inferrule*
   {\Gamma \vdash}
@@
 
-  \inferrule* [right=DiaIntro]
+  \inferrule*
   {\Gamma \vdash}
@@
 
-  \inferrule* [right=DiaEta]
+  \inferrule*
   {\Gamma \vdash M \stackrel0: \Diamond}
@@
 \end{mathpar}
-
-Link to Selinger et al: the $\eta$ rule for $\Diamond$s means that are
-essentially depending on the ``shape'' of anything that depends on
-diamonds.
-
-\subsection{LFPL-style Natural Numbers}
-
+The $\Diamond$ type also supports an $\eta$-rule in the $\sigma = 0$
+fragment, indicating that, in this fragment, it acts the same as a
+unit type. This allows us to use freely use diamonds in types, and to
+not have to care about the identity of particular diamonds, since by
+this rule all diamonds are definitionally
+equal\footnote{\emph{Fungible}, if one wishes to use a monetary
+  metaphor.}.
+% Link to Selinger et al: the $\eta$ rule for $\Diamond$s means that are
+% essentially depending on the ``shape'' of anything that depends on
+% diamonds.
+
+Construction of natural numbers now requires a $\Diamond$ for
+$\conZero$ and a $\Diamond$ and a predecessor for $\conSucc$:
 \begin{mathpar}
@@
 \end{mathpar}
-
+In the $\sigma = 0$ fragment, we can construct $\Diamond$s for free,
+and so construct natural numbers freely as well just as we did for the
+Cons-free system above.
+
+The dependently typed recursor for LFPL-style natural numbers again
+augments the simply typed linear recursor from
+\autoref{sec:lfpl-intro} with dependency information:
 \begin{mathpar}
@@
   \inferrule*
-  {0\Gamma, m \stackrel0: \tyNat \vdash X~\istype \\\\
-    0\Gamma, d \stackrel1: \Diamond \vdash M_z \stackrel\sigma: X[\conZero(*)/m] \\\\
-    0\Gamma, d \stackrel1: \Diamond, n \stackrel0: \tyNat, x \stackrel1: X[n/m] \vdash M_s \stackrel\sigma : X[\conSucc(*,n)/m] \\\\
-    \Gamma \vdash N \stackrel\sigma: \tyNat}
-  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,n/x) \mapsto M_s\} \stackrel\sigma: X[N/m]}
-\end{mathpar}
-
-Note: we have used $* : \Diamond$ as the value in the types for the
-zero and successor cases. By the $\eta$-rule for diamonds, we could
-have equally well used the $d$ variable that is in scope in each case.
-
-\subsection{Explicit Potential}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Soundness via Realisability}
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    \Gamma \vdash M \stackrel\sigma: \tyNat
+    0\Gamma, d \stackrel1: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
+    0\Gamma, d \stackrel1: \Diamond, n \stackrel0: \tyNat, p \stackrel1: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
+  {\Gamma \vdash \tmRec\,M\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
+\end{mathpar}
+We have used $* : \Diamond$ as the value in the types for the zero and
+successor cases. By the $\eta$-rule for diamonds, we could have
+equally well used the $d$ variable that is in scope in each case.
+
+Just as for the Cons-free system iterator above, in the $\sigma = 0$
+fragment this rule is identical to the usual dependently typed
+recursor for the natural numbers, so it can be used in the types to
+reason about programs. Moreover, we will see in
+\autoref{sec:lfpl-sound} that this rule is also sound for polynomial
+time in a system with $\Diamond$s.
+
+\subsection{Reflection of Realisability}
+\label{sec:qtt-reflection}
+
+Our final addition to QTT is \emph{reflection of realisability}. In
+QTT thus far, it has been possible to reason about the non-resourced
+behaviour of programs. This is because the $0$-ing process moving from
+the $\sigma = 1$ fragment to the $\sigma = 0$ fragment erases all
+resource information. This is sufficient for reasoning about the
+extensional behaviour of programs, but it is useful to be able to make
+statements like ``this function is realisable in polynomial time'' in
+the types of QTT, something that is not current possible.
+
+We remedy this by adding a \emph{realisable} type to QTT with the
+following type formation and introduction and elimination rules:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma \vdash A~\istype}
+  {0\Gamma \vdash \Rtype(A)~\istype}
+
+  \inferrule*
+  {0\Gamma \vdash M \stackrel1: A}
+  {0\Gamma \vdash \rIntro(M) \stackrel\sigma: \Rtype(A)}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel\sigma: \Rtype(A)}
+  {\Gamma \vdash \rElim(M) \stackrel{\sigma'}: A}
+\end{mathpar}
+Intuitively, the type $\Rtype(A)$ is inhabited whenever the type $A$
+is realisable in the $\sigma = 1$ fragment of the system. In
+particular, the type $\Rtype(\tyNat \to \tyNat)$ is the type of all
+realisable functions from natural numbers to natural numbers. In the
+polynomial time systems we are concerned with here, this is exactly
+the type of polynomial functions. Note that in the introduction rule,
+the premise is required to be in the $\sigma = 1$ fragment, to ensure
+that the type is realisable, while in the elimination rule, the
+conclusion is in an arbitrary fragment $\sigma'$. This flexibility is
+require to maintain the admissibility of the $0$-ing rule.
+
+With just the rules given here, the type $\Rtype(A)$ is nothing more
+than a statement that a given type is realisable with a polytime
+implementation. This is enough to do some proofs that we present in
+the next section, e.g., that polytime functions are closed under
+composition, but one could imagine stronger reflection principles that
+allow deeper logical consequences of polytime realisability to be
+proved internally. We discuss this further in
+\autoref{sec:towards-synthetic-complexity-theory}.
+
+Readers familiar with Benton's Linear/Non-linear system
+\cite{benton94} will note that the $\Rtype(A)$ constructor is the QTT
+analogue of the right adjont $G$ type constructor in that system. The
+$\Sigma$-types describe above play the role of the left adjoint $F$
+types, in a similar way to the dependent linear type system of Pradic,
+Krishnaswami and Benton \cite{pradic}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Programming and Proving with Polytime}
+\label{sec:programming-polytime}
+
+\begin{enumerate}
+\item Non-deterministic polytime via a bit oracle -- needs a statement
+  of success
+\item Previous could also be interpreted as probabilistic polytime
+  (access to coin flips); can we formalise some kind of cryptographic
+  property this way? Have a look at hash functions in the
+  Authenticated Data Structures paper.
+\item Insertion sort, quicksort as polytime functions.
+\item A type of polytime functions; via a reflection operator
+\item Programming with explicit resource annotations, including
+  reasoning about the resource requirements directly.
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Polytime Soundness via Realisability}
 \label{sec:soundness}
@@
 
-FIXME: introductory paragraph.
+\bob{introductory paragraph.}
 
@@
 \end{mathpar}
-  \caption{Language with Big-step Costed Evaluation Semantics}
+  \caption{Language with CBV Big-step Costed Evaluation Semantics}
   \label{fig:opsem}
@@
 
+\bob{move this to below the definition of resource monoid}
+
 An alternative perspective on resource monoids is to see them as a
@@
 
-FIXME: Relational definition, used for the formalisation.
-
 \subsection{Models of Quantitative Type Theory}
@@
 \begin{definition}
-  A \emph{linear poset} is a partially ordered set $(A, \leq)$ with
+  A \emph{linear preorder} is a preordered set $(A, \leq)$ with
   \begin{enumerate}
@@
     $x \leq y \multimap z$.
-  \end{enumerate}
-  The collection of all linear posets and functions that preserve the
-  order and the operations forms a category $\LinPoset$.
+  \item Right adjoint
+  \end{enumerate}
+  The collection of all linear preorders and functions that preserve the
+  order and the operations forms a category $\LinPreorder$.
 \end{definition}
@@
     $\Pi$ and $\Sigma$ types.
-  \item A functor $L : \cat{C}^\op \to \LinPoset$, such that
+  \item A functor $L : \cat{C}^\op \to \LinPreorder$, such that
     reindexing along projections has a right adjoint:
@@
   natural numbers.
-\item Assume a functor $L : \cat{C}^\top \to \LinPoset$ with
+\item Assume a functor $L : \cat{C}^\top \to \LinPreorder$ with
   reindexing along projections having a right adjoint.
@@
 
-\begin{proposition}\label{prop:basic-qtt-realisability-model}
-  Get a model QTT with $\Pi$, $\Sigma$, and a universe.
-\end{proposition}
-
 \subsubsection{Non-iterable Data Types}
@@
 show how non-iterable versions may be realised, we focus on two
-representative examples. Booleans
+representative examples. Booleans are the simplest case, with only two
+cases and no chance of iteration. Lists are more complex; we can have
+non-iterable lists containing iterable data. Combining lists and
+booleans allows us to represent other datatypes, such as the
+$\TapeType$ type used to represent Turing machine tapes in
+\autoref{sec:completeness}.
 
@@
 \begin{displaymath}
-  \mathrm{bool} = \{ (\mathit{tt}, \alpha, \synTrue) \mid \alpha \in M \} \cup \{ (\mathit{ff}, \alpha, \synFalse) \mid \alpha \in M \}
+  \mathrm{Bool} = \{ (\mathit{tt}, \alpha, \synTrue) \mid \alpha \in M \} \cup \{ (\mathit{ff}, \alpha, \synFalse) \mid \alpha \in M \}
 \end{displaymath}
@@
 
-\paragraph{Lists}
+Realisability of the construction and elimination of booleans amounts
+to the existence of the following inequalities. In an arbitrary fibre
+$L(A)$, we have $I_A \leq \mathit{tt}^* \mathrm{Bool}$ and
+$I_A \leq \mathit{ff}^* \mathrm{Bool}$ (treating $\mathit{tt}$ and
+$\mathit{ff}$ as constant functions $A \to \mathbb{B}$). These
+inequalities are realised by the corresponding $\synTrue/\synFalse$
+expression. For conditionals, the types involved are a little more
+complex to ensure agreement between boolean manipulations at the
+$\Set$-level and the realising computations. To get a realiser for a
+condition, we require a set $A$, an element $X \in L(A)$ (standing for
+the context) and an element $Y \in L(A \times \mathbb{B})$ (standing
+for the target type) and the existence in $L(A)$ of inequalities
+$X \leq (\lambda a. (a, \mathit{tt}))^* Y$, for the true case, and
+$X \leq (\lambda a. (a, \mathit{ff}))^* Y$, for the false case. When
+we have all these, we get in $L(A \times \mathbb{B})$ an inequality
+$\pi_1^* X \otimes \pi_2^* \mathrm{Bool} \leq Y$. \bob{back ref to the
+  typing rules for non-iterable data types}
+
+\paragraph{Lists} Lists are a little more involved, due to the need to
+explicitly manage a context that applies to all elements of the
+list. Let $\mathrm{List}(B)$ be the set of lists with elements from a
+set $B$. If we have $A : \Set$ and $B : A \to \Set$ and
+$X \in L(\Sigma a:A.\, B a)$, then we the resourced lists predicate
+$\mathrm{RList}(X) \in L(\Sigma a : A.\, \mathrm{List}(B a))$ must
+satisfy the equation:
+\begin{displaymath}
+  \mathrm{RList}(X) =
+  \begin{array}[t]{l}
+    \{ ((a, []), \alpha, (\synFalse, *)) \mid \alpha \in M \} \\
+    \cup \\
+    \{ \begin{array}[t]{@{}l}
+         ((a, b :: bs), \alpha, (\synTrue, (v_1, v_2))) \mid \\
+         \quad \exists \alpha_1, \alpha_2. 0 \leq M(\alpha, \alpha_1 \oplus \alpha_2) \land ((a, b), \alpha_1, v_1) \in X \land ((a, bs), \alpha_2, v_2) \in \mathrm{RList}(X) \}
+       \end{array}
+  \end{array}
+\end{displaymath}
+This equation has a least solution, by induction on the length of the
+list being realised. This definition is somewhat involved, but in
+essence states that a list is represented by tagged pairs, where
+$\synFalse$ represents nil and $\synTrue$ represents cons, and that
+the potential is distributed amongst the elements of the list as
+needed.
+
+Realisability of the nil and cons constructors, and of pattern
+matching on lists is similar to case for booleans. We refer the reader
+to the Agda development for the details.
+
+\subsubsection{Realising QTT}
+
+Summing up the previous sections, we have the following key theorem
+paving the way to soundness:
+\begin{theorem}\label{thm:basic-qtt-realisability-model}
+  For every resource monoid with a distinguished sub-monoid
+  $(M, M_0)$, we obtain a realisability model of $\mathbb{N}$-QTT with
+  $\Pi$ and $\Sigma$ types, a universe, and non-iterable datatypes.
+
+  \bob{State the realisability property of closed programs.}
+\end{theorem}
 
@@
 
-Explain why the polytime iterators work.
+The models constructed in the previous section only allow for
+constant-time programs to be realised.
+
+If we use the natural number resource monoid (\bob{Ref})...
+
+\subsection{Iteration Resource Monoids}
+
+To interpret iteration over a resource monoid $(M, M_0)$, we require
+additional structure, which we call an \emph{Iteration Resource
+  Monoid} to account for measurement of the sizes of iterable data
+structures and the effects of iteration on potentials.
+
+\subsubsection{Definition}
+We require:
+\begin{enumerate}
+\item a function $\mathit{size} : \mathbb{N} \to M$ that gives the
+  potential of an iterable data structure of a given
+  size;
+\item a function $\mathit{raise} : M \to M$ that raises the
+  (polynomial) degree of some potential; and
+\item a function $\mathit{scale} : \mathbb{N} \times M \to M$ that
+  scales a potential for a fixed number of iterations.
+\end{enumerate}
+These functions must satisfy the following properties:
+\begin{enumerate}
+\item $M_0$ is closed under the $\mathit{raise}$ operation;
+\item for all $\alpha$ and $n$,
+  $0 \leq M(\mathit{raise}(\alpha) \oplus \mathit{size}(n),
+  \mathit{scale}(n,\alpha) \oplus \mathit{size}(n))$; and
+\item for all $\alpha \in M_0$ and $n$,
+  $0 \leq M(\mathit{scale}(1 + n, \alpha), \alpha \oplus
+  \mathit{scale}(n, \alpha))$.
+\end{enumerate}
+The first property states that $\mathit{raise}$ is suitable as
+potential for whole programs2, meaning that it does not make any
+requirements on the existence of iterable data. Note that we do
+\emph{not} require $M_0$ to contain $\mathit{size}(n)$ -- programs
+themselves may not contain iterable data, all potential for iteration
+must be delivered externally. A useful intuition is that
+$\mathit{scale}(n,\alpha)$ represents the potential required for at
+most $n$ iterations that require potential $\alpha$, whereas
+$\mathit{raise}(\alpha)$ represents the potential required for a
+number of iterations that depends on the context. This is the
+motivation behind the second required property, which states that
+having $\mathit{raise}(\alpha)$ potential implies having
+$\mathit{scale}(n,\alpha)$ potential when the current size is $n$. The
+third property states that $\mathit{scale}$ decomposes as expected on
+potentials that do not include any size potential.
+
+Note that $\mathit{scale}(n,\alpha)$ is not the same as the action
+$n \cdot \alpha$ defined in \autoref{FIXME}. The latter operation
+scales both size and function potential, but the former only scales
+the function potential.
+
+\subsubsection{Polynomial Iteration Resource Monoids}
+
+Both of the polymonial resource monoids defined in
+\autoref{sec:polynomial-resource-monoids} support the structure of an
+Iteration Resource Monoid. We define:
+\begin{enumerate}
+\item $\mathit{size}(n) = (n, 0)$
+\item $\mathit{raise}(n, p) = (n, xp)$
+\item $\mathit{scale}(m, (n, p)) = (n, m \cdot p)$
+\end{enumerate}
+Note that $\mathit{raise}$ does indeed raise the degree of the
+polynomial involved. Property 2 above is satisfied because for any
+polynomial we have $(m \cdot p)(x) \leq (xp)(x)$ whenever $m \leq x$.
+
+\subsubsection{Realising Iterable Natural Numbers}
+
+For any natural number $n$, we define its representation as a value
+$\mathrm{natValue}(n) \in \ValSet$ by recursion.
+\begin{displaymath}
+  \begin{array}{lcl}
+    \mathrm{natValue}(0) &=& (\synTrue , *) \\
+    \mathrm{natValue}(1 + n) &=& (\synFalse, \mathrm{natValue}(n))
+  \end{array}
+\end{displaymath}
+This representation uses a tagged pair approach similar to our
+representation of lists in \autoref{sec:lists}. Using this, we can
+define what it means for a natural number to be realisable via
+$\mathrm{Nat} \in L(\mathbb{N})$:
+\begin{displaymath}
+  \mathrm{Nat} = \{(n, \alpha, \mathrm{natValue}(n)) \mid n \in \mathbb{N}, 0 \leq M(\alpha, \mathit{size}(n)) \}
+\end{displaymath}
+So a natural number $n$ is realised by the value
+$\mathrm{natValue}(n)$ as long as we have at least $\mathit{size}(n)$
+potential.  This gives us the ability to represent natural numbers as
+a type in QTT, but in order to iterate (and construct in the case of
+LFPL), we need to construct specific realisers for the Cons-Free and
+LFPL systems.
+
+\subsection{The Cons-Free System}
+
+The Cons-Free system uses the $\MaxPoly$ resource monoid, with the
+distinguished sub-monoid being those elements that are $0$ in the size
+component.
+
+\begin{enumerate}
+\item We can duplicate natural numbers
+\item We \emph{cannot} construct natural numbers
+\item We can construct a recursor
+\end{enumerate}
 
@@
   \label{thm:cons-free-soundness}
+  \bob{State the poly-time property for this system}
 \end{theorem}
+
+\subsection{The LFPL System}
+
+The LFPL system uses the $\PlusPoly$ resource monoid, with the
+distinguished sub-monoid again being those elements that are $0$ in
+the size component.
+
+\begin{enumerate}
+\item We can no longer duplicate natural numbers
+\item Define diamonds
+\item We can construct natural numbers given diamonds
+\item We can construct a recursor
+\end{enumerate}
 
@@
   \label{thm:lfpl-soundness}
+  \bob{State the poly-time property for this system}
 \end{theorem}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Realising Explicit Amortised Complexity}
-\label{sec:explicit-amortised-complexity}
+\subsection{Beyond Natural Numbers}
+
+\bob{Theoretically possible, bit useless in the Cons-Free system,
+  useful in the LFPL one}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Realising Explicit Amortised Complexity}
+% \label{sec:explicit-amortised-complexity}
 
@@
   \end{enumerate}
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Programming in Polytime}
-\label{sec:programming-polytime}
-
-\begin{enumerate}
-\item Non-deterministic polytime via a bit oracle -- needs a statement
-  of success
-\item Previous could also be interpreted as probabilistic polytime
-  (access to coin flips); can we formalise some kind of cryptographic
-  property this way? Have a look at hash functions in the
-  Authenticated Data Structures paper.
-\item Insertion sort, quicksort as polytime functions.
-\item A type of polytime functions; via a reflection operator
-\item Programming with explicit resource annotations, including
-  reasoning about the resource requirements directly.
 \end{enumerate}

### 2023-07-10 paper/paper.tex
[ paper ] lots more writing

@@
 \subsection{No Recursion, Only Case Analysis}
+\label{sec:linear-case-analysis}
 
@@
 \subsection{Diamond Trading with LFPL}
+\label{sec:lfpl-intro}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Quantitative Type Theory}
+\section{Polytime Quantitative Type Theory}
 \label{sec:qtt}
 
-We have now seen two systems for controlling resource usage by means
-of linear typing. We now look to extend these systems to include
-dependent types. Our motivation is threefold:
-\begin{enumerate}
-\item We would like to have a means for reasoning about the behaviour
-  of the programs we write in our systems. FIXME: why dependent types
-  then? Why not a program logic?
-\item In case of the explicitly resource accounted system, simple
-  typing is limited in the reasoning the programmer can perform about
-  the resources required by a particular program. While we could in
-  principle
-\item As well as constructing proofs \emph{about} complexity
-  constrained programs, we would also be able to write constructive
-  proofs that are themselves of constrained complexity.
-\end{enumerate}
+We have now seen the \emph{Cons-free} and \emph{LFPL} systems for
+capturing polytime by means of linear typing and restricted
+iteration. We now look to extend these systems to include dependent
+types.
 
@@
 dependent types we use \emph{Quantitative Type Theory} (QTT)
-\cite{atkey18qtt}. In this section, we review the definition of QTT
-and how we have adapted it to the polytime systems we saw in the
-previous section.
+\cite{atkey18qtt}. We now review the definition of QTT and describe
+how we have adapted it to the polytime systems we saw in the previous
+section.
 
@@
 
-\bob{Explain the judgements and general idea of QTT, and the admissibility of $0$-ing}
+Integrating linear and dependent types is not straightforward due to
+the conflict between the linear typing discipline regarding presence
+of a variable as only bestowing the right to use it once, and the
+dependent typing regime that uses variables both in types (for
+specification purposes) and in terms (for computational purposes),
+syntactically yield multiple uses of the same variable.
+
+QTT is a system that resolves this conflict by recording usage of
+variables with annotations from a semiring. It sits in the general
+area of systems that use semiring annotatons to measure resource usage
+\cite{BLAH}. The key feature of QTT, originally due to
+\cite{mcbride16}, is that usage of variables in types counts for
+$0$-usage in terms of the semiring used. This allows us to use normal
+type theory as a specification language, while also enjoying the
+benefits of linear typing for programs. The term typing judgement of
+QTT has the following form:
+\begin{displaymath}
+  x_1 \stackrel{\rho_1}: S_1, \dots, x_n \stackrel{\rho_n}: S_n \vdash M \stackrel\sigma: T
+\end{displaymath}
+where the annotations $\rho_i$ are all from the semiring being
+used. The annotation $\sigma$ is either $0$ or $1$, indicating whether
+we are in the erased ($\sigma = 0$) fragment, where all the normal
+rules of type theory apply, or the in the non-erased (``present'',
+$\sigma = 1$) fragment, where a restricted typing discipline applies.
+As we shall see below in the cases of $\Sigma$-types, iterable types,
+and LFPL's $\Diamond$ type, the separation of QTT into these two
+fragments allows an expressive combination of reasoning using full
+type theory with the benefits of linear typing.
+
+In the remainder of this sub-section, we describe the core of QTT. As
+well as the term typing judgement given above, QTT also has judgements
+for well-formed contexts ($\Gamma~\isctxt$) and types
+($\Gamma \vdash T~\istype$), and definition equality of types and
+terms ($\Gamma \vdash S \equiv T~\istype$ and
+$\Gamma \vdash M \equiv N \stackrel\sigma: S$). It is an invariant of
+the system that types are always well-formed in a context with all
+annotations $0$, i.e., $\Gamma \vdash S~\istype$ implies
+$0\Gamma = \Gamma$. An important admissible rule of the system, along
+with weakening and substitution, is that of $0$-ing:
+\begin{displaymath}
+  \inferrule*
+  {\Gamma \vdash M \stackrel1: S}
+  {0\Gamma \vdash M \stackrel0: S}
+\end{displaymath}
+This rule allows us to take any term $M$ in the $\sigma = 1$ fragment
+and treat it as if it were in the $\sigma = 0$ fragment, and hence use
+it for specification purposes in types. As we add novel rules to QTT
+in the following sections, we will be careful to maintain the
+admissibility of this rule.
+
+\subsubsection{Natural-number usages}
+
+We use an instantiation of QTT with the natural number semiring, with
+the usual semiring structure of addition and multiplication. In a mild
+generalisation of the original presentation of QTT, we also allow
+sub-usaging via the \emph{reverse} ordering on the naturals. That is,
+if a variable is marked as usage $n$ and $m \geq n$, then we can also
+regard it as usage $m$. This makes the system more like affine linear
+logic, since $m \leq 0$ for all $m$, matching the system in
+\autoref{sec:linear-intro}. We do \emph{not} have an unrestricted
+usage $\omega$, since this would allow the possibility of unrestricted
+duplication, and hence violate our polytime soundness properties.
 
 \subsubsection{Contexts, variables, and conversion}
-\begin{mathpar}
-  \inferrule* [right=Ctxt-Emp]
+
+As we saw above, contexts in QTT are comprised of variable
+$\stackrel\rho:$ type triples, where $\rho$ is a natural number
+indicating how many times the variable $x$ is available for use in a
+$\sigma = 1$ term. There are two operations on raw contexts: scaling
+$\pi\Gamma$, which multiplies each $\rho$ in $\Gamma$ by $\pi$, and
+addition $\Gamma_1 + \Gamma_2$, which adds two contexts' usage
+annotations assuming that the lengths and types are equal.  The basic
+usage-annotation discipline of QTT is demonstrated by the context
+formation and variable rules:
+\begin{mathpar}
+  \inferrule*
   { }
@@
 
-  \inferrule* [right=Ctxt-Ext]
+  \inferrule*
   {\Gamma~\isctxt \\ 0\Gamma \vdash S~\istype}
   {\Gamma, x \stackrel\rho: S~\isctxt}
-\end{mathpar}
-
-\begin{mathpar}
-  \inferrule* [right=Tm-Var]
+
+  \inferrule*
   {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
   {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}
-
-  \inferrule* [right=Tm-Conv]
-  {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
-  {\Gamma \vdash M \stackrel\sigma: T}
-\end{mathpar}
+\end{mathpar}
+As with most dependent type theories, contexts are built inductively
+from the empty context $\epsilon$ and extension of a context by a
+variable with a type that is well-formed in the preceding
+context. Usage annotations $\rho$ on variables are arbitrary, and
+types are always judged in a $0$-annotated context. The variable rule
+marks unused variables as usage $0$ and the selected variable with
+usage $\sigma$.
+
+% \begin{mathpar}
+
+%   \inferrule* [right=Tm-Conv]
+%   {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
+%   {\Gamma \vdash M \stackrel\sigma: T}
+% \end{mathpar}
 
 \subsubsection{$\Pi$- and $\Sigma$-types}
-\begin{mathpar}
-  \inferrule* [right=$\Pi$-type]
+
+QTT's $\Pi$-types have the form $(x \stackrel\rho: S) \to T$,
+indicating functions that, in the $\sigma = 1$ fragment, use their
+arguments $\rho$-many times. The formation, introduction and
+elimination rules are similar to the standard ones, except for the
+addition of usage annotations:
+\begin{mathpar}
+  \inferrule*
   {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
@@
 
-  \inferrule* [right=$\Pi$-Intro]
+  \inferrule*
   {\Gamma, x \stackrel{\sigma\rho}: S \vdash M \stackrel\sigma: T}
@@
 
-  \inferrule* [right=$\Pi$-Elim]
+  \inferrule*
   {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
     \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
-    FIXME}
+    0\Gamma_1 = 0\Gamma_2 \\
+    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
   {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
 \end{mathpar}
-
-\bob{$\Sigma$-types, from the LICS paper}
+The side conditions on the elimination rule state that (i) both
+$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
+defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
+either the function does not use its argument, or we are in the
+$\sigma = 0$ fragment and everything is being erased.
+
+$\Sigma$-types are a little more involved, and demonstrate the
+flexibility in QTT in allowing additional power in the $\sigma = 0$
+fragment, where we do not need to care about
+implementability. Formation and introduction are given by the rules:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
+  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T}
+
+  \inferrule*
+  {\Gamma_1 \vdash M \stackrel{\sigma'}: S \\
+    \Gamma_2 \vdash N \stackrel\sigma: T[M/x] \\\\
+    0\Gamma_1 = 0\Gamma_2 \\
+    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)}
+  {\pi\Gamma_1 + \Gamma_2 \vdash (M, N) \stackrel\sigma: (x \stackrel\pi: S) \otimes T}
+\end{mathpar}
+As with $\Pi$-typed, the first component of a $\Sigma$ type is
+annotated with a usage for how many times it can be used, and this is
+respected by the introduction rule. Elimination of $\Sigma$-types
+depends on whether we are in the $\sigma = 0$ fragment or not. In the
+$\sigma = 0$ fragment, we are free to disregard usage restrictions,
+and use projections as normal:
+\begin{mathpar}
+  \inferrule*
+  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
+  {\Gamma \vdash \mathrm{fst}(M) \stackrel0: S}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
+  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
+\end{mathpar}
+Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
+we can use them as for normal type-theoretic constructions.  In the
+$\sigma = 1$ fragment, we must take into account the resource content
+of objects and use a pattern matching construct; the dependently typed
+analogue of the $\otimes$-eliminator in \autoref{sec:lineartype}:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
+    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
+    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
+\end{mathpar}
 
 \subsubsection{The Identity Type}
-\begin{mathpar}
-  \inferrule* [right=Eq-type]
+
+QTT also supports an extensional equality type with equality reflection:
+\begin{mathpar}
+  \inferrule*
   {0\Gamma \vdash S~\istype \\
@@
 
-  \inferrule* [right=Eq-Intro]
+  \inferrule*
   {\Gamma \vdash M \stackrel\sigma: S}
   {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
-\end{mathpar}
-\bob{Equality reflection}
-
-\subsubsection{Universe}
-\bob{Copy from LICS paper}
+
+  \inferrule*
+  {\Gamma \vdash N \stackrel\sigma: M_1 =_S M_2}
+  {\Gamma \vdash M_1 \equiv M_2 \stackrel\sigma: S}
+\end{mathpar}
+The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
+as the only proof of equality \cite{hoffmann99dependent}.
+
+\subsubsection{Universe} QTT can also be equipped in universe types,
+just as in standard type theory. We refer to the rules presented in
+\cite{atkey18qtt}.
 
@@
 presentation of the simply typed linear systems in
-\autoref{sec:lin-intro}, we add two kinds of datatype to QTT. First,
+\autoref{sec:linear-intro}, we add two kinds of datatype to QTT. First,
 \autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
@@
 
-\subsubsection{Booleans} The boolean type was already described in
+\subsubsection{Booleans} The boolean type for QTT was described in
 \cite{atkey18qtt}, but we repeat it here for completeness. Booleans
-offer no possibility for iteration in any case, but it is useful to
-see how their typing rules handle resource distribution before moving
-to more complex types.
+offer no possibility for iteration, but it is useful to see how the
+QTT versions of the rules extend the simply typed rules from
+\autoref{sec:linear-case-analysis} before moving to more complex
+types.
 \begin{mathpar}
@@
 
-\bob{Reword this paragraph in light of whatever is said in the
-  previous section.} The datatypes of the previous section still only
-allow us to write programs in the $\sigma = 1$ fragment that are
-constant time in the size of their input. If we are handed a list of
-an unknown length, we can only explore it to a fixed depth determined
-statically. As we saw above, to write programs that perform work
-proportional to the length of their input, we need some form of
-iterable datatype. In both our cons-free and LFPL-style systems, we
-use a natural number datatype.
+The datatypes of the previous section still only allow us to write
+programs in the $\sigma = 1$ fragment that are constant time in the
+size of their input. As with the simply typed linear system, if we are
+handed a list of an unknown length, we can only explore it to a fixed
+depth determined statically. To write programs that do work
+proportional to the size of their input, we need some form of iterable
+datatype. In both our Cons-free and LFPL-style QTT systems, we use a
+natural number datatype.
 
@@
 by polynomial-time amortised computation, and so is sound for
-polynomial time. \bob{And we do some examples?}
+polynomial time.
 
@@
 \subsection{The Cons-Free System}
+\label{sec:cons-free-sound}
 
@@
 \subsection{The LFPL System}
+\label{sec:lfpl-sound}
 

### 2023-07-11 paper/paper.tex
[ paper ] various writing

@@
-\documentclass[acmsmall,review]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review,screen]{acmart}
+\settopmatter{printfolios=true,printccs=true,printacmref=true}
 
@@
 \renewcommand{\subsubsectionautorefname}{Section}
+
+\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
+\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
+\newcommand{\judgeref}[1]{\hyperref[#1]{Judgement \ref*{#1}}}
+\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
+\newcommand{\diagref}[1]{\hyperref[#1]{Diagram \ref*{#1}}}
+\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
+\newcommand{\defref}[1]{\hyperref[#1]{Definition \ref*{#1}}}
+
+\newcommand{\AgdaModule}[1]{\textcolor{blue}{\tt #1}}
 
@@
 \begin{abstract}
-  We combine dependent types for reasoning about programs with linear
-  type systems for implicit polynomial time.
+  We combine dependent types with linear type systems that soundly and
+  completely capture polynomial time computation. To combine dependent
+  and linear types, we use \emph{Quantitative Type Theory}. We explore
+  two systems for capturing polynomial time: one system that disallows
+  construction of iterable data, and one, based on the LFPL system of
+  Martin Hofmann, that controls construction via a payment
+  method. Both of these are extend to full dependent types, allowing
+  for arbitrary computation in types alongside guaranteed polynomial
+  time computation in terms.
+
+  The resulting system smoothly combines reasoning with dependent
+  types and restricted computational complexity. \bob{Elaborate after
+    Section 3 is written.}We prove the soundness of the systems using
+  a realisability technique due to Hofmann and Dal Lago.
 \end{abstract}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{type theory, implicit computational complexity, linear types}
+\keywords{type theory, implicit computational complexity, linear logic, computational complexity}
 
@@
 \subsection{Quantitative Type Theory}
+\label{sec:qtt-basic}
 
@@
 
+\bob{Usage weakening somewhere}
+
 % \begin{mathpar}
@@
   \inferrule*
-  {\Gamma \vdash N \stackrel\sigma: M_1 =_S M_2}
-  {\Gamma \vdash M_1 \equiv M_2 \stackrel\sigma: S}
+  {\Gamma \vdash N \stackrel0: M_1 =_S M_2}
+  {\Gamma \vdash M_1 \equiv M_2 \stackrel0: S}
 \end{mathpar}
 The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
-as the only proof of equality \cite{hoffmann99dependent}.
+as the only proof of equality \cite{hoffmann99dependent}. Note that
+equality reflection only targets the $\sigma = 0$ fragment, we cannot
+use equality reflection to convey realisability information.
 
@@
 
-\begin{enumerate}
-\item Non-deterministic polytime via a bit oracle -- needs a statement
-  of success
-\item Previous could also be interpreted as probabilistic polytime
-  (access to coin flips); can we formalise some kind of cryptographic
-  property this way? Have a look at hash functions in the
-  Authenticated Data Structures paper.
-\item Insertion sort, quicksort as polytime functions.
-\item A type of polytime functions; via a reflection operator
-\item Programming with explicit resource annotations, including
-  reasoning about the resource requirements directly.
-\end{enumerate}
+% \begin{enumerate}
+% \item Non-deterministic polytime via a bit oracle -- needs a statement
+%   of success
+% \item Previous could also be interpreted as probabilistic polytime
+%   (access to coin flips); can we formalise some kind of cryptographic
+%   property this way? Have a look at hash functions in the
+%   Authenticated Data Structures paper.
+% \item Insertion sort, quicksort as polytime functions.
+% \item A type of polytime functions; via a reflection operator
+% \item Programming with explicit resource annotations, including
+%   reasoning about the resource requirements directly.
+% \end{enumerate}
+
+\subsection{Building Datatypes}
+\label{sec:prog-datatypes}
+
+We have only defined an iterable natural number datatype for both of
+our systems above. Using the power of dependent types with a universe
+we can use these natural numbers to create further datatypes whose
+size is measured by a natural. Iteration on the size yields iteration
+over the full datastructure. For example, in the LFPL system, we can
+define a type of iterable lists by pairing a size with a type of
+elements defined by recursion:
+\begin{displaymath}
+  \mathrm{IList}(A) = \Sigma n \stackrel1: \tyNat.\,\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}
+\end{displaymath}
+
+Nil and Cons constructors.
+
+List iterator:
+
+\bob{U notation for the universe above}
+
+\begin{enumerate}
+\item Insertion sort in LFPL.
+\item Insertion sort in LFPL with correctness specification.
+\end{enumerate}
+
+\subsection{Polytime Problems}
+
+Uses
+
+\begin{displaymath}
+  (A : \mathsf{U}, P : A \to \mathsf{U})
+\end{displaymath}
+
+Cartesian container morphisms
+
+\subsection{Polytime-based Complexity Classes}
+
+Monads
+
+\subsubsection{Non-deterministic Polynomial Time}
+
+\subsubsection{Probabilistic Polynomial Time}
 
@@
 
-\newcommand{\cstaccess}{c_{\mathit{access}}}
-\newcommand{\cstmkclo}{c_{\mathit{mkclo}}}
-\newcommand{\cstapp}{c_{\mathit{app}}}
-\newcommand{\cstmkpair}{c_{\mathit{mkpair}}}
-\newcommand{\cstmkunit}{c_{\mathit{mkunit}}}
-\newcommand{\cstTrue}{c_{\mathit{mktrue}}}
-\newcommand{\cstFalse}{c_{\mathit{mkfalse}}}
-\newcommand{\cstLetpair}{c_{\mathit{letpair}}}
-\newcommand{\cstSeq}{c_{\mathit{seq}}}
-\newcommand{\cstIf}{c_{\mathit{if}}}
+\newcommand{\cstaccess}{1}
+\newcommand{\cstmkclo}{1}
+\newcommand{\cstapp}{1}
+\newcommand{\cstmkpair}{1}
+\newcommand{\cstmkunit}{1}
+\newcommand{\cstTrue}{1}
+\newcommand{\cstFalse}{1}
+\newcommand{\cstLetpair}{1}
+\newcommand{\cstSeq}{1}
+\newcommand{\cstIf}{1}
+
+% \newcommand{\cstaccess}{c_{\mathit{access}}}
+% \newcommand{\cstmkclo}{c_{\mathit{mkclo}}}
+% \newcommand{\cstapp}{c_{\mathit{app}}}
+% \newcommand{\cstmkpair}{c_{\mathit{mkpair}}}
+% \newcommand{\cstmkunit}{c_{\mathit{mkunit}}}
+% \newcommand{\cstTrue}{c_{\mathit{mktrue}}}
+% \newcommand{\cstFalse}{c_{\mathit{mkfalse}}}
+% \newcommand{\cstLetpair}{c_{\mathit{letpair}}}
+% \newcommand{\cstSeq}{c_{\mathit{seq}}}
+% \newcommand{\cstIf}{c_{\mathit{if}}}
 
@@
 
-\bob{introductory paragraph.}
-
-The Hofmann-Dal Lago approach to proving the soundness of implicit
-computational complexity systems is based on a three way coupling
+In this section and the next, we establish the polytime soundness of
+our extensions of QTT by adapting a realisability method due to
+\cite{hofmann-dallago}. This approach is based on a three way coupling
 between abstract mathematical elements (the \emph{what}), values from
@@
 
+\paragraph{Agda Formalisation} The key soundness results in this
+section have been formalised in the Agda proof assistant
+\cite{agda}. The Agda formalisation can be found in the supplementary
+material. After each definition and result we provide a pointer to the
+Agda modules where the corresponding formalisation can be found, and
+note interesting features of the mechanisation.
+
 \subsection{Machine Model and Operational Semantics}
@@
 
-Note: the Agda formalisation is intrinsically well-scoped, which
-makes expressing things a bit easier.
-
-The syntax and rules of our target language are given in
+We demonstrate that every program that can be written in extensions of
+QTT has the complexity bounds that we claim by translating QTT terms
+into an untyped CBV $\lambda$-calculus with a costed operational
+semantics. The syntax and rules of our target language are given in
 \autoref{fig:opsem}.
 
-\begin{enumerate}
-\item Justification that looking up variables is constant: Due to
-  lexical scoping, the size of the maximum variable lookup is bounded
-  for a fixed program.
-\item An advantage of our more ``elementary'' operational semantics
-  (rather than encoding everything in the $\lambda$-calculus)
-\end{enumerate}
+Variables are represented as de Bruijn indicies $i, j$. Expressions
+$E \in \ExpSet$ can be (anonymous) $\lambda$-abstractions, unit,
+pairing and boolean values, variables, sequencing, application, pair
+elimination, and conditionals. Note that, with the exception of
+$\lambda$-abstraction and sequencing, expressions never contain nested
+expressions; instead referring to variables already defined. Values
+$V \in \ValSet$ can be closures $\clo{E}{\eta}$, where $\eta$ is an
+environment for the closure, unit values, pairs and booleans.
+
+Costed evaluation of expressions in environments is defined by a
+big-step operational semantics $E, \eta \Downarrow_k V$, where $k$ is
+the number of steps. For simplicity, all operations cost $1$ unit,
+though this could be generalised to allow for different operations to
+have different costs. We use $\eta[i]$ to access the $i$th variable in
+the environment, counting from the right. The evaluation rules are
+mostly as one would expect, except for the application rule which
+includes a self reference to the closure being invoked, in order to
+allow recursive definitions.
+
+\paragraph{Agda Formalisation} The machine model is defined in the
+Agda module \AgdaModule{MachineModel}. We use an intrinsicaly
+well-scoped syntax, which ensures that all variable look up operations
+are always well defined.
 
@@
 
-Resource potentials are attached to values to represent the amount of
-intrinsic potential they have to fuel computation. Resource potentials
-are organised into resource monoids. To motivate the definition, we
-first enumerate the structure we will. To be able to account for the
-combined potential attached to composite data and programs (e.g.,
-pairs, or functions applied to arguments) we will require monoid
-structure on potentials. The action of turning potential difference
-into fuel for computation will be modelled by a difference
-function. Finally, we require that our resource monoid contains
-sufficient elements to fuel constant time operations. We gather these
-requirements into a formal definition as follows:
+As we mentioned above, resource potentials are attached to values to
+represent the amount of intrinsic potential they have to fuel
+computation. Resource potentials are organised into resource
+monoids. To be able to account for the combined potential attached to
+composite data and programs (e.g., pairs, or functions applied to
+arguments) we will require monoid structure on potentials. The action
+of turning potential difference into fuel for computation will be
+modelled by a difference function. Finally, we require that our
+resource monoid contains sufficient elements to fuel constant time
+operations. We gather these requirements into a formal definition as
+follows, which is a reformulation of the resource monoids of
+\cite{hoffman-dallago}:
 
@@
     we can add potentials.
-  \item A \emph{difference function} $M : |M| \times |M| \to \natinf$,
+  \item a \emph{difference function} $M : |M| \times |M| \to \natinf$,
     where $\natinf$ is the natural numbers extended with a negative
@@
     \item for all $\alpha, \beta, \gamma$,
-      $M(\alpha, \beta) \rplus M(\beta, \gamma) \leq M(\alpha, \gamma)$.
+      $M(\alpha, \beta) + M(\beta, \gamma) \leq M(\alpha, \gamma)$.
     \end{enumerate}
@@
     moving from $\alpha$ to $\gamma$ directly.
-  \item \bob{Interaction between $(\rplus,\rzero)$ and $M(\alpha,\beta)$.}
+  \item Differencing and the commutative monoid structure must satisfy:
+    \begin{enumerate}
+    \item $M(\alpha, \beta) \leq M(\alpha \rplus \gamma, \beta \rplus \gamma)$; and
+    \item $M(\alpha, \rzero) = 0$.
+    \end{enumerate}
   \item An \emph{accounting function}
@@
 \end{definition}
+
+For any resource monoid $M$, we can define an action of $\mathbb{N}$
+on $M$ as $n \cdot \alpha = \alpha \oplus \cdots \oplus \alpha$, where
+the right-hand side has $n$ summands.
+
+\paragraph{Alternative definition} Every resource monoid induces a
+partial ordering on its carrier set by $\alpha \leq \beta$ iff
+$0 \leq M(\alpha, \beta)$. Taking this idea further, we can
+reformulate a resource monoid as a symmetric monoidal category
+enriched in the symmetric monoidal category $\natinf$, where the
+monoid structure is addition. Isomorphism of objects $\alpha, \beta$
+in such a category amounts to having $M(\alpha, \beta) = 0$. The
+conditions in the definition above amount to the usual identity and
+composition laws for enriched categories.
+
+\paragraph{Agda Formalisation} Resource monoids are defined in the
+module \AgdaModule{Algebra.ResourceMonoid}. We use a formulation
+closer to the enriched category theory definition for the actual
+formalisation, because it avoids having to treat equality in the
+monoid structure separately from the induced preorder on elements.
 
@@
   resource monoid due to the requirement that the $\mathit{acct}$
-  function exists. \bob{Forward ref to where this is used.}
+  function must exist.
 \end{definition}
-
-Every resource monoid induces a partial ordering on its carrier set by
-$\alpha \leq \beta$ iff $0 \leq M(\alpha, \beta)$. In the case of the
-natural number resource monoid, this yields the usual ordering on the
-naturals.
 
@@
 The differencing operator of the natural number resource monoid can
-only supply as much fuel as is contained in the potential. This makes
-it suitable for amortised analysis, where we directly store potential
-in data structures. For the two polynomial time systems, we need more
-sophisticated structures originally presented by Dal Lago and
-Hofmann. The fundamental idea with both is to represent potentials as
-pairs $(m,p)$, where $m$ is a natural number and $p$ is a
-polynomial. The $m$ tracks the ``size'' of data as it pertains to the
-number of times an operation will be repeated by iterating over it ---
-for example, an iterable natural number will have size equal to
-itself, but a non-iterable natural number may be assigned zero
-size. The polynomial $p$ tracks the complexity of a program as a
-function of the size of the input. This leads to a differencing
-operator that evaluates the polynomial with the size of the data:
-
-\begin{definition}[Polynomial Resource Monoids]
-  The \emph{Max-Polynomial} resource monoid $\MaxPoly$ has carrier
-  set consisting of pairs $(m,p)$ where $m$ is natural number and $p$
-  is a polynomial with natural number coefficients. Addition of
-  elements is defined as $(m,p) \oplus (n,q) = (m \sqcup n, p+q)$ with
-  $\emptyset = (0,0)$. Difference is defined as:
+only supply as much fuel as is contained in the potential.
+% This makes
+% it suitable for amortised analysis, where we directly store potential
+% in data structures.
+For the two polynomial time systems, we need more sophisticated
+structures, both originally presented by Dal Lago and Hofmann. The
+fundamental idea with both is to represent potentials as pairs
+$(m,p)$, where $m$ is a natural number and $p$ is a polynomial. The
+$m$ tracks the ``size'' of data as it pertains to the number of times
+an operation will be repeated by iterating over it --- for example, an
+iterable natural number will have size equal to itself, but a
+non-iterable natural number may be assigned zero size. The polynomial
+$p$ tracks the complexity of a program as a function of the size of
+the input. This leads to a differencing operator that evaluates the
+polynomial with the size of the data:
+
+\begin{definition}[Polynomial Resource Monoids]\label{def:polynomial-resource-monoids}
+  The \emph{Max-Polynomial} resource monoid $\MaxPoly$ has carrier set
+  consisting of pairs $(m,p)$ where $m$ is natural number and $p$ is a
+  polynomial with natural number coefficients. Addition of elements is
+  defined as $(m,p) \oplus (n,q) = (m \sqcup n, p+q)$, where $\sqcup$
+  is the max operator, with $\emptyset = (0,0)$. Difference is defined
+  as:
   \begin{displaymath}
@@
 complexity $p$, then running the combination with no expectation of
-remaining potential yields $p(m)$ available steps.
-
-The $\MaxPoly$ and $\PlusPoly$ resource monoids will be used for the
-cons-free and LFPL-style (i.e., with diamonds) systems
-respectively. We will elaborate on how in
-\autoref{sec:realisability-model} and how they support the two kinds
-of natural number iteration in \autoref{sec:realising-iteration}.
-
-FIXME: difference between this definition and Hofmann-Dal Lago's: we
-use polynomials with natural number coefficients instead of a function
-that is bounded by a polynomial.
+remaining potential yields $p(m)$ available steps. The $\MaxPoly$ and
+$\PlusPoly$ resource monoids will be used for the \emph{Cons-free} and
+LFPL-style systems respectively, as we explain in
+\autoref{sec:realising-iteration} and show how these resource monoids
+yield the required polytime bounds on programs.
+
+\paragraph{Agda Formalisation} The natural number resource monoid is
+defined in the Agda module \AgdaModule{Algebra.ResourceMonoid.Nat} and the
+ polynomial monoids are both defined in
+\AgdaModule{Algebra.ResourceMonoid.Polynomial}. The definition is
+parameterised by the ``size monoid'' operation (either $\sqcup$ or
+$+$) used to compose sizes.
+
+% FIXME: difference between this definition and Hofmann-Dal Lago's: we
+% use polynomials with natural number coefficients instead of a function
+% that is bounded by a polynomial.
 
@@
 we will use for interpreting programs. We will call these sub-monoids
-$\MaxPoly_0$ and $\PlusPoly_0$. We make use of resource sub-monoids
-in our definitions of the interpretations of simultaneous
-substitutions (\autoref{defn:rl-morphism}) and terms
-(\autoref{defn:rl-terms}).
-
-\subsubsection{Resource Monoids as Enriched Categories}
-
-\bob{move this to below the definition of resource monoid}
-
-An alternative perspective on resource monoids is to see them as a
-kind of entriched category \cite{kelly}. Enriched categories replace
-the sets of morphisms between objects with objects from some category
-other than $\Set$. Define the posetal category $\natinf$ with objects $\mathbb{N} \cup \{\infty\}$ and
-
-$\natinf$ as a category.
-
-\bob{remember to look at the Marsden and thingy paper about
-  quantitative resources and enriched categories.}  A concise
-definition of Hofmann and Dal Lago's resource monoids can be given as:
-\begin{definition}
-  A \emph{resource monoid} is a strict symmetric monoidal
-  $\natinf$-enriched category.
-\end{definition}
-
-\subsection{Models of Quantitative Type Theory}
+$\MaxPoly_0$ and $\PlusPoly_0$.
+
+\subsection{Models of Quantitative Type Theory from Indexed Preorders}
 \label{sec:qtt-models}
 
-\bob{PLAN: RCwFs, and state that they can be constructed from Indexed Linear Posets}
-
-\cite{atkey18} describes a general class of QTT models termed
+\cite{atkey18qtt} described a general class of QTT models termed
 \emph{Quantitative Categories with Families} (QCwFs). This class of
-models fits will with the syntax of QTT, but to actually build models
-it is revealing to construct them from certain indexed partial orders,
-as we explain now. In the next section, we will construct a class of
-specific models that prove the soundness of our complexity constrained
-systems.
-
-\begin{definition}[Category with Families]
-  BLAH
-\end{definition}
-
-\begin{definition}
-  A \emph{linear preorder} is a preordered set $(A, \leq)$ with
-  \begin{enumerate}
-  \item A commutative monoid $(I, -\otimes-)$ that is monotone
-    w.r.t. the order; and
-  \item Is closed: there is an operation
-    $\multimap : A \times A \to A$ such that $x \otimes y \leq z$ iff
-    $x \leq y \multimap z$.
-  \item Right adjoint
+models fits well with the syntax of QTT, but to actually build models
+it is convenient to construct them from certain indexed linear
+preorders. The BCI-realisability model construction in
+\cite{atkey18qtt} is a special case of this construction, but most of
+the details are the same, so we only sketch a special case here to
+concentrate on the specifics of complexity bounded realisability.
+
+\begin{definition}\label{def:indexed-linear-preorder}
+  A \emph{$\mathbb{N}$-linear preorder} is a preordered set
+  $(L, \leq)$ with
+  \begin{enumerate}
+  \item a commutative monoid $(I, -\otimes-)$ that is monotone
+    w.r.t. the order;
+  \item is closed: there is an operation
+    $\multimap : L \times L \to L$ such that $x \otimes y \leq z$ iff
+    $x \leq y \multimap z$; and
+  \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
+    resource requirement adjustments, which satisfies the following
+    inequalities:
+    \begin{enumerate}
+    \item $\oc_0 X \simeq I$, for discarding;
+    \item $\oc_{m+n} X \leq (\oc_m X) \otimes (\oc_n X)$, for duplication;
+    \item $\oc_m\oc_n X \leq \oc_{mn} X$ for nesting;
+    \item $\oc_1 X \leq X$ for extraction / dereliction;
+    \item $(\oc_n X) \otimes (\oc_n Y) \leq \oc_n(X \otimes Y)$, for distribution; and
+    \item $n \leq m$ implies $\oc_n X \leq \oc_m X$, for usage weakening.
+    \end{enumerate}
   \end{enumerate}
@@
   order and the operations forms a category $\LinPreorder$.
+
+  An \emph{indexed linear preorder} $L : \Set^\op \to \LinPreorder$ is
+  a contravariant function, where we write $f^* : L(B) \to L(A)$ for
+  the action of $L$ on functions $f : A \to B$, such that such that
+  reindexing along projections has a right adjoint
+  $L_{\Sigma_{a \in A}.B}(\pi_1^*X, Y) \cong L_{A}(X, \forall_B Y)$
+  that commutes with reindexing.
 \end{definition}
 
-\begin{definition}
-  \begin{enumerate}
-  \item A CwF $(\cat{C}, \top, \langle-,-\rangle, ...)$ that supports
-    $\Pi$ and $\Sigma$ types.
-  \item A functor $L : \cat{C}^\op \to \LinPreorder$, such that
-    reindexing along projections has a right adjoint:
-    \begin{displaymath}
-      L_{\Gamma.A}(\pi^*X, Y) \cong L_{\Gamma}(X, \forall_A Y)
-    \end{displaymath}
-  \end{enumerate}
-\end{definition}
-
-\bob{define the construction}
-\begin{enumerate}
-\item Start with a CwF $\cat{C}$ that supports $\Pi$, $\Sigma$ and
-  natural numbers.
-\item Assume a functor $L : \cat{C}^\top \to \LinPreorder$ with
-  reindexing along projections having a right adjoint.
-\item Construct a new CwF with the same base category as $\cat{C}$,
-  but with $\Ty'(\Delta) = \Sigma A : \Ty(\Delta).~L(\Delta.A)$ and
-  $\Tm'(\Delta, (T, X)) = \Tm(\Delta, T)$.
-\item
-\end{enumerate}
-
-This construction is implicit in \cite{atkey2018qtt}, where the
-indexed linear poset is defined from a $\mathcal{R}$-LCA.
+Given an indexed linear preorder $L : \Set^\op \to \LinPreorder$, we
+construct a QCwF model of QTT with the basic type formers from
+\autoref{sec:qtt-basic}:
+\begin{enumerate}
+\item Define a category $\cat{L}$ of interpretations of contexts with
+  objects that are pairs $(A \in \Set, X \in L(A))$ and morphisms
+  $f : (A,X) \to (B, Y)$ that are functions $f : A \to B$ such that
+  $X \leq f^*Y$ (this is the Grothendieck category of $L$). There is a
+  faithful functor $U : \cat{L} \to \Set$. The category $\cat{L}$ will
+  be used for interpreting contexts in the $\sigma = 1$ fragment of
+  QTT.
+\item Define scaling of objects of $\cat{L}$ by
+  $\pi(A,X) = (A, \oc_\pi X)$, and addition of $(A, X)$ and $(A, Y)$
+  as $(A, X \otimes Y)$.
+\item For each set $A$, define the collection of semantics types
+  $\Ty(A)$ as the collection of $B : A \to \Set$ and
+  $X \in L(\Sigma_{a \in A}.B(a))$. Thus a QTT type consists of an
+  extensional meaning $B$ and its realisability specification $X$.
+\item For each $A$ and $(B,X) \in \Ty(A)$, the $\sigma = 0$ fragment
+  terms $\Tm(A,(B,X))$ are functions $\Pi_{a \in A}.\,B(a)$. For each
+  context interpretation $(A, X)$ in $\cat{L}$ and type intepretation
+  $(B,Y) \in \Ty(A)$, the $\sigma = 1$ fragment terms
+  $\RTm((A,X),(B,Y))$ are functions $f : \Pi_{a \in A}.\,B(a)$ such
+  that $X \leq \overline{f}^*Y$, where
+  $\overline{f} : A \to \Sigma_{a \in A}.\,B(a)$ is the section
+  associated with $f$.
+\item The empty context is interpreted as $(\{*\}, I)$ and context
+  extension $(A,X).(B,Y)$ (i.e., comprehension) by
+  $(\Sigma_{a \in A}.B(a), \pi_1^* X \otimes Y)$.
+\item Given $(A,X) \in \Ty(C)$ and
+  $(B,Y) \in \Ty(\Sigma_{c \in C}.A(c))$, $\Sigma$-types are
+  interpreted similarly to context extension and $\Pi$-types are
+  interpreted as
+  $(\lambda c.~(\Pi_{a \in A(c)}.\,B(c,a)), \forall_{A}(X \multimap
+  (\mathit{ev}\,f)^*Y))$, where
+  $\mathit{ev}\,f : (\Sigma_{c \in C}.A(c)) \to (\Sigma_{c \in
+    C}.\Sigma_{a \in A(c)}.B(c,a))$ is defined using application of
+  $f$.
+\item Universe and Equality types are interpreted as normal in $\Set$
+  with the realisability component set to $I$ in both cases. Note that
+  the universe of small types includes realisability information for
+  each type.
+\item Realisability reflection for a type $(B,X) \in \Ty(A)$ is
+  intepreted as the type
+  $(\lambda a. \{ b \in B(a) \mid I \leq (\lambda a. (a,b))^*X \},
+  I)$. Thus the set-component of the type is restricted to the
+  elements that are realisable, while the actual realisability
+  component is the ``empty'' $I$ realisability specification.
+\end{enumerate}
+
+\paragraph{Agda Formalisation} The indexed linear preorders are
+defined in the Agda module \AgdaModule{IndexedLinear}. We have not yet
+completed a formalisation of the construction of a full model of QTT
+from an indexed linear preorder, nor have we constructed the syntax of
+QTT in Agda, so these parts are currently unmechanised.
 
@@
 (\autoref{sec:machine-model}) and a compositional notion of resource
-potential (\autoref{sec:resource-monoids}), we construct models of QTT
-that witness the resource and type soundness of our complexity
-constrained systems.
-
-\bob{Tie back to previous section}
-
-We fix a resource monoid $M$ with distinguished sub-monoid $M_0$.
-
-\subsubsection{Base CwF}
-\label{sec:realisability-base-cwf}
-
-As our base CwF, we use the category $\Set$ of sets and functions with
-the standard interpretation of $\Pi$, $\Sigma$ types and the
-$\mathbb{N}$ type.
+potential (\autoref{sec:resource-monoids}), we can construct models of
+QTT that witness the resource and type soundness of our complexity
+constrained systems. We fix a resource monoid $M$ with distinguished
+sub-monoid $M_0$.
 
@@
 that ties together our ``mathematical'' model of types in $\Set$ with
-our machine model and resource monoid. For a set $A$, the carrier of
-$L(A)$ is the set of ternary relations
-$X \subseteq A \times M \times \ValSet$ and we define the ordering
-$X \leq Y$ to hold iff there exists realising expression
-$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
-$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
+our machine model and resource monoid. This construction is a
+reformulation of \cite{hoffmann-dallago} suitable for modelling
+dependent types. For a set $A$, the carrier of $L(A)$ is the set of
+ternary relations $X \subseteq A \times M \times \ValSet$ and we
+define the ordering $X \leq Y$ to hold iff there exists realising
+expression $E \in \ExpSet$ and potential $\gamma \in M_0$ such that
+for all $a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
 $(a,\alpha,v) \in X$, we have that there exists a result
@@
 $\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
-action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above \bob{do
-  this}. At the level of realising values, the modality $\oc_n$ has no
-effect; it only serves to alter the required potentials.
+action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above. At the
+level of the realising values, the modality $\oc_n$ has no effect; it
+only serves to alter the required potentials.
 \begin{displaymath}
@@
 need to give realisers for each of the required inequalities in
-\autoref{def:indexed-linear-preorder}. In each case, this is a matter
-of programming in the language of \autoref{sec:machine-model}. The
-corresponding potentials are calculated by counting the steps in the
-ensuing programs. Note that, so far, all realisers are constant time
-(relative to their input), so the constructions so far work for any
-resource monoid. For systems that require iteration, we will need more
-structure on our resource monoids \bob{fwd ref}.
-
-\begin{proposition}
-  $L$, with $I$, $\otimes$, $\multimap$, $\oc_n$, and $\forall_{B}$ defined above, is
-  an indexed linear preorder.
+\defref{def:indexed-linear-preorder}. In each case, this is a matter
+of programming in the language of \autoref{sec:machine-model}. For
+example, transitivity of the order is realised by sequencing of
+realising expressions. The corresponding potentials are calculated by
+counting the steps in the ensuing programs.
+
+\begin{proposition}\label{prop:realisability-model}
+  $L$, with $I$, $\otimes$, $\multimap$, $\oc_n$, and $\forall_{B}$
+  defined above, is an indexed linear preorder.
 \end{proposition}
 
-\begin{proof}
-  See the Agda development.
-\end{proof}
+\paragraph{Agda Formalisation} The construction of this indexed linear
+preorder and the proof of \propref{prop:realisability-model} are
+formalised in the Adga module
+\AgdaModule{AmortisedRealisabilityModel}.
 
 \subsubsection{Non-iterable Data Types}
-
-The model of QTT constructed in
-\autoref{prop:basic-qtt-realisability-model} does not yet include any
-useful base types. Handling iterable types, which are the ones that
-induce non-constant-time complexities, requires specific properties of
-resource monoids that we introduce in
+\label{sec:noniter-model}
+
+The model of QTT constructed in \propref{prop:realisability-model}
+does not yet include any useful base types. Handling iterable types,
+which are the ones that induce non-constant-time complexities,
+requires specific properties of resource monoids that we introduce in
 \autoref{sec:realising-iteration}.
@@
 Rather than perform a complex generic construction of datatypes to
-show how non-iterable versions may be realised, we focus on two
-representative examples. Booleans are the simplest case, with only two
-cases and no chance of iteration. Lists are more complex; we can have
-non-iterable lists containing iterable data. Combining lists and
-booleans allows us to represent other datatypes, such as the
-$\TapeType$ type used to represent Turing machine tapes in
-\autoref{sec:completeness}.
+show how non-iterable versions may be realised, we focus on the two
+representative examples from \autoref{sec:linear-case-analysis} and
+\autoref{sec:noniter-qtt}. Booleans are the simplest case, with only
+two cases and no chance of iteration. Lists are more complex; we can
+have non-iterable lists containing iterable data.
 
 \paragraph{Booleans} Fix $\mathbb{B} = \{ \mathit{tt}, \mathit{ff} \}$
-as our set of boolean elements. We define an element of $L(\mathbb{B}$
-to represent boolean values:
+as our set of boolean elements. We define an element of
+$L(\mathbb{B})$ to represent boolean values:
 \begin{displaymath}
@@
 we have all these, we get in $L(A \times \mathbb{B})$ an inequality
-$\pi_1^* X \otimes \pi_2^* \mathrm{Bool} \leq Y$. \bob{back ref to the
-  typing rules for non-iterable data types}
+$\pi_1^* X \otimes \pi_2^* \mathrm{Bool} \leq Y$. This construction
+suffices to realise the rules for QTT booleans in
+\autoref{sec:noniter-qtt}.
 
@@
 
+\paragraph{Agda Formalisation} The construction of realisers for
+booleans is carried out in the Agda module
+\AgdaModule{AmortisedModel.Bool}. The constructions for lists with
+case analysis are carried out in the module
+\AgdaModule{AmortisedModel.List}.
+
 \subsubsection{Realising QTT}
@@
 Summing up the previous sections, we have the following key theorem
-paving the way to soundness:
+paving the way to polytime soundness:
 \begin{theorem}\label{thm:basic-qtt-realisability-model}
@@
   $(M, M_0)$, we obtain a realisability model of $\mathbb{N}$-QTT with
-  $\Pi$ and $\Sigma$ types, a universe, and non-iterable datatypes.
-
-  \bob{State the realisability property of closed programs.}
+  $\Pi$ and $\Sigma$ types, a universe, extensional equality types,
+  realisability reflection, and non-iterable datatypes.
 \end{theorem}
@@
 The models constructed in the previous section only allow for
-constant-time programs to be realised.
-
-If we use the natural number resource monoid (\bob{Ref})...
+constant-time programs to be realised. To interpret the iterators of
+the \emph{Cons-free} and LFPL-style systems, we need to use the
+$\MaxPoly$ and $\PlusPoly$ resource monoids. We do this in this
+section, where first we establish some operations that will be useful
+to see how they capture the nesting of iterations inherent to polytime
+computation.
 
@@
 Note that $\mathit{scale}(n,\alpha)$ is not the same as the action
-$n \cdot \alpha$ defined in \autoref{FIXME}. The latter operation
-scales both size and function potential, but the former only scales
-the function potential.
+$n \cdot \alpha$ defined in \autoref{sec:resource-monoids}. The latter
+operation scales both size and function potential, but the former only
+scales the function potential.
 
@@
 Both of the polymonial resource monoids defined in
-\autoref{sec:polynomial-resource-monoids} support the structure of an
+\defref{def:polynomial-resource-monoids} support the structure of an
 Iteration Resource Monoid. We define:
@@
 This representation uses a tagged pair approach similar to our
-representation of lists in \autoref{sec:lists}. Using this, we can
-define what it means for a natural number to be realisable via
+representation of lists in \autoref{sec:noniter-model}. Using this, we
+can define what it means for a natural number to be realisable via
 $\mathrm{Nat} \in L(\mathbb{N})$:
 \begin{displaymath}
-  \mathrm{Nat} = \{(n, \alpha, \mathrm{natValue}(n)) \mid n \in \mathbb{N}, 0 \leq M(\alpha, \mathit{size}(n)) \}
+  \mathrm{Nat} = \{(n, \alpha, \mathrm{natValue}(n)) \mid n \in \mathbb{N}, 0 \leq M(\alpha, \mathit{size}(n+1)) \}
 \end{displaymath}
 So a natural number $n$ is realised by the value
-$\mathrm{natValue}(n)$ as long as we have at least $\mathit{size}(n)$
-potential.  This gives us the ability to represent natural numbers as
-a type in QTT, but in order to iterate (and construct in the case of
-LFPL), we need to construct specific realisers for the Cons-Free and
-LFPL systems.
+$\mathrm{natValue}(n)$ as long as we have at least
+$\mathit{size}(n+1)$ potential (we add one to make the LFPL soundness
+proof easier).  This gives us the ability to represent natural numbers
+as a type in QTT, but in order to iterate (and construct in the case
+of LFPL), we need to construct specific realisers for the
+\emph{Cons-free} and LFPL systems.
 
@@
 
-The Cons-Free system uses the $\MaxPoly$ resource monoid, with the
-distinguished sub-monoid being those elements that are $0$ in the size
-component.
-
-\begin{enumerate}
-\item We can duplicate natural numbers
-\item We \emph{cannot} construct natural numbers
-\item We can construct a recursor
-\end{enumerate}
-
+The \emph{Cons-free} system uses the $\MaxPoly$ resource monoid, with
+the distinguished sub-monoid being those elements that are $0$ in the
+size component. We enumerate the features of the \emph{Cons-free}
+system and justify their realisability with the $\MaxPoly$ resource
+monoid:
+\begin{enumerate}
+\item Duplication of natural numbers by $\dupNat(M)$ is realisable by
+  the expression $(0,0)$, which creates a pair by copying the input
+  variable twice. By the cost semantics in
+  \autoref{sec:machine-model}, this takes $1$ step of computation (we
+  assume that it is actually implemented via some pointer copy).  The
+  resource accounting for this realiser works because the size
+  component required for the output is the maximum of the size
+  components of the two elements, and since $n \sqcup n = n$, we have
+  enough resources to fulfil this.
+\item Construction of natural numbers is not realisable. In a putative
+  $\conSucc$ rule, we would need to get an additional unit of size
+  resource from nowhere.
+\item Iteration is realised by constructing a realising expression in
+  the expression language from the given expressions for the zero and
+  successor cases that uses the in-built recursion of the
+  language. The proof that resources are correctly accounted for is
+  carried out by induction on the natural being iterated over. For
+  $n$, we require potential
+  $\mathit{scale}(n,\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
+  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, where
+  $\gamma_{\mathit{succ}}$ and $\gamma_{\mathit{zero}}$ are the
+  potentials required by the successor and zero cases respectively. By
+  Property (2) of iteration resource monoids, above, we know that
+  $\mathit{raise}(\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
+  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$ always
+  dominates this requirement when paired with the potential
+  $\mathit{size}(n)$ from the input. Therefore, this latter
+  expression, plus some administrative set up costs, is the required
+  potential for the whole iterator.
+\end{enumerate}
+Together, we have a soundness result for the \emph{Cons-free} system,
+that ensures that every term in the $\sigma = 1$ fragment is
+realisable by a \emph{correct} program that terminates in polynomial
+time for all inputs:
 \begin{theorem}[Soundness for the Cons-free System]
   \label{thm:cons-free-soundness}
-  \bob{State the poly-time property for this system}
+  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
+  exists a realising expression $E$ and polynomial $p$ such that for
+  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
+  $k \in \mathbb{N}$ such that
+  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n)$ and $v$ is
+  a realising value for
+  $\llbracket M \rrbracket(n) \in \llbracket X \rrbracket(n)$.
 \end{theorem}
+
+\paragraph{Agda Formalisation} The realisability of the
+\emph{Cons-free} system iterator and the soundness property of the
+whole system are formalised in the Agda modules \AgdaModule{ConsFree}
+and \AgdaModule{ConsFree.Iterator}. The soundness theorem above is a
+combination of the inequality definition in our indexed linear
+preorder and the QTT model sketched in \autoref{sec:qtt-models}.
 
@@
 distinguished sub-monoid again being those elements that are $0$ in
-the size component.
-
-\begin{enumerate}
-\item We can no longer duplicate natural numbers
-\item Define diamonds
-\item We can construct natural numbers given diamonds
-\item We can construct a recursor
-\end{enumerate}
-
+the size component. With this resource monoid, the capabilities
+offered at the QTT level are altered:
+\begin{enumerate}
+\item We can no longer duplicate natural numbers, because
+  $\tyNat \otimes \tyNat$ requires twice as much size resource as
+  $\tyNat$, due to the combining operation on size potentials being
+  addition.
+\item We define the realisability specification for diamonds
+  $\Diamond \in L(1)$ to be:
+  \begin{displaymath}
+    \Diamond = \{ (*, \alpha, *) \mid 0 \leq M(\alpha, \mathit{size}(1)) \}
+  \end{displaymath}
+  Hence a diamond represents at least one unit of size resource,
+  matching the intuitive explanation given in
+  \autoref{sec:lfpl-intro}.
+\item With this definition of realisability for $\Diamond$s, it is
+  possible to realise the $\conZero$ and $\conSucc$ constructors for
+  natural numbers. By the additive combination of size resources we
+  get $1$ from the diamond and $n+1$ from the predecessor to total
+  $n+2$ for a new number. Note that, even if we add a $\Diamond$ type
+  to the \emph{Cons-free} system, it would still not be possible to
+  realise the constructors, because we would only have
+  $1 \sqcup (n+1) = n+1$ size resource for the output.
+\item The construction of the recursor follows a very similar proof to
+  the realisability of \emph{Cons-free} iterator, up to some
+  additional work to make sure that the dummy $*$ values representing
+  the diamond components end up in the right places. This additional
+  work is revealed in the required potential for the LFPL iterator
+  being
+  $\mathit{raise}(\mathit{acct}(8) \rplus \gamma_{\mathit{succ}})
+  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, so
+  slightly higher in the successor case.
+\end{enumerate}
+
+Soundness for the LFPL system is similar to the \emph{Cons-free}
+system, except for a $+1$ to the input to the polynomial, to account
+for the fact that we cost one size unit for the $\conZero$
+constructor.
 \begin{theorem}[Soundness for the LFPL-style System]
   \label{thm:lfpl-soundness}
-  \bob{State the poly-time property for this system}
+  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
+  exists a realising expression $E$ and polynomial $p$ such that for
+  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
+  $k \in \mathbb{N}$ such that
+  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n+1)$ and $v$
+  is a realising value for
+  $\llbracket M \rrbracket(n) \in \llbracket X \rrbracket(n)$.
 \end{theorem}
 
-\subsection{Beyond Natural Numbers}
-
-\bob{Theoretically possible, bit useless in the Cons-Free system,
-  useful in the LFPL one}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Realising Explicit Amortised Complexity}
-% \label{sec:explicit-amortised-complexity}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Relationship to Second-order MALL with Modalities}
-\label{sec:poly-mall}
-
-\begin{enumerate}
-\item Compare the SAL and LFPL modalities to the natural number
-  constructions
-  \begin{enumerate}
-  \item In the SAL case, we have $\oc A \multimap A^n$, which allows
-    us to
-  \end{enumerate}
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+The proofs of well-accounted realisability for the LFPL iterator, and
+the \emph{Cons-free} iterator, could be adapted to any other
+inductively defined type that is finitely branching. This is not
+immediately necessary, as evidenced by the construction of other
+datatypes in \autoref{sec:prog-datatypes}. Nevertheless, native tree
+type where the iterability is proportional to the total number of
+nodes would be useful.
+
+\paragraph{Agda Formalisation} The realisability of the LFPL system
+iterator and the soundness property of the whole system are formalised
+in the Agda modules \AgdaModule{LFPL} and \AgdaModule{LFPL.Iterator}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related and Future Work}
 \label{sec:conclusion}
 
-\begin{enumerate}
-\item Brunel's classical quantitative realisability
-\item Works on polytime realisability for proof extraction (Schwictenberg, Soft Set Theory)
+We have presented an two extensions of Quantitative Type Theory that
+soundly and completely captures polynomial time.
+
+\subsection{Systems for Polytime}
+
+BLL, SAL, LAL
+
+THe ones from Fossacs 2022, and POPL23(?) Romain Per...
+
+Neil Jones
+
+\subsection{Other Complexity Classes}
+
+\begin{enumerate}
 \item Lower complexity classes, like LOGSPACE
 \item Higher complexity classes, like EXPTIME, ELEMENTARY
-\item More fine-granied classes, like PTIME($k$)
-\item Automation, in the style of Jan's stuff
-\item Circuits and shit.
-\end{enumerate}
+\end{enumerate}
+
+\subsection{Explicit Resource Accounting}
+
+\begin{enumerate}
+\item RAML, Hofmann-Jost, etc.
+\item lambda-amor
+\item
+\end{enumerate}
+
+\subsection{Linear and Substructural Dependent Types}
+
+Linear LF
+
+Pradic et al
+
+Grtt
+
+Penn
+
+Abel
+
+\subsection{Implementation} Could use Idris, but it has the wrong
+semiring, and its datatype definition facility is too permissive.
+
+\subsection{Towards Synthetic Computational Complexity Theory?}
+
+Blah
 

### 2023-07-11 paper/paper.tex
[ paper ] some writing, mostly introduction

@@
-\documentclass[acmsmall,review,screen]{acmart}
+\documentclass[acmsmall,review,screen,anonymous]{acmart}
 \settopmatter{printfolios=true,printccs=true,printacmref=true}
@@
   for arbitrary computation in types alongside guaranteed polynomial
-  time computation in terms.
-
-  The resulting system smoothly combines reasoning with dependent
-  types and restricted computational complexity. \bob{Elaborate after
-    Section 3 is written.}We prove the soundness of the systems using
-  a realisability technique due to Hofmann and Dal Lago.
+  time computation in terms. We prove the soundness of the systems
+  using a realisability technique due to Hofmann and Dal Lago.
+
+  Our long-term goal is to combine the extensional reasoning of type
+  theory with intensional reasoning about the resources intrinsically
+  consumed by programs. This paper is a step along this path, which we
+  hope will lead both to practical systems for reasoning about
+  programs' resource usage, and to theoretical use as a form of
+  \emph{synthetic computational complexity theory}.
 \end{abstract}
@@
 
-{\bf New story:}
+Type theory is often claimed to be ideal for Computer Science,
+combining programming and proof in one unifying system, so a happy
+programmer can verify while they program, and program while they
+verify. From a broader Computer Science view, however, type theory
+lacks the ability to talk about the very thing that makes Computer
+Science interesting -- the fact that computation is everywhere bounded
+by the resources in time and space that we can afford to give it.
+
+Typical Type Theory only speaks of the public face that programs
+present to the world -- if you input things like this, you get things
+that look like that -- but cannot bring itself to mention the true
+cost of programs' execution. One can encode costs by embedding another
+programming language in type theory \cite{pottier}, or one can
+synthesise costs by treating resource counting as a computational
+effect \cite{daniellsson,calf}, but neither of these capture the
+intrinsic costs of the programs we write in type theory. These
+techniques deliver only \emph{conspicuous consumption}, not speaking
+of the real resources consumed.
+
+In this paper, we propose a method for extending dependent type theory
+with a means for constraining the intrinsic computational complexity
+of programs written in the theory. We concentrate on linear type
+systems that soundly and completely capture polynomial time
+computation, the commonly used standard for feasible resource usage,
+and extend these systems to dependent types. The additional
+expressivity of dependent types allows us use these characterisations
+of polytime to further functionally characterise
+
+We use techniques from \emph{Implicit Computational Complexity}
+theory, which seeks to provide intrinsic characterisations of
+complexity classes in terms of logical systems or programming
+languages. We review the techniques that we use in
+\autoref{sec:linear-intro}. To adapt these systems to dependent types,
+we use \emph{Quantitative Type Theory} (QTT)
+\cite{atkey18qtt,mcbride16}, a combination of linear and dependent
+types. We review QTT in \autoref{sec:qtt}.
+
+Our long-term goal is to combine the extensional reasoning of type
+theory with intensional reasoning about the resources intrinsically
+consumed by programs. This paper is a first step along this path,
+which we hope will lead both to practical systems for reasoning about
+programs' resource usage, and to theoretical use as a form of
+\emph{synthetic computational complexity theory}. We discuss these
+possibilities further in \autoref{sec:conclusion}.
+
+% The goal of this paper is to unify implicit computational complexity
+% with full dependent types. ICC enables us to ensure that
+% \emph{programs} are within a certain complexity class, while dependent
+% types allow us to state and prove that these programs solve the
+% problems that we intend them to, and to construct reductions between
+% problems. We can marry complexity-controlled implementations with
+% problems, paving the way for a \emph{synthetic computational
+%   complexity theory}.
+
+% \item Section 5: Extensions
+%   \begin{enumerate}
+%   \item Explicit Complexity via Diamond Counting
+%   \item Poly-time univalence
+%   \item Towards P- and NP-completeness proofs. Issue here is the
+%     extensionality. Towards a synthetic (poly-time based)
+%     computational complexity theory.
+%   \end{enumerate}
+% \item Section 6: Related Work
+%   \begin{enumerate}
+%   \item ICC stuff
+%     \begin{enumerate}
+%     \item LFPL
+%     \item CONS-free
+%     \item Soft Affine Logic; which has the same underlying idea as
+%       CONS-free (is this a novel observation?)
+%     \item Other systems based on information flow.
+%     \end{enumerate}
+%   \item Explicit complexity: lambda-amor, and things by Deepak Garg
+%     and Marco Gaboardi.
+%   \item Recent CMU stuff on calf
+%   \end{enumerate}
+
+\subsection{Contributions and Content}
+
+This paper makes the following contributions to the theory and use of
+linear dependent type theory and implicit computational complexity:
 \begin{enumerate}
-\item Introduction: The goal of this paper is to unify implicit
-  computational complexity with full dependent types. ICC enables us
-  to ensure that \emph{programs} are within a certain complexity
-  class, while dependent types allow us to state and prove that these
-  programs solve the problems that we intend them to, and to construct
-  reductions between problems. We can marry complexity-controlled
-  implementations with problems, paving the way for a \emph{synthetic
-    computational complexity theory}.
-\item Section 1: Linear types and Implicit Polynomial Time.
-  \begin{enumerate}
-  \item Total typed functional programming, as we see in type theory,
-    is restricted to the use of higher order functions and iteration
-    over inductive data types.
-  \item Polynomial time is the smallest class closed under nested
-    iteration over the input -- i.e. we consider iteration over the
-    input as feasible, and compositions thereof. The key thing is to
-    avoid iteration over data constructed from the input.
-  \item Key is to control iteration. So we split inductive datatypes
-    into iterable and non-iterable versions. We can
-  \item One way to capture polytime is to ensure that all iteration is
-    only done over the input. So we can duplicate iterable things, but
-    we can't construct new iterable things (at least not from an
-    iteration). We also can't duplicate higher order things, because
-    that effectively acts as a way to duplicate iterations (half
-    instantiated iterators). Compare to Jones' CONS-FREE programming.
-  \item Another approach is to allow ``borrowing'' of iterability by
-    means of diamonds. Iterability becomes a first-class object, so we
-    can use it to construct new data structures from the input. This
-    allows the construction of new iterable data structures. LFPL.
+\item We formulate two systems that combine linear type theory for
+  polytime computation with full dependent types, using Quantitative
+  Type Theory. The system is presented in \autoref{sec:qtt}. The
+  linear typing discipline required for enforcing polytime is provided
+  by QTT, but we also need to carefully add constructs for
+  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
+  of natural number iterator that we consider
+  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
+  the natural number iterators from the simply typed to the
+  dependently typed setting requires careful annotation of the rules
+  to ensure that the correct information is available for type
+  checking, while also not allowing too much information to be made
+  available at runtime that would violate the polytime soundness
+  property. A further contribution of this paper is the addition of
+  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
+  statements about polytime realisability to be reflected into
+  types. We explain in \autoref{Sec:realisability-future} how this
+  will have applications in QTT beyond the polytime systems presented
+  here.
+\item We demonstrate the utility of the combination of polytime and
+  dependent types in \autoref{sec:programming-polytime}. Just as in
+  the simply typed world, we have an expressive language for writing
+  polytime programs. With the additional power of dependent types, we
+  can also prove properties of these programs. A simple example is
+  proving that a sorting program actually sorts. Using QTT reflection,
+  we can go further and represent the class of polytime problems, with
+  polytime reductions between them, as dependent pairs
+  (\autoref{sec:polytime-problems}). Our final examples use dependent
+  types to give monadic presentations of the complexity classes of
+  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
+  (PP) time. Since these classes rely on specific correctness
+  criteria, it is not possible to capture them in a simply typed
+  system for polytime.
+\item We prove the polytime soundness of our systems via a
+  realisability argument in \autoref{sec:soundness}. Our construction
+  is an extension of the amortised complexity realisability
+  constructions of \cite{hofmann-dallago}. We extend their work to our
+  dependently typed setting, and also give a realisability
+  interpretation of datatypes directly, instead of via second-order
+  impredicative encodings. The technical content of this section has
+  been formalised in the Agda proof assistant \cite{agda}, and the
+  development is available in the supplementary material.
 \end{enumerate}
-\item Section 2: Extending to Dependent Types
-  \begin{enumerate}
-  \item Combining linear types and dependent types is subtle, because
-    there is a fundamental incompatibility in the syntax of the two
-    systems.
-  \item There have been several approaches, though they are all
-    somewhat similar (list them all here). We will use QTT, because it
-    offers a firm semantic foundation as well as a relatively
-    straightforward syntax.
-  \item Recap the syntax of QTT, with extensional equality. Key point
-    is that there is a normal dependently typed language with a linear
-    sub-language. We can have special constructs in the linear
-    sub-language that ``erase'' to normal type theory -- products are
-    the prime example.
-  \item Non iterable datatypes -- have normal eliminators in the DTT
-    fragment, but
-  \end{enumerate}
-\item Section 3: Programming and Proving with Poly-time
-  \begin{enumerate}
-  \item Internalised completeness proofs.
-  \item Construct the type of problems that are solvable in polynomial
-    time.
-  \item Construct reductions between problems.
-  \item NP. Doesn't make sense to talk about this without dependent
-    types, because we can't say when we've solved the problem being
-    stated.
-  \item PP (Probabilistic), using a monad. Again, need to be able to
-    state connection with the problem to know that we are solving it.
-  \end{enumerate}
-\item Section 4: Soundness via Realisability
-  \begin{enumerate}
-  \item The underlying CBV language and its
-  \end{enumerate}
-\item Section 5: Extensions
-  \begin{enumerate}
-  \item Explicit Complexity via Diamond Counting
-  \item Poly-time univalence
-  \item Towards P- and NP-completeness proofs. Issue here is the
-    extensionality. Towards a synthetic (poly-time based)
-    computational complexity theory.
-  \end{enumerate}
-\item Section 6: Related Work
-  \begin{enumerate}
-  \item ICC stuff
-    \begin{enumerate}
-    \item LFPL
-    \item CONS-free
-    \item Soft Affine Logic; which has the same underlying idea as
-      CONS-free (is this a novel observation?)
-    \item Other systems based on information flow.
-    \end{enumerate}
-  \item Explicit complexity: lambda-amor, and things by Deepak Garg
-    and Marco Gaboardi.
-  \item Recent CMU stuff on calf
-  \end{enumerate}
-
-\end{enumerate}
-
-{\bf Old story:}
-\begin{enumerate}
-\item There are systems for implicit computational complexity based on
-  linear logic (e.g. LFPL, SAL, LLL, BLL) that seek to capture
-  complexity classes (most often polynomial time) implicitly
-  (i.e. without directly talking about time bounds). This work has
-  also led to followup work on explicitly resourced calculi such as
-  RAML, which seek to use static analysis to determine complexity
-  bounds on programs.
-\item In dependent type theory, we can reason about the extensional
-  behaviour of programs (within the limits of intensional type
-  theory), but not about intenstional properties. The only way to
-  formalise something like ``for all polynomial time functions
-  $\mathbb{N} \to \mathbb{N}$'' is to develop a deep emebdding of
-  polynomial time functions and to work with that, which is tedious.
-\item The goal of this paper is to unify dependent types for reasoning
-  about programs with implicit computational complexity type
-  systems. In the first instance, we do not seek to use dependent
-  types to state precise bounds on programs' time consumption, but we
-  intend to use dependent types to reason about the extensional
-  behaviour of programs that we know to be polynomial time. At the end
-  of the paper, we show how to adapt our framework to explicit
-  resource accounting, in the style of RAML.
-\item The plan of the paper is as follows:
-  \begin{enumerate}
-  \item We first introduction the ideas behind the use of linear logic
-    for capturing polynomial time, by studying two systems Lafont's
-    SAL, and Hofmann's LFPL. In Hofmann and Dal Lago's presentation,
-    these are two variants of Second order Multiplicative Linear
-    Logic, with restricted $\oc$ modalities for duplication. We also
-    recall the realisability semantics of Dal Lago and Hofmann (FIXME:
-    move this later?)
-  \item The most modern presentations of these systems use second
-    order quantification to represent data types. This is incompatible
-    with the form of dependent types we use here, so we seek a direct
-    representation of natural numbers that captures
-  \item We then recall QTT, a extension of Martin-Löf type theory with
-    linear types, which can accomodate intensional restrictions on
-    computability as well as full reasoning about programs. Since
-    allowing an unrestricted $\oc$ modality would violate our
-    polynomial time bounds, we instead choose the natural number
-    semiring for resource accounting.
-  \item We then transplant the simply typed rules into QTT, showing
-    how to carefully distinguish between data that is available to the
-    program, and data that is only available for type formation.
-  \item We finish by showing that we can also do explicit resource
-    tracking.
-  \end{enumerate}
-\end{enumerate}
-
-Related work: only applications of Linear TT so far are erasure (Abel
-et al.) and quantum stuff (Selinger et al.), and (sort of) dependency
-analysis. This is another one.
+
+Before we get to the contributions above, we present, in
+\autoref{sec:linear-intro}, two linear simply typed systems for
+polytime, adaptations of systems already present in the
+literature. Our paper concludes with a discussion of further related
+work and the outlook for future work in \autoref{sec:conclusion}.
 
@@
 
-For this section, the linear $\lambda$-calculus we will use will have
-linear functions and $\otimes$-products:
+For this section, the affine linear $\lambda$-calculus we will use
+will have linear functions and $\otimes$-products:
 \begin{mathpar}
@@
   { }
-  {x : A \vdash x : A}
+  {\Gamma, x : A, \Gamma' \vdash x : A}
 
@@
 These rules are standard, so we do not describe them further except to
-note how linear typing uses presence or absences in a context to
-control resource usage -- if a variable is in the context it must be
-used exactly once. The fact that this discipline interferes with
-dependent types is one of the reasons we turn to QTT when we wish to
-add dependent types in \autoref{sec:qtt}.
+note how affine linear typing uses presence or absences in a context
+to control resource usage. If a variable is in the context it must be
+used at most once (variables that are not used are absorbed by the
+additional context in the variable rule). The fact that this
+discipline interferes with dependent types is one of the reasons we
+turn to QTT when we wish to add dependent types in \autoref{sec:qtt}.
 
@@
 of objects and use a pattern matching construct; the dependently typed
-analogue of the $\otimes$-eliminator in \autoref{sec:lineartype}:
+analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
 \begin{mathpar}
@@
 
-\subsubsection{Universe} QTT can also be equipped in universe types,
-just as in standard type theory. We refer to the rules presented in
-\cite{atkey18qtt}.
+\subsubsection{Universe} QTT can also be equipped in universe types
+$\mathsf{U}$, just as in standard type theory. We refer to the rules
+presented in \cite{atkey18qtt}.
 
@@
 \subsection{Polytime Problems}
+\label{sec:polytime-problems}
 
@@
 enriched in the symmetric monoidal category $\natinf$, where the
-monoid structure is addition. Isomorphism of objects $\alpha, \beta$
-in such a category amounts to having $M(\alpha, \beta) = 0$. The
-conditions in the definition above amount to the usual identity and
-composition laws for enriched categories.
+monoid structure is addition. The conditions in the definition above
+amount to the usual identity and composition laws for enriched
+categories.
 
@@
 
-\paragraph{Agda Formalisation} The natural number resource monoid is
-defined in the Agda module \AgdaModule{Algebra.ResourceMonoid.Nat} and the
- polynomial monoids are both defined in
+\paragraph{Agda Formalisation} The $\mathbb{N}$ resource monoid is
+defined in \AgdaModule{Algebra.ResourceMonoid.Nat} and the polynomial
+monoids are both defined in
 \AgdaModule{Algebra.ResourceMonoid.Polynomial}. The definition is
@@
 \begin{enumerate}
-\item RAML, Hofmann-Jost, etc.
-\item lambda-amor
-\item
+\item RAML, Hofmann-Jost, Raja, etc.
+\item lambda-amor,
+\item calf from CMU
 \end{enumerate}
@@
 Abel
+
+Related work: only applications of Linear TT so far are erasure (Abel
+et al.) and quantum stuff (Selinger et al.), and (sort of) dependency
+analysis. This is another one.
+
+What we didn't do: polytime typechecking guarantees.
 
@@
 
+\subsection{Towards Applications}
+
+Cryptography?
+
 \subsection{Towards Synthetic Computational Complexity Theory?}
+
+Synthetic computability theory Bauer, Forster
 

### 2023-07-11 paper/paper.tex
[ paper ] minor tweaks

@@
-\documentclass[acmsmall,review,screen,anonymous]{acmart}
+\documentclass[acmsmall,review,screen]{acmart}
 \settopmatter{printfolios=true,printccs=true,printacmref=true}
@@
 addition $\Gamma_1 + \Gamma_2$, which adds two contexts' usage
-annotations assuming that the lengths and types are equal.  The basic
-usage-annotation discipline of QTT is demonstrated by the context
-formation and variable rules:
+annotations assuming that the lengths and types are equal. Contexts
+are ordered pointwise $\Gamma' \sqsubseteq \Gamma$ on the usage
+annotations (which is the \emph{reverse} ordering on naturals) The
+basic usage-annotation discipline of QTT is demonstrated by the
+context formation and variable rules:
 \begin{mathpar}
@@
   {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}
+
+  \inferrule*
+  {\Gamma \vdash M : S \\ \Gamma' \sqsubseteq \Gamma}
+  {\Gamma' \vdash M : S}
 \end{mathpar}
@@
 usage $\sigma$.
-
-\bob{Usage weakening somewhere}
-
-% \begin{mathpar}
-
-%   \inferrule* [right=Tm-Conv]
-%   {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
-%   {\Gamma \vdash M \stackrel\sigma: T}
-% \end{mathpar}
 
@@
 amount to the usual identity and composition laws for enriched
-categories.
+categories. With this reading, we can see the value $M(\alpha,\beta)$
+when it is $\geq 0$ as the possibility of moving from $\alpha$ to
+$\beta$ levels of potential resource with some amount of residual
+resource emitted for computation; when it is $-\infty$, moving from
+$\alpha$ to $\beta$ is not possible.
 
@@
 formalisation, because it avoids having to treat equality in the
-monoid structure separately from the induced preorder on elements.
+monoid structure separately from the induced preorder on
+elements. Thinking of proofs involving the resource monoid as a
+process of finding a composable sequence of morphisms in a category
+was a helpful intuition when constructing the realisability model
+below.
 
@@
 Note that the definition of realisablity is uniform in the element $a$
--- the realising expression $e$ and the potential $\gamma$ must work
+-- the realising expression $E$ and the potential $\gamma$ must work
 for all $a$ -- thus the implementation and complexity measure of the
@@
 runtime. Moreover note that the potential $\gamma$ attached to the
-expression $e$ must come from the sub-monoid $M_0$, indicating that is
+expression $E$ must come from the sub-monoid $M_0$, indicating that is
 intended to be data-free, while the potential $\alpha$ for the input
@@
 We have presented an two extensions of Quantitative Type Theory that
-soundly and completely captures polynomial time.
+soundly and completely capture polynomial time. We have seen that this
+allows for an expressive combination of verification and complexity
+constrained computation. We now discuss some related work in implicit
+computational complexity and linear dependent types, and take a look
+at where the combination of resource conciousness and dependent types
+could take us.
 
@@
 
-BLL, SAL, LAL
-
-THe ones from Fossacs 2022, and POPL23(?) Romain Per...
+BLL, SAL, LAL, LFPL
+
+The ones from Fossacs 2022, and POPL23(?) Romain Per...
 
@@
 Synthetic computability theory Bauer, Forster
-
-Blah
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
+\bibliography{bibliography}
 

### 2023-07-11 paper/paper.tex
[ paper ] Complete programming bit, start on bibliography

@@
 %% the next command will enable that style.
-%%\citestyle{acmauthoryear}
+\citestyle{acmauthoryear}
 
@@
 \email{robert.atkey@strath.ac.uk}
-\orcid{1234-5678-9012}
+\orcid{0000-0002-4414-5047}
 \affiliation{%
@@
   time computation in terms. We prove the soundness of the systems
-  using a realisability technique due to Hofmann and Dal Lago.
+  using a realisability technique due to Dal Lago and Hofmann.
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{type theory, implicit computational complexity, linear logic, computational complexity}
+\keywords{type theory, implicit computational complexity, linear logic}
 
@@
 cost of programs' execution. One can encode costs by embedding another
-programming language in type theory \cite{pottier}, or one can
-synthesise costs by treating resource counting as a computational
-effect \cite{daniellsson,calf}, but neither of these capture the
-intrinsic costs of the programs we write in type theory. These
-techniques deliver only \emph{conspicuous consumption}, not speaking
-of the real resources consumed.
+programming language in type theory, for example \cite{GueneauCP18},
+or one can synthesise costs by treating resource counting as a
+computational effect, for example \cite{Danielsson08,NiuSGH22}, but
+neither of these capture the intrinsic costs of the programs we write
+in type theory. These techniques deliver only \emph{conspicuous
+  consumption}, not speaking of the real resources consumed.
 
@@
   statements about polytime realisability to be reflected into
-  types. We explain in \autoref{Sec:realisability-future} how this
-  will have applications in QTT beyond the polytime systems presented
-  here.
+  types. % We explain in \autoref{Sec:realisability-future} how this
+  % will have applications in QTT beyond the polytime systems presented
+  % here.
 \item We demonstrate the utility of the combination of polytime and
@@
   is an extension of the amortised complexity realisability
-  constructions of \cite{hofmann-dallago}. We extend their work to our
-  dependently typed setting, and also give a realisability
+  constructions of \citet{dallago11realisability}. We extend their
+  work to our dependently typed setting, and also give a realisability
   interpretation of datatypes directly, instead of via second-order
   impredicative encodings. The technical content of this section has
-  been formalised in the Agda proof assistant \cite{agda}, and the
-  development is available in the supplementary material.
+  been formalised in the Agda proof assistant
+  \cite{norell2008dependently}, and is included in the supplementary
+  material.
 \end{enumerate}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Linear $\lambda$-Calculus and Polytime}
+\section{Affine Linear $\lambda$-Calculus and Polytime}
 \label{sec:linear-intro}
@@
 types. For example, to simulate Turing machines, one can construct a
-$\mathit{Tape}$ type as a Zipper \cite{huet}
+$\mathrm{Tape}$ type as a Zipper \cite{huet}
 $\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
@@
 either the function does not use its argument, or we are in the
-$\sigma = 0$ fragment and everything is being erased.
+$\sigma = 0$ fragment and everything is being erased. In the
+following, when we write $S \to T$ for a non-dependent function type,
+we mean that the argument is to be used linearly:
+$(x \stackrel1: S) \to T$.
 
@@
 \end{mathpar}
+QTT also supports a unit type $I$ with constructor $*$ and
+pattern-matching elimination. See \cite{atkey18qtt} for details.
 
@@
   {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    \Gamma \vdash M \stackrel\sigma: \tyNat
-    0\Gamma, d \stackrel1: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
-    0\Gamma, d \stackrel1: \Diamond, n \stackrel0: \tyNat, p \stackrel1: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
+    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
   {\Gamma \vdash \tmRec\,M\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
@@
 \label{sec:programming-polytime}
+
+We now explore the possibilities afforded by the combination of
+polytime guarantees with the specification expressivity of dependent
+types.
 
@@
 We have only defined an iterable natural number datatype for both of
-our systems above. Using the power of dependent types with a universe
-we can use these natural numbers to create further datatypes whose
-size is measured by a natural. Iteration on the size yields iteration
-over the full datastructure. For example, in the LFPL system, we can
-define a type of iterable lists by pairing a size with a type of
-elements defined by recursion:
-\begin{displaymath}
-  \mathrm{IList}(A) = \Sigma n \stackrel1: \tyNat.\,\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}
-\end{displaymath}
-
-Nil and Cons constructors.
-
-List iterator:
-
-\bob{U notation for the universe above}
-
-\begin{enumerate}
-\item Insertion sort in LFPL.
-\item Insertion sort in LFPL with correctness specification.
-\end{enumerate}
+our systems above. We could extend both systems to include further
+iterable inductive types, although in the \emph{Cons-free} system this
+is not particularly useful due to the prohibition of on
+construction. However, sticking with just the natural numbers, we can
+use the power of dependent types with a universe to create further
+datatypes whose size is measured by some iterable natural
+number. Iteration on the size yields iteration over the full
+datastructure. For example, in the LFPL system, we can define a type
+of iterable lists by pairing a size with a type of elements defined by
+recursion on the size:
+\begin{displaymath}
+  \mathrm{IList}\,A = (n \stackrel1: \tyNat) \otimes \left(\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}\right)
+\end{displaymath}
+The $\cNil$ and $\cCons$ constructors can now be defined in terms of
+$\conZero$ and $\conSucc$, provided the caller supplies sufficient
+$\Diamond$s. These definitions live in the $\sigma = 1$ fragment, so
+we annotate them appropriately:
+\begin{mathpar}
+  \begin{array}{l}
+    \mathrm{nil} \stackrel1: \Diamond \to \mathrm{IList}\,A \\
+    \mathrm{nil}\,d = (\conZero(d), *)
+  \end{array}
+
+  \begin{array}{l}
+    \mathrm{cons} \stackrel1: \Diamond \to A \to \mathrm{IList}\,A \to \mathrm{IList}\,A \\
+    \mathrm{cons}\,d\,x\,\mathit{xs} = \mathrm{let}\,(n,\mathit{elems}) = \mathit{xs}\,\textrm{in}\,(\conSucc(d,n),(x,\mathit{elems}))
+  \end{array}
+\end{mathpar}
+Using the LFPL iterator it is also possible to construct a dependently
+typed iterator for $\mathrm{IList}\,A$ values. Unfortunately, the
+current types of the LFPL system are not sufficient to type this as a
+function, as we have no way of stating that the successor case must be
+arbitrarily duplicable. Lifting this restrict by means of some
+modality is future work. The typing rule for the derived list iterator
+is:
+\begin{mathpar}
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma \vdash A~\istype \\
+    0\Gamma, x \stackrel0: \mathrm{IList}\,A \vdash P~\istype \\\\
+    \Gamma \vdash M \stackrel\sigma: \mathrm{IList}\,A \\\\
+    0\Gamma, d \stackrel1: \Diamond \vdash N_1 \stackrel\sigma: P[\mathrm{nil}(*)/x] \\\\
+    0\Gamma, d \stackrel1: \Diamond, x \stackrel\sigma: A, \mathrm{xs} \stackrel0: \mathrm{IList}\,A, p \stackrel\sigma: P[\mathit{xs}/x] \vdash N_2 \stackrel\sigma : P[\mathrm{cons}(*,x,\mathit{xs})/x]}
+  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\mathrm{nil}(d) \mapsto N_1; \mathrm{cons}(d,x,\mathit{xs};p) \mapsto N_2\} \stackrel\sigma: P[M/x]}
+\end{mathpar}
+Note that, in the cons case, we have access to the result of iterating
+over the tail of the list ($p$), but not to the actual tail of the
+list ($\mathit{xs}$).
+
+With our list iterator, it is now possible to write interesting
+polytime programs. For example, insertion of a natural into a sorted
+list:
+\begin{displaymath}
+  \mathrm{insert} \stackrel1: \Diamond \to \tyNat \to \mathrm{IList}\,\tyNat \to \mathrm{IList}\,\tyNat
+\end{displaymath}
+which requires some ingenuity to write to handle the case where we
+find the place to insert the item and need access to the remainder of
+the list. Note that, also, the function consumes a $\Diamond$ to
+construct the new element of the output list, and also that the items
+in the list are themselves iterable natural numbers. This is needed to
+account for the
+
+Insertion sort is repeated insertion of elements from an original list
+into a new list. The new list is constructed from the $\Diamond$s
+yielded by the original list:
+\begin{displaymath}
+  \mathrm{insertionSort} \stackrel1: \mathrm{IList}\,A \to \mathrm{IList}\,A
+\end{displaymath}
+The immediate benefit of dependent types in this situation is that it
+is now possible to state and prove the correctness property of this
+sorting procedure. Using the fact that the $\sigma = 0$ fragment of
+QTT is exactly normal type theory, we can use normal dependently typed
+programmming techniques to establish:
+\begin{displaymath}
+  \mathrm{insertionSortCorrect} \stackrel0: (\mathit{xs} \stackrel1: \mathrm{IList}\,A) \to \mathrm{Sorted}(\mathit{xs},\mathrm{insertionSort}\,\mathrm{xs})
+\end{displaymath}
+where $\mathrm{Sorted}(x,y)$ is some predicate stating that $y$ is a
+sorted permutation of $x$.  Note that, despite the $1$ annotation on
+the $\Pi$-type here, we are free to duplicate $\mathit{xs}$ because
+types are constructed in the $\sigma = 0$ fragment.
 
@@
 
-Uses
-
-\begin{displaymath}
-  (A : \mathsf{U}, P : A \to \mathsf{U})
-\end{displaymath}
-
-Cartesian container morphisms
+Define a decision problem to be a pair $(A, P)$, where $A$ is a type
+in the universe $\mathsf{U}$, and $P : A \to \mathsf{U}$ is a
+predicate on $A$. For what follows, we are only interested in whether
+or not $P\,a$ is inhabited for each $a$. Therefore, we use
+$P \Leftrightarrow Q$ to stand for equi-inhabitation of two
+$P, Q : \mathsf{U}$, i.e.,
+$P \Leftrightarrow Q \equiv (P \to Q) \times (Q \to P)$.
+
+We can use the reflection type former defined in
+\autoref{sec:qtt-reflection} to define a predicate on decision
+problems that establishes whether or not they are polytime decision
+problems. Specifically, we can state that there is a polytime
+realisable boolean-value predicate that reports true exactly when the
+given element of $a$ is in the predicate:
+\begin{displaymath}
+  \mathrm{PTIME}(A,P) =
+    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
+\end{displaymath}
+Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
+decision problem $(A,P)$ is deciable in polytime. We make two notes
+about this definition: (i) this definition is intrinsic, in the sense
+that, whichever of the polytime systems is chosen, proving that a
+decision problem is solvable in polytime is a matter of programming,
+without having to reason directly about machine models and step
+counting; and (ii) we have defined problems to have arbitrary types
+$A$ as domains, rather than bitstrings, and so the notion of size
+attached to an input is intrinsic to the type $A$ chosen.
+
+We can also declare a type of polytime \emph{reductions} between
+problems. A problem $(A, P)$ can be polytime reduced to a problem
+$(B,Q)$ if there is an inhabitant of the following type:
+\begin{displaymath}
+  (A,P) \stackrel{\textsc{Poly}}\Rightarrow (B,Q) = (f \stackrel1: \mathbf{R}(A \to B)) \otimes (a \stackrel1: A) \to Q(\mathbf{R}^{-1}(f)\,a) \Leftrightarrow P\,a
+\end{displaymath}
+In words, there must be a polytime function $f$ that preserves and
+reflects decisions. With this definition, it is possible to prove in
+our systems that polytime computations are closed under polytime
+reductions. We note that this definition is, up to the reflection
+modality, the same as the definition of cartesian container morphism,
+well known in dependent type theory \cite{abbott03categories}, and
+speaks to a general conception of containers as ``problem/solution''
+pairings.
 
@@
 
-Monads
+The fact that we can characterise polytime decision problems is
+perhaps to be expected from a system designed to capture polynomial
+time realisable programs. However, we can go further to capture the
+complexity classes of Non-deterministic Polynomial time (NP) and
+Probabilistic Polynomial time (PP), both of which are based on
+polytime. We do this by augmenting our polytime functions with
+additional power in the form of computation effects.
 
@@
 
+To capture the complexity class NP, we use polynomial time programs
+augmented with non-determinism, as one might expect. We will not need
+to reason about equality of these non-deterministic programs, so we
+can represent non-deterministic choices as \emph{additive} binary
+trees. We suppose a non-iterable datatype defined like so:
+\begin{displaymath}
+  \begin{array}{l}
+    \textbf{data}\,\mathrm{ND}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
+    \quad \mathsf{return} : A \to \mathrm{ND}\,A \\
+    \quad \mathsf{choice} : (\BoolTy \to \mathrm{ND}\,A) \to \mathrm{ND}\,A
+  \end{array}
+\end{displaymath}
+The crucial point here is that the subtrees are represented as a
+function $\BoolTy \to \mathrm{ND}\,A$. By the typing rules of QTT,
+this means that the two branches of this function can share resources
+(see the encoding of the additive product types in
+\cite{atkey18qtt}). Thus, each branch of this tree is polynomial time,
+not the whole tree itself.
+
+The type $\mathrm{ND}$ supports a monad interface via the usual free
+monad construction, as well as an effect
+$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
+bit of non-deterministic information. Thus a program of type
+$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
+program with access to an oracle. In the $\sigma = 0$ fragment, we can
+write a function that resolves non-determinism using a list of
+booleans. This function returns $\mathsf{nothing}$ if the list of
+booleans is insufficient to resolve all the $\mathsf{choice}$s:
+\begin{displaymath}
+  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
+\end{displaymath}
+With these definitions, we can define Non-deterministic Polynomial
+time as a predicate on problems:
+\begin{displaymath}
+  \mathrm{NP}(A,P) =
+  \begin{array}[t]{l}
+    (f \stackrel1: \mathbf{R}(A \multimap \mathrm{ND}(\BoolTy)))\, \otimes \\
+    \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
+  \end{array}
+\end{displaymath}
+Thus, a problem is in NP if there is a non-deterministic
+boolean-valued polynomial time function that has a path to returning
+$\cTrue$ exactly when the input is in predicate. Moreover, it is a
+quick matter of programming to see that problems in NP are closed
+under the type of polytime reductions given above.
+
+% \begin{displaymath}
+%   \begin{array}{l}
+%     \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A \\
+%     \mathrm{runWithOracle}\,(\mathsf{return}\,a)\,\mathit{bs} = \mathsf{just}\,a \\
+%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cTrue :: \mathit{bs}) = \mathrm{runWithOracle}\,x\,\mathit{bs} \\
+%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cFalse :: \mathit{bs}) = \mathrm{runWithOracle}\,y\,\mathit{bs} \\
+%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,\cNil = \mathsf{nothing}
+%   \end{array}
+% \end{displaymath}
+
 \subsubsection{Probabilistic Polynomial Time}
+
+By changing the computation effects supplied to a program, we can
+change the complexity class. To capture Probabilistic Polynomial time,
+we use a (non-iterable) data structure representing trees of
+probabilistic choices, where $\mathbb{Q}[0,1]$ is some type of
+(non-iterable) rationals in the closed interval $[0,1]$:
+\begin{displaymath}
+  \begin{array}{l}
+    \textbf{data}\,\mathrm{Dist}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
+    \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
+    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy\ to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
+  \end{array}
+\end{displaymath}
+As in the non-deterministic case, a function $A \to \mathrm{Dist}\,B$
+in the $\sigma = 1$ fragment is a polytime probabilistic
+computation. Again, the use of a function type here ensures that each
+branch of the tree is constructable in polynomial time, not the whole
+tree. In the $\sigma = 0$ fragment we can write a function that
+computes the probability of a $\mathrm{Dist}\,\BoolTy$ computation
+being true:
+\begin{displaymath}
+  \mathrm{probTrue} \stackrel0: \mathrm{Dist}\,\BoolTy \to \mathbb{Q}[0,1]
+\end{displaymath}
+We can now define the class of probabilistic polynomial time decision
+problems, where the decider is allowed to make probabilistic choices
+as long as it is correct with probability at least $\frac{1}{3}$:
+\begin{displaymath}
+  \mathrm{PP}(A,P) =
+  \begin{array}[t]{l}
+    (f \stackrel1: \mathbf{R}(A \multimap \mathrm{Dist}(\BoolTy))) \, \otimes
+    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{1}{3}) \Leftrightarrow P\,a
+  \end{array}
+\end{displaymath}
+Again, problems in PP are easily seen to be closed under polytime
+reductions.
 
@@
 section have been formalised in the Agda proof assistant
-\cite{agda}. The Agda formalisation can be found in the supplementary
-material. After each definition and result we provide a pointer to the
-Agda modules where the corresponding formalisation can be found, and
-note interesting features of the mechanisation.
+\cite{norell2008dependently}. The Agda formalisation can be found in
+the supplementary material. After each definition and result we
+provide a pointer to the Agda modules where the corresponding
+formalisation can be found, and note interesting features of the
+mechanisation.
 
@@
 completed a formalisation of the construction of a full model of QTT
-from an indexed linear preorder, nor have we constructed the syntax of
-QTT in Agda, so these parts are currently unmechanised.
+from an indexed linear preorder so this part is currently unmechanised.
 
@@
 QTT that witness the resource and type soundness of our complexity
-constrained systems. We fix a resource monoid $M$ with distinguished
-sub-monoid $M_0$.
+constrained systems. We fix a resource monoid $M$ with sub-monoid
+$M_0$ and proceed to build an indexed linear preorder of resource
+accounted realisers.
 
@@
 
-We have presented an two extensions of Quantitative Type Theory that
+We have presented two extensions of Quantitative Type Theory that
 soundly and completely capture polynomial time. We have seen that this
@@
 
-The ones from Fossacs 2022, and POPL23(?) Romain Per...
+The ones from Fossacs 2022, and POPL23(?) Romain Perchoux
 
@@
 \item lambda-amor,
+\item Daniellson
 \item calf from CMU
@@
 \subsection{Towards Synthetic Computational Complexity Theory?}
+\label{sec:towards-synthetic-complexity-theory}
 

### 2023-07-11 paper/paper.tex
[ paper ] Complete paper?

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Polynomial Time and Full Dependent Types}
+\title{Polynomial Time and Dependent Types}
 
@@
   We combine dependent types with linear type systems that soundly and
-  completely capture polynomial time computation. To combine dependent
-  and linear types, we use \emph{Quantitative Type Theory}. We explore
-  two systems for capturing polynomial time: one system that disallows
+  completely capture polynomial time computation. We explore two
+  systems for capturing polynomial time: one system that disallows
   construction of iterable data, and one, based on the LFPL system of
   Martin Hofmann, that controls construction via a payment
-  method. Both of these are extend to full dependent types, allowing
-  for arbitrary computation in types alongside guaranteed polynomial
-  time computation in terms. We prove the soundness of the systems
-  using a realisability technique due to Dal Lago and Hofmann.
+  method. Both of these are extend to full dependent types via
+  Quantitative Type Theory, allowing for arbitrary computation in
+  types alongside guaranteed polynomial time computation in terms. We
+  prove the soundness of the systems using a realisability technique
+  due to Dal Lago and Hofmann.
 
@@
 expressivity of dependent types allows us use these characterisations
-of polytime to further functionally characterise
+of polytime to further functionally characterise the classes of
+non-deterministic and probabilistic polynomial time.
 
 We use techniques from \emph{Implicit Computational Complexity}
-theory, which seeks to provide intrinsic characterisations of
-complexity classes in terms of logical systems or programming
-languages. We review the techniques that we use in
-\autoref{sec:linear-intro}. To adapt these systems to dependent types,
-we use \emph{Quantitative Type Theory} (QTT)
-\cite{atkey18qtt,mcbride16}, a combination of linear and dependent
-types. We review QTT in \autoref{sec:qtt}.
-
-Our long-term goal is to combine the extensional reasoning of type
-theory with intensional reasoning about the resources intrinsically
+theory, which provides intrinsic characterisations of complexity
+classes in terms of logical systems or programming languages. We
+review the techniques that we use in \autoref{sec:linear-intro}. To
+adapt these systems to dependent types, we use \emph{Quantitative Type
+  Theory} (QTT) \cite{atkey18qtt,mcbride16}, a combination of linear
+and dependent types. We review QTT in \autoref{sec:qtt}.
+
+Our long-term goal is to combine the extensional reasoning of Type
+Theory with intensional reasoning about the resources intrinsically
 consumed by programs. This paper is a first step along this path,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Affine Linear $\lambda$-Calculus and Polytime}
+\section{Affine Linear Typing and Polytime}
 \label{sec:linear-intro}
@@
 that characterise polytime include Bounded Linear Logic (BLL), which
-uses explicit polynomials in the formulas \cite{bll} and Soft Affine
-Logic (SAL) \cite{lafont}, which does not explicitly represent
-time information in formulas, but uses a restricted form of Linear
-Logic's $\oc$ modality instead. Light Affine Logic (LAL) \cite{lal} is
+uses explicit polynomials in the formulas \cite{bll92} and Soft Affine
+Logic (SAL) \cite{Lafont04}, which does not explicitly represent time
+information in formulas, but uses a restricted form of Linear Logic's
+$\oc$ modality instead. Light Linear Logic (LLL) \cite{lll98} is
 another ``counting-free'' system for polytime.
@@
 programming languages that characterise complexity classes such as
-polytime. SAL has been transformed into a programming language for
-polytime by \cite{gaboardi}. Hofmann proposed a new programming
+polytime. SAL has been transformed into a programming language by
+\citet{BaillotM04}, and likewise for LLL by
+\citet{BaillotGM10}. \citet{hofmann99lfpl} proposed a new programming
 language, Linear Functional Programming Language (LFPL), that uses a
@@
 review the use of linear types to capture polytime by presenting two
-systems, one based on ideas from SALL and the second more explicitly
-based on LFPL. With a view to adapting these systems to dependent
-types in \autoref{sec:qtt}, we take an approach slightly different to
-much of the polytime linear logic literature. We emphasise the use of
-explicit datatypes and eliminators, rather than using impredicative
-encodings via universal types. Thus, we are closer to Hofmann's
-original LFPL (though not later presentations of it
-\cite{dallago-hofmann}) than BLL, SALL, or LAL.
+systems, one based on ideas from SAL and the second more explicitly
+based on LFPL. With a view to extending to dependent types in
+\autoref{sec:qtt}, we take an approach slightly different to much of
+the polytime linear logic literature. We emphasise the use of explicit
+datatypes and eliminators, rather than using impredicative encodings
+via universal types. We are closer to Hofmann's original LFPL (though
+not a later presentation of it by \citet{dallago11realisability}) than
+BLL, SAL or LLL.
 
@@
 to the size of the term. This is because every $\beta$-redex
-substitutes into exactly one variable, reducing the size of the term
+substitutes into at most one variable, reducing the size of the term
 by one each time.
@@
 preserved in these rules, and to foreshadow their dependently typed
-counterparts in \autoref{sec:qtt-noniter}. The first type is the
-booleans, which are anyway non-recursive and so wouldn't allow
+counterparts in \autoref{sec:noniter-qtt}. The first type is the
+booleans, which are anyway non-recursive and so would not allow
 iteration anyway:
@@
 \end{mathpar}
-Hence, we can construct lists arbitrarily, but we can only do case
-analysis on them. If we wish to explore a list to a arbitrary depth it
-must be driven by an iterable type of the form that we will see in the
-next two sections.
+We can construct lists arbitrarily but only do case analysis on
+them. If we wish to explore a list to a arbitrary depth it must be
+driven by a type we can iterate over.
 
@@
 types. For example, to simulate Turing machines, one can construct a
-$\mathrm{Tape}$ type as a Zipper \cite{huet}
+$\mathrm{Tape}$ type as a Zipper (\citet{Huet97})
 $\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
@@
 (we will see this in action in the completeness construction below and
-soundness proofs in \autoref{sec:iterator-soundness}).
+soundness proofs in \autoref{sec:realising-iteration}).
 
@@
 
-A more flexible system was proposed by \cite{hofmann99}. The
-\emph{Linear Functional Programming Language} (LFPL) takes a different
-approach to the \emph{Cons-free} system. Instead of completely
-prohibiting construction of new data, constructions is allowed if it
-is paid for by values of new type $\Diamond$ (``diamonds''):
+A more flexible system was proposed by \citet{hofmann99lfpl}. Instead
+of completely prohibiting construction of data, the \emph{Linear
+  Functional Programming Language} (LFPL) allows construction if it is
+paid for by values of type $\Diamond$ (``diamonds''):
 \begin{mathpar}
@@
 \end{mathpar}
-Thus, to construct a $\conZero$, we must have a $\Diamond$ to pay for
-it, and likewise, to construct a $\conSucc$ we must have a diamond. We
+To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
+and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
 can think of $\Diamond$s as an unit of iterable data. Iterability is
@@
 The difference with the \emph{Cons-free} iterator above is that the
-$\conZero$ and $\conSucc$ cases now both have an addition binding of
+$\conZero$ and $\conSucc$ cases now both have an additional binding of
 type $\Diamond$. This allows some form nesting of iterations: during
@@
 use for iteration over substructures that are smaller than the current
-point in the iteration. A construction, due to Aelhig and
-Schwictenberg \cite{syntactic-lfpl}, illustrates how this leads to all
-polytime computations. As above, we assume a polynomial $p(n)$ and a
-step function $f : \mathit{St} \multimap \mathit{St}$ that needs to be
-iterated $p(n)$ times. We construct a linear iterator:
+point in the iteration. A construction, due to \citet{AehligS02},
+illustrates how this leads to all polytime computations. As above, we
+assume a polynomial $p(n)$ and a step function
+$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
+$p(n)$ times. We construct a linear iterator:
 \begin{displaymath}
@@
 iteration. We now look to extend these systems to include dependent
-types.
-
-To integrate linear typing for controlling time complexity with
-dependent types we use \emph{Quantitative Type Theory} (QTT)
-\cite{atkey18qtt}. We now review the definition of QTT and describe
+types by building upon \emph{Quantitative Type Theory} (QTT)
+\cite{atkey18qtt,mcbride16}. This section reviews QTT and describes
 how we have adapted it to the polytime systems we saw in the previous
@@
 area of systems that use semiring annotatons to measure resource usage
-\cite{BLAH}. The key feature of QTT, originally due to
-\cite{mcbride16}, is that usage of variables in types counts for
-$0$-usage in terms of the semiring used. This allows us to use normal
-type theory as a specification language, while also enjoying the
-benefits of linear typing for programs. The term typing judgement of
-QTT has the following form:
+\cite{BrunelGMZ14,GhicaS14,OrchardLE19}. The key feature of QTT, an
+insight owing to \citet{mcbride16}, is that usage of variables in
+types counts for $0$-usage in terms of the semiring used. This allows
+us to use normal type theory as a specification language, while also
+enjoying the benefits of linear typing for programs. The term typing
+judgement of QTT has the following form:
 \begin{displaymath}
@@
 rules of type theory apply, or the in the non-erased (``present'',
-$\sigma = 1$) fragment, where a restricted typing discipline applies.
-As we shall see below in the cases of $\Sigma$-types, iterable types,
-and LFPL's $\Diamond$ type, the separation of QTT into these two
-fragments allows an expressive combination of reasoning using full
-type theory with the benefits of linear typing.
+``realisable'', $\sigma = 1$) fragment, where a restricted typing
+discipline applies.  As we shall see below in the cases of
+$\Sigma$-types, iterable types, and LFPL's $\Diamond$ type, the
+separation of QTT into these two fragments allows an expressive
+combination of reasoning using full type theory with the benefits of
+linear typing.
 
@@
 for well-formed contexts ($\Gamma~\isctxt$) and types
-($\Gamma \vdash T~\istype$), and definition equality of types and
+($\Gamma \vdash T~\istype$), and definitional equality of types and
 terms ($\Gamma \vdash S \equiv T~\istype$ and
@@
 we mean that the argument is to be used linearly:
-$(x \stackrel1: S) \to T$.
+$(x \stackrel1: S) \to T$, where $x$ does not appear in $T$.
 
@@
 flexibility in QTT in allowing additional power in the $\sigma = 0$
-fragment, where we do not need to care about
-implementability. Formation and introduction are given by the rules:
+fragment, where we do not need to care about polytime
+realisability. Formation and introduction are given by the rules:
 \begin{mathpar}
@@
 \end{mathpar}
-Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
-we can use them as for normal type-theoretic constructions.  In the
+$\Sigma$-types are unrestricted in the $\sigma = 0$ fragment, and we
+can use them as normal for type-theoretic constructions.  In the
 $\sigma = 1$ fragment, we must take into account the resource content
@@
 QTT also supports a unit type $I$ with constructor $*$ and
-pattern-matching elimination. See \cite{atkey18qtt} for details.
+pattern-matching \cite{atkey18qtt}.
 
@@
 The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
-as the only proof of equality \cite{hoffmann99dependent}. Note that
+as the only proof of equality \cite{hofmann97syntax}. Note that
 equality reflection only targets the $\sigma = 0$ fragment, we cannot
@@
 
-\subsubsection{Universe} QTT can also be equipped in universe types
-$\mathsf{U}$, just as in standard type theory. We refer to the rules
-presented in \cite{atkey18qtt}.
+\subsubsection{Universe} QTT has universe types $\mathsf{U}$, as in
+standard type theory \cite{atkey18qtt}.
 
@@
 presentation of the simply typed linear systems in
-\autoref{sec:linear-intro}, we add two kinds of datatype to QTT. First,
-\autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
-construction and case analysis, but no recursion. In
-\autoref{sec:cons-free-qtt} we describe how to extend QTT to be a
-dependently typed adaptation of the Cons-free system of
+\autoref{sec:linear-intro}, we add two kinds of datatype to
+QTT. First, in \autoref{sec:noniter-qtt}, we add non-iterable
+datatypes that allow construction and case analysis, but no
+recursion. In \autoref{sec:cons-free-qtt} we describe how to extend
+QTT to be a dependently typed adaptation of the Cons-free system of
 \autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
@@
 offer no possibility for iteration, but it is useful to see how the
-QTT versions of the rules extend the simply typed rules from
+QTT rules extend the simply typed rules from
 \autoref{sec:linear-case-analysis} before moving to more complex
@@
     0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} : P[M/x]}
+  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
 \end{mathpar}
@@
 prove properties of programs just as we do in normal type theory. We
-will see in \autoref{sec:cons-free-sound} that this rule is realisable
-by polynomial-time amortised computation, and so is sound for
-polynomial time.
-
-\subsection{LFPL-style Diamonds and Natural Numbers}
+will see in \autoref{sec:consfree-sound} that this rule is realisable
+by polynomial-time computation, and so is sound for polynomial time.
+
+\subsection{LFPL-style Diamonds, Natural Numbers, and a Recursor that Gives Back}
 \label{sec:lfpl-qtt}
@@
 resource information. This is sufficient for reasoning about the
-extensional behaviour of programs, but it is useful to be able to make
-statements like ``this function is realisable in polynomial time'' in
-the types of QTT, something that is not current possible.
+extensional behaviour of programs via types, but it is useful to be
+able to make statements like ``this function is realisable in
+polynomial time'' in the types of QTT, something that is not currently
+possible with the system we have seen so far.
 
@@
 
-With just the rules given here, the type $\Rtype(A)$ is nothing more
-than a statement that a given type is realisable with a polytime
-implementation. This is enough to do some proofs that we present in
-the next section, e.g., that polytime functions are closed under
+With just the rules given here, the type $\Rtype(A)$ is no more than a
+statement that a given type is realisable with a polytime
+implementation. This is enough to do the constructions that we present
+in the next section, e.g., that polytime functions are closed under
 composition, but one could imagine stronger reflection principles that
@@
 
-Readers familiar with Benton's Linear/Non-linear system
-\cite{benton94} will note that the $\Rtype(A)$ constructor is the QTT
-analogue of the right adjont $G$ type constructor in that system. The
-$\Sigma$-types describe above play the role of the left adjoint $F$
-types, in a similar way to the dependent linear type system of Pradic,
-Krishnaswami and Benton \cite{pradic}.
+Readers familiar with \citet{Benton94}'s Linear/Non-linear system will
+note that the $\Rtype(A)$ constructor is the QTT analogue of the right
+adjont $G$ type constructor in that system. The $\Sigma$-types play
+the role of the left adjoint $F$ types, in a similar way to the
+dependent linear type system of \citet{KrishnaswamiPB15}.
 
@@
 types.
-
-% \begin{enumerate}
-% \item Non-deterministic polytime via a bit oracle -- needs a statement
-%   of success
-% \item Previous could also be interpreted as probabilistic polytime
-%   (access to coin flips); can we formalise some kind of cryptographic
-%   property this way? Have a look at hash functions in the
-%   Authenticated Data Structures paper.
-% \item Insertion sort, quicksort as polytime functions.
-% \item A type of polytime functions; via a reflection operator
-% \item Programming with explicit resource annotations, including
-%   reasoning about the resource requirements directly.
-% \end{enumerate}
 
@@
 With our list iterator, it is now possible to write interesting
-polytime programs. For example, insertion of a natural into a sorted
-list:
+polytime programs. For example, the example used by \citet{Hofmann03}
+to demonstrate the expressivity of LFPL is insertion sort. First we
+define insertion of a natural into a sorted list:
 \begin{displaymath}
@@
 modality, the same as the definition of cartesian container morphism,
-well known in dependent type theory \cite{abbott03categories}, and
-speaks to a general conception of containers as ``problem/solution''
-pairings.
+well known in dependent type theory \cite{AbbottAG05}, and speaks to a
+general conception of containers as ``problem/solution'' pairings.
 
@@
 \subsubsection{Non-deterministic Polynomial Time}
+\label{sec:np-class}
 
@@
   \mathrm{NP}(A,P) =
-  \begin{array}[t]{l}
-    (f \stackrel1: \mathbf{R}(A \multimap \mathrm{ND}(\BoolTy)))\, \otimes \\
+  \begin{array}[t]{@{}l}
+    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
     \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
@@
   \begin{array}[t]{l}
-    (f \stackrel1: \mathbf{R}(A \multimap \mathrm{Dist}(\BoolTy))) \, \otimes
+    (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
     (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{1}{3}) \Leftrightarrow P\,a
@@
 our extensions of QTT by adapting a realisability method due to
-\cite{hofmann-dallago}. This approach is based on a three way coupling
-between abstract mathematical elements (the \emph{what}), values from
-a machine model (the \emph{how}), and resource potentials (the
-\emph{fuel}). Each type in the system is defined as a three way
+\citet{dallago11realisability}. This approach is based on a three way
+coupling between abstract mathematical elements (the \emph{what}),
+values from a machine model (the \emph{how}), and resource potentials
+(the \emph{fuel}). Each type in the system is defined as a three way
 relation between these elements. The set of abstract elements depends
@@
 into \emph{resource monoids} that we define in
-\autoref{sec:resource-monoids}. Following Hofmann and Dal Lago, we
-will select resource monoids appropriate to the kind of system that we
-want to prove complexity soundness for.
+\autoref{sec:resource-monoids}. Unlike \citet{dallago11realisability},
+we explicitly construct realisers for inductive datatypes (both
+iterable and non-iterable) instead of relying on second-order
+polymorphic encodings and special $\oc$-style modalities. These
+explicit constructions are essential to construct a model of dependent
+types.
 
@@
 \caption{Language with CBV Big-step Costed Evaluation Semantics}
+\Description{Syntax and evaluation rules of the untyped CBV
+  $\lambda$-calculus used for realisers.}
   \label{fig:opsem}
@@
 operations. We gather these requirements into a formal definition as
-follows, which is a reformulation of the resource monoids of
-\cite{hoffman-dallago}:
+follows, which is a slight reformulation of the resource monoids of
+\citet{dallago11realisability}:
 
@@
 \paragraph{Alternative definition} Every resource monoid induces a
-partial ordering on its carrier set by $\alpha \leq \beta$ iff
+pre-ordering on its carrier set by $\alpha \leq \beta$ iff
 $0 \leq M(\alpha, \beta)$. Taking this idea further, we can
@@
 \subsubsection{Specific Resource Monoids}
+\label{sec:specific-monoids}
 
@@
 
-\cite{atkey18qtt} described a general class of QTT models termed
+\citet{atkey18qtt} described a general class of QTT models termed
 \emph{Quantitative Categories with Families} (QCwFs). This class of
@@
 it is convenient to construct them from certain indexed linear
-preorders. The BCI-realisability model construction in
-\cite{atkey18qtt} is a special case of this construction, but most of
-the details are the same, so we only sketch a special case here to
-concentrate on the specifics of complexity bounded realisability.
+preorders. The BCI-realisability model construction given by
+\citet{atkey18qtt} is a special case of this construction, so we only
+sketch a special case here to concentrate on the specifics of
+complexity bounded realisability.
 
@@
   \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
-    resource requirement adjustments, which satisfies the following
-    inequalities:
+    resource requirement adjustments, satisfying:
     \begin{enumerate}
@@
 \item The empty context is interpreted as $(\{*\}, I)$ and context
-  extension $(A,X).(B,Y)$ (i.e., comprehension) by
-  $(\Sigma_{a \in A}.B(a), \pi_1^* X \otimes Y)$.
+  extension $(A,X).n(B,Y)$ (i.e., comprehension) by
+  $(\Sigma_{a \in A}.B(a), \pi_1^* X \otimes \oc_n Y)$.
 \item Given $(A,X) \in \Ty(C)$ and
@@
   with the realisability component set to $I$ in both cases. Note that
-  the universe of small types includes realisability information for
-  each type.
+  the universe of small types includes resource-relevant realisability
+  information for each type.
 \item Realisability reflection for a type $(B,X) \in \Ty(A)$ is
@@
 our machine model and resource monoid. This construction is a
-reformulation of \cite{hoffmann-dallago} suitable for modelling
-dependent types. For a set $A$, the carrier of $L(A)$ is the set of
-ternary relations $X \subseteq A \times M \times \ValSet$ and we
-define the ordering $X \leq Y$ to hold iff there exists realising
-expression $E \in \ExpSet$ and potential $\gamma \in M_0$ such that
-for all $a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
+reformulation of \citet{dallago11realisability}'s realisability models
+suitable for modelling dependent types. For a set $A$, the carrier of
+$L(A)$ is the set of ternary relations
+$X \subseteq A \times M \times \ValSet$ and we define the ordering
+$X \leq Y$ to hold iff there exists realising expression
+$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
+$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
 $(a,\alpha,v) \in X$, we have that there exists a result
@@
 \begin{displaymath}
-  \begin{array}{lcl}
+  \begin{array}{l@{\,}c@{\,}l}
     X \otimes Y &=& \{ (a, \alpha, (v_1, v_2)) \mid \exists \alpha_1, \alpha_2.~0 \leq M(\alpha, \alpha_1 \rplus \alpha_2) \land X(a,\alpha_1,v_1) \land Y(a,\alpha_2,v_2) \}\\
@@
 $\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
-action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above. At the
-level of the realising values, the modality $\oc_n$ has no effect; it
-only serves to alter the required potentials.
+action of $(\mathbb{N}, \leq)$ on $M$ defined above. The modality
+$\oc_n$ has no effect on realising values. It only serves to alter the
+resource potentials.
 \begin{displaymath}
@@
 example, transitivity of the order is realised by sequencing of
-realising expressions. The corresponding potentials are calculated by
-counting the steps in the ensuing programs.
+expressions. The potentials are calculated by counting the steps in
+the ensuing programs.
 
@@
 The model of QTT constructed in \propref{prop:realisability-model}
-does not yet include any useful base types. Handling iterable types,
-which are the ones that induce non-constant-time complexities,
-requires specific properties of resource monoids that we introduce in
+does not yet include any useful base types. Iterable types, which are
+the ones that induce non-constant time complexities, require specific
+properties of resource monoids that we introduce in
 \autoref{sec:realising-iteration}.
 
-Rather than perform a complex generic construction of datatypes to
-show how non-iterable versions may be realised, we focus on the two
-representative examples from \autoref{sec:linear-case-analysis} and
-\autoref{sec:noniter-qtt}. Booleans are the simplest case, with only
-two cases and no chance of iteration. Lists are more complex; we can
-have non-iterable lists containing iterable data.
+Before that, we show how to define realisers for the representative
+examples of non-iterable types from \autoref{sec:linear-case-analysis}
+and \autoref{sec:noniter-qtt}. Booleans are the simplest case, with
+only two cases and no chance of iteration. Lists are more complex; we
+can have non-iterable lists containing iterable data.
 
@@
 and $\mathit{ff}$ is represented by $\synFalse$. In both case, we
-allow arbitrary potentials to be attached.
+allow arbitrary potential $\alpha$ to be attached.
 
 Realisability of the construction and elimination of booleans amounts
-to the existence of the following inequalities. In an arbitrary fibre
+to the existence of the following inequalities. In any preorder
 $L(A)$, we have $I_A \leq \mathit{tt}^* \mathrm{Bool}$ and
@@
 $\Set$-level and the realising computations. To get a realiser for a
-condition, we require a set $A$, an element $X \in L(A)$ (standing for
+conditional, we require a set $A$, an element $X \in L(A)$ (standing for
 the context) and an element $Y \in L(A \times \mathbb{B})$ (standing
@@
 
-Realisability of the nil and cons constructors, and of pattern
-matching on lists is similar to case for booleans. We refer the reader
-to the Agda development for the details.
-
 \paragraph{Agda Formalisation} The construction of realisers for
-booleans is carried out in the Agda module
-\AgdaModule{AmortisedModel.Bool}. The constructions for lists with
-case analysis are carried out in the module
-\AgdaModule{AmortisedModel.List}.
-
-\subsubsection{Realising QTT}
-
-Summing up the previous sections, we have the following key theorem
-paving the way to polytime soundness:
-\begin{theorem}\label{thm:basic-qtt-realisability-model}
-  For every resource monoid with a distinguished sub-monoid
-  $(M, M_0)$, we obtain a realisability model of $\mathbb{N}$-QTT with
-  $\Pi$ and $\Sigma$ types, a universe, extensional equality types,
-  realisability reflection, and non-iterable datatypes.
-\end{theorem}
+booleans and lists are carried out in the Agda modules
+\AgdaModule{AmortisedModel.Bool} and \AgdaModule{AmortisedModel.List}.
+
+% \subsubsection{Realising QTT}
+
+% Summing up the previous sections, we have the following key theorem
+% paving the way to polytime soundness:
+% \begin{theorem}\label{thm:basic-qtt-realisability-model}
+%   For every resource monoid with a distinguished sub-monoid
+%   $(M, M_0)$, we obtain a realisability model of $\mathbb{N}$-QTT with
+%   $\Pi$ and $\Sigma$ types, a universe, extensional equality types,
+%   realisability reflection, and non-iterable datatypes.
+% \end{theorem}
 
@@
 For any natural number $n$, we define its representation as a value
-$\mathrm{natValue}(n) \in \ValSet$ by recursion.
-\begin{displaymath}
-  \begin{array}{lcl}
-    \mathrm{natValue}(0) &=& (\synTrue , *) \\
-    \mathrm{natValue}(1 + n) &=& (\synFalse, \mathrm{natValue}(n))
-  \end{array}
-\end{displaymath}
+$\mathrm{natValue}(n) \in \ValSet$ by recursion:
+\begin{mathpar}
+  \mathrm{natValue}(0) = (\synTrue , *)
+
+  \mathrm{natValue}(1 + n) = (\synFalse, \mathrm{natValue}(n))
+\end{mathpar}
 This representation uses a tagged pair approach similar to our
@@
 \subsection{The Cons-Free System}
-\label{sec:cons-free-sound}
+\label{sec:consfree-sound}
 
@@
 whole system are formalised in the Agda modules \AgdaModule{ConsFree}
-and \AgdaModule{ConsFree.Iterator}. The soundness theorem above is a
-combination of the inequality definition in our indexed linear
-preorder and the QTT model sketched in \autoref{sec:qtt-models}.
+and \AgdaModule{ConsFree.Iterator}. The soundness theorem is a
+combination of this and the QTT model sketched in
+\autoref{sec:qtt-models}.
 
@@
 \item We define the realisability specification for diamonds
-  $\Diamond \in L(1)$ to be:
-  \begin{displaymath}
-    \Diamond = \{ (*, \alpha, *) \mid 0 \leq M(\alpha, \mathit{size}(1)) \}
-  \end{displaymath}
-  Hence a diamond represents at least one unit of size resource,
-  matching the intuitive explanation given in
+  $\Diamond \in L(1)$ to be
+  $\Diamond = \{ (*, \alpha, *) \mid 0 \leq M(\alpha,
+  \mathit{size}(1)) \}$.  Thus, a diamond represents at least one unit
+  of size resource, matching the intuitive explanation given in
   \autoref{sec:lfpl-intro}.
@@
 
-\subsection{Systems for Polytime}
-
-BLL, SAL, LAL, LFPL
-
-The ones from Fossacs 2022, and POPL23(?) Romain Perchoux
-
-Neil Jones
-
-\subsection{Other Complexity Classes}
-
-\begin{enumerate}
-\item Lower complexity classes, like LOGSPACE
-\item Higher complexity classes, like EXPTIME, ELEMENTARY
-\end{enumerate}
-
-\subsection{Explicit Resource Accounting}
-
-\begin{enumerate}
-\item RAML, Hofmann-Jost, Raja, etc.
-\item lambda-amor,
-\item Daniellson
-\item calf from CMU
-\end{enumerate}
-
-\subsection{Linear and Substructural Dependent Types}
-
-Linear LF
-
-Pradic et al
-
-Grtt
-
-Penn
-
-Abel
-
-Related work: only applications of Linear TT so far are erasure (Abel
-et al.) and quantum stuff (Selinger et al.), and (sort of) dependency
-analysis. This is another one.
-
-What we didn't do: polytime typechecking guarantees.
-
-\subsection{Implementation} Could use Idris, but it has the wrong
-semiring, and its datatype definition facility is too permissive.
-
-\subsection{Towards Applications}
-
-Cryptography?
-
-\subsection{Towards Synthetic Computational Complexity Theory?}
+\subsection{Related Work}
+
+\paragraph{Implicit Computational Complexity with Linear Types}
+Implicit Computational Complexity \cite{Lago11} is a vast field, so we
+only survey closely related works. We have already mentioned the
+Bounded Linear Logic \cite{bll92}, Soft Affine Logic \cite{Lafont04},
+Light Linear Logic \cite{lll98} and LFPL \cite{hofmann99lfpl} systems,
+which all use linear typing to implicitly capture polynomial
+time. \citet{Jones01} characterises polynomial time using first-order
+functional programs without constructors. Thus it shares a method with
+our \emph{Cons-free} system, but we use linear typing to permit
+controlled use of higher-order functions. Other approaches to
+polynomial time use stratification or information flow tracking to
+ensure that the outputs of iteration may not be used unrestrictedly to
+drive further iteration. For example, \cite{BellantoniC92} and
+\cite{HainryP23}. Below polynomial time, systems have be devised to
+capture LOGSPACE \cite{LagoS16}. Above polynomial time, systems such
+as Elementary Affine Logic (EAL) capture all Elementary-time functions
+\cite{CoppolaM01}.
+
+We have used \citet{dallago11realisability}'s technique to prove
+soundness of our extension of QTT. This technique has been
+successfully applied to many other linear typing based systems, such
+as BLL \cite{hofmann04bll-realisability,dallago10bll} and LLL
+\cite{LagoH10} and EAL. In contrast to most of those systems, we do
+not use restricted $\oc$-modalities and second order encodings to
+express datatypes. Our explicit datatype approaches enabled our
+combination of dependent types and polynomial time.
+
+\paragraph{Explicit Resource Accounting with Dependent Types} In
+contrast to the implicit systems, previous works have constructed
+systems that give explicit resource bounds via typing. Examples
+include \citet{HoffmannDW17}'s Resource Allocated ML (RAML) and
+\citet{RajaniG0021}, both of which are based on ideas of type-based
+amortised complexity analysis arising from \cite{hofmann99lfpl}'s
+ideas, via the work of \cite{HofmannJ03}. More details are to be found
+in the survey paper of \citet{HoffmannJ22}. A approach is to track
+costs at the value level instead of the types. \citet{Danielsson08}
+describes a system that uses a ``tick'' effect to count steps of
+computation, which can be reasoned about via dependent
+types. \citet{NiuSGH22} take this idea further by employing a
+modality-based phase separation to ensure that tick counting never
+interferes with the functional business of
+programs. \citet{McCarthyFNFF16} is another tick effect based system
+in Coq. All of these techniques rely on the programmer correctly
+annotating the program with tick effects to count the resource usage
+they are interested in, in contrast to the intrinsic approach we have
+taken here.
+
+\paragraph{Linear and Substructural Dependent Types} We chose QTT as
+the particular combination of linear and dependent types for our
+systems. Other systems include systems such as those by
+\cite{CervesatoP02,KrishnaswamiPB15,Vakar14} which all use a strict
+separation between linear and non-linear variables. This strict
+separation would mean that we could not as easily move programs from
+the linear fragment into the types, as in
+\autoref{sec:programming-polytime}. Systems that are more like QTT in
+that they do not have a strict separation of variables include those
+of \cite{MoonEO21,ChoudhuryEEW21,Abel18}. These systems differ from
+QTT in that they do not include a complete copy of unrestricted type
+theory as QTT does in its $\sigma = 0$ fragment, because they all
+track usage in types as well as terms, so it is not clear how to use
+them for unrestricted reasoning as we do with QTT. \citet{FuKS22}
+present a system that is closer to QTT but does not include a universe
+type.
+
+\subsection{Future Work}
 \label{sec:towards-synthetic-complexity-theory}
 
-Synthetic computability theory Bauer, Forster
+\paragraph{Implementation} We currently lack an implementation of our
+extension of QTT, which hampers further investigation of programming
+and proving with polytime along the lines of
+\autoref{sec:programming-polytime}. Idris 2 \cite{Brady21} is an
+implemenation of QTT, but cannot be used directly because its facility
+for defining datatypes is too liberal, not making a distinction
+between iterable and non-iterable datatypes. A further
+implementation-focused question is whether or not the term-level
+polytime guarantees can be turned to type-level guarantees to
+guarantee polytime typechecking.
+
+\paragraph{Other Complexity Classes, and Explicit Resource Tracking}
+It seems feasible to adapt the techniques presented here to other
+complexity classes such as LOGSPACE and ELEMENTARY, given the simply
+typed linear systems mentioned above. Also, our construction already
+includes soundness of a system with intrinsic but explicit resource
+tracking where $\Diamond$s are used to pay for every step of
+computation but never returned, via the natural number resource monoid
+defined in \autoref{sec:specific-monoids}. Investigation of such a
+system may yield a system that tracks costs intrinsically and
+precisely.
+
+\paragraph{Towards Synthetic Computational Complexity Theory?}
+
+The realisability type $\mathbf{R}(A)$ described in
+\autoref{sec:qtt-reflection} allows us to internalise the
+realisability of certain functions into the logical ($\sigma = 0$)
+fragment of the calculus. However, it is not possible to derive any
+logical consequences from this other than turning it back into a
+function. This limitation becomes acute when trying to prove results
+from standard Computational Complexity theory. Even though we can
+characterise the class NP, as we did in \autoref{sec:np-class}, we
+cannot prove the Cook-Levin theorem that 3-SAT is NP-complete. This is
+because the proofs relies on obtaining the \emph{source code} of the
+program solving an NP problem and then encoding that program in
+3-SAT. To do this in our setting, we would need to internalise the
+soundness property (\thmref{thm:lfpl-soundness}) as an axiom, stating
+that for a realisable polytime function there (merely) exists a
+realising expression $E$ that completes in polynomial time, and then
+writing polytime enodings into 3-SAT. We hope that the addition of
+such an axiom to our system would lead to an expressive machine-free
+\emph{Synthetic Computational Complexity Theory}, analogous to the
+Church-Turing axiom for Synthetic Computability Theory \cite{Bauer06}.
 

### 2023-07-11 paper/paper.tex
[ paper ] De-anonymised version

@@
   Martin Hofmann, that controls construction via a payment
-  method. Both of these are extend to full dependent types via
+  method. Both of these are extended to full dependent types via
   Quantitative Type Theory, allowing for arbitrary computation in

### 2023-09-14 paper/response.txt
[ paper ] POPL reviews and response

Thank you to all the reviewers for their engaging and detailed reviews. We are glad that all of the reviewers said they enjoyed reading our submission and came up with interesting questions. We have enjoyed writing a response to them, and think that incorporating the many suggestions will make a much improved paper.

## Proposed Changes

- We will address Reviewer A and Reviewer C's point that the exact definitional equality of QTT is unclear by including the equality rules in an appendix. We will also include a discussion of the metatheory of the systems presented in this work, as described in the response to Reviewer A below. (Reviews A and C).

- We will elaborate on the exact meaning of poly-time soundness earlier in the paper, to avoid confusion (Review C).

- We add more examples of programming in the LFPL-style, to highlight the exact way that the resource discipline shows itself (Reviews B and D).

- We will consider switching the presentation of resource monoid to the enriched category theory definition used in the formalisation (Review C).

- We will add some more background on Implicit Computational Complexity (Review D).

- We will fix the many minor typos and confusing bits of wording found by the reviewers.

## Review A

Thank you for the review. You bring up important points about the presentation of QTT and its metatheoretic properties, and the prospects for a theory of synthetic computational complexity. We have tried to itemize and summarise these points to respond to them.

1. *Lack of closed system of PTIME-QTT*

   As we also state below in our response to Review C, the omission of the definitional equality rules was for brevity. QTT has the usual $\beta\eta$-laws for $\Pi$-types and $\Sigma$-types, and $\beta$-laws for the inductive types. In the $\sigma=1$ fragment this means that the $\eta$-law for the `let` eliminator also includes commuting conversions, which might actually be infeasible for implementation. The $R$ and $R^{-1}$ terms are definitionally mutually inverse.

   We will include the full definitional equality rules in any revision of the submission.

   The definitional equality rules are all justified by the Set part of the model and the fact that the resource component is a preorder.

   You are right that the Agda development does not prove anything about equalities (this is the main component of the gap in the development identified on ine 917). The construction of the preorder is proof relevant "by default" in Agda. Our preferred way to proceed is to actually construct a Setoid QCwF in Agda, which simply ignores the realisability component for the purposes of equality. Definitional equality is then interpreted up to the setoid equality.

2. *Lack of metatheory*

   Again, this was for brevity. The weakening and substitution properties carry over from (Atkey, 2018). Type checking is undeciable for the presented system due to equality reflection. We conjecture that a system with intensional equality would have decidable conversion and typechecking, but we have not proved this.

   Consistency is a consequence of the consistency of the Set model, though as you say it would need an empty type to state it. It is possible to interpret the whole system in the usual Set CwF, ignoring the linearity. This would no longer be possible if axioms were added to the system to internalise the polytime property however, please see the next point.

   For canonicity, there are two questions, external and internal. The realisability semantics described in the paper demonstrates that every closed term of boolean type has a realiser that evaluates to a boolean, or, more usefully, that every closed term of function type $\mathrm{Nat} \multimap \mathrm{Nat}$ has a realiser that evaluates to a function that evaluates to natural number values when supplied with natural number values.

   Internal canonicity, that every closed *source* language term of ground type is $\beta\eta$-equal to a closed value, is not proven. It would probably be a corollary of decidability of conversion. We do not think there would be a conflict between the full $\beta\eta$ equational theory and the underlying CBV implementation. The realisability model ensures that they compute the same answers, no matter how they get there.

3. *Prospects for a synthetic complexity theory*

   We think that the relativisation that you propose would be accounted for by use of the realisability predicate we introduced in Section 3.5, which attaches realisability information to the values manipulated by the set-theoretic parts.

   The idea then would be that there would be a strict proposition type (like Agda's Prop or Coq's SProp) with (somehow) an axiom in that proposition type that states $(f : \mathbf{R}(\mathrm{Nat} \to \mathrm{Nat})) \to \exists c:\mathrm{Code}.~\exists p:\mathrm{Poly}.~\forall n:\mathrm{Nat}.~\exists k:\mathrm{Nat}.~\llbracket c \rrbracket n = (\mathrm{R}^{-1}(f)~n,k) \times (k \leq p(n))$, where $\mathrm{Code}$ is some type of representations of code and $\llbracket - \rrbracket$ is an interpretation function that returns the result and the number of computation steps. This axiom is not interpretable in a plain Set model, but it is justified by the realisability model. The existential must live in Prop so that the code representation cannot leak out and destroy extensionality of functions. (We hope this quick sketch is understandable!)

## Review B

Thank you for the review. You raise an important point about the relationship between dependent types and the kind of complexity control we are interested in here. We respond to this first, and then your smaller comments.

> Can you provide an example program with a (tight) cost bound that can be expressed in this system but not with just linear types?

The systems presented are not intended to provide tight cost bounds on programs as part of its typing judgement. They are only intended to ensure that all terms (in the $\sigma = 1$ fragment) are realisable by polynomial time implementations.

It was not our goal with this work to use dependent types to prove complexity bounds, or to express tight bounds on complexity in the types.

Our goal is to combine *implicit* control of complexity with dependent types. Dependent types are used to reason about the extensional behaviour of programs, while the linear types ensure the complexity bounds.

The purpose of the insertion sort example was to show that writing "normal" functions was actually feasible in the LFPL-style system.

Where dependent types come in indispensible is in expressing complexity classes that build on PTIME, such as NP and PP, as we do in Section 4.3. These complexity classes combine polytime-ness with specific correctness criteria, something that is not expressible as a type without the combination of linear and dependent types. The correctness of polynomial time reductions also relies on dependent types to prove that the reduction does indeed reduce the problem.

That said, we think that two further things are possible, building on this work:

  1. The realisability semantics does provide a concrete polynomial for each term, so it might be possible to make a type system that exposes this information.

  2. The natural number resource monoid (Definition 5.2, line 815) is not used in this submission, but it could be used to make a system where $\Diamond$s are explicitly counted in the types (so there'd be a type $\Diamond : \mathrm{Nat} \to \mathsf{U}$) and are used to pay for each step of computation. The user would then be able to prove tight cost bounds on the $\Diamond$ fuel provided to functions. We think this would be an interesting system to explore in a future paper, more along the lines of a fully dependently typed and explicitly resourced version of (Jan) Hoffmann et al's RAML system (POPL 2017).

On the topic of more examples, please see the response to Review D below.

From the detailed comments:

1. *The var rule in the affine $\lambda$-calculus.*

   Yes, the context could just be $\Gamma, x : A$.

2. *Variable clash on lines 208-213*

   The re-use of the variable name $s$ was intentional, because the one in the $\mathsf{succ}$ case represents the intermediate state being updated, while the $\lambda$-bound one is the initial state, so conceptually they are both "the state". We can see how this is confusing though. We will rename the $\lambda$-bound one to $s_0$.

3. > Why is the construction of booleans cost-free?

   Roughly put, there is no way to "save up" iterability from the input in a boolean because they are non-recursive. Constructors for iterable natural numbers are disallowed or controlled by $\Diamond$s because they would otherwise allow the input to be duplicated to exceed polynomial bounds. More formally, because it is justified by the realisability model.

4. *Sentence missing on line 599*

   The sentence ought to be: "This [the fact that the list elements are iterable naturals] is needed to account for the comparisons between elements."

## Review C

Thank you for the very detailed comments. We will fix the minor typos. Here are responses to the questions and comments in the *Detailed comments* section. We have paraphrased the original comments before each response.

1. *No equational theory given*

   This was just for brevity, but in hindsight it should have been explicitly commented on. We responded to this comment in more depth for Review A, above.

2. *What does it mean to represent all polytime computations?*

   We agree that it is a bit confusing. The soundness property is stated in terms of "this program can be realised by a polynomial time implementation", not "this piece of code will reduce in polynomial time in the size of the program" (which is another plausible reading, and one used elsewhere in implicit computational complexity).

   We will add text explaining this point. The best place seems to be around line 103 where complexity of cut-elimination is (perhaps misleadingly) mentioned, and then elaborated upon in Section 2.2

3. *Where do Nats come from?*

   In the Cons-free system, all iterable natural numbers come from the input. In the case of the `iterate` function, the unstated assumption is that the input is supplied in the form of a pair of a natural number and the initial state. The natural number is intended to be the "size" of the input, e.g. the length of the initial tape (and this could be enforced using dependent types).

4. *What is the status of the admissible rules?*

   The $0$-ing admissible rule is more like subst/wk. It shows up explicitly in the QCwF semantics as the faithful functor $U : \mathcal{L} \to \mathcal{C}$ that drops the resource information, and the corresponding operation on semantic terms. Its presence as an admissible operation is part of the "trick" of QTT that keeps the syntactic form of the two fragments the same.

   We think that a discussion of this point might be out of scope for this paper, but certainly a syntax-free description of QTT would be something worthwhile to explore. We note that the system of Fu, Kishida, and Selinger (2022) does have an explicit "forget the resources" operation in the system (called the "shape"), but the system is different in that this is also applied to the types as well as the terms, unlike in QTT.

5. > In §3.5 (“Reflection of Realisability”) ...

   The fragment-asymmetry is odd, and it would perhaps be better if it were decomposed into separate operations of unreflection and projection into the $0$ fragment, but that would not be compatible with the way that QTT is currently set up.

   The definitional equality rules for this type state that $R^{-1}$ and $R$ are mutually inverse, and they hold in both fragments.

6. *Where do proofs about PTIME(A,P) live?*

   Your interpretation is correct. Proofs of `PTIME(A,P)` need to be constructed in the $0$ fragment because the proof that a problem is solved by a particular function may not be polynomial time (or it might be inconvenient to express it in such a way that polytime-ness can be verified by typing). We will make this clearer in any future revision.

7. > In l. 710 it says “These explicit constructions are essential to construct a model of dependent types”

   On reflection, "essential" is too strong. Using impredicative encodings in the surface type theory is not compatible with many other things that one might want to do with a type theory. However, it might be possible to construct a model that uses $\lambda$-/Church-encodings of inductive datatypes in the implementation, but we think such a construction would be less clear that the explicit first order construction.

8. *Why not use the enriched category definition of resource monoid?*

   We thought that the monoid + difference operator version would be more digestable to readers without some enriched category theory knowledge, and it stays closer to Dal Lago and Hofmann's original presentation. We will certainly consider making the enriched category definition primary though.

9. *Why an extra slot in the closures for self-reference?*

   Closures *themselves* don't have an extra slot for self-reference, their code portions expect to be supplied with a self-reference when invoked. We found this presentation made the proofs of realisability for the iterators manageable.

   We're not sure what you are referring to by "the normal way"? Encoding recursion by a CBV Y-combinator?

10. *In what sense is this a special case of something?*

    Yes, we agree the wording is rough. There's two things going on:

	 - There's a mistake in (Atkey, 2018). The category for interpreting linear contexts shouldn't be the category of assemblies, rather it should be the total category of the realizability hyperdoctrine as it is in this submisson. Thus, the construction in this submission generalises the corrected version of (Atkey, 2018).

     - The second "special case" in the cited sentence was meant to refer to the fact that this construction only handles the $\mathbb{N}$-annotated system, not for an arbitrary semiring, but this got lost in an edit. We will make this more clear, or remove the reference to being a special case.

## Review D

Thank you for your thought provoking comments on implicit computational complexity. We have tried to address them below. We are excited that you identified cryptography as a potential application area, we are very interested in this for future work.

Responses to comments from the *Assessment* section follow. We have paraphrased the questions for brevity.

1. *Why hasn't Implicit Computational Complexity had a greater impact?*

   We don't regard ourselves as complexity theorists, so we're perhaps not the best people to answer this question. Certainly implicit computational complexity theory ("complexity via functional programming") has not had the same impact as descriptive complexity ("complexity via logic programming").

   We speculate several possible reasons that ICC hasn't taken off outside a niche in PL theory, and ways that our work may help:

     - Sometimes the languages are arguably quite obscure, being based on Linear Logic sequent calculi or proof nets with impredicative encodings of data. We have tried to address this by being a restriction of normal type theory with inductive data types and eliminators.
     - Each complexity class seems to require an overhaul of the programming language used to characterise it, and there doesn't seem to be a systematic way of doing so. We have tried to address this, at least for classes based on polynomial time, with the effects-based characterisations of NP and PP in Section 4.3 of the submission. It would be interesting to explore what the effect of other notions of computational effect would be.
     - We have not seen any system that integrates restricted-complexity systems into larger ones that encompass non-restricted reasoning, as we have done in this submission. Our hypothesis is that this will enable reasoning about polynomial time, perhaps even with mechanised assistance.

   Of course, there is no guarantee that this is enough to push ICC into the mainstream, but we think it is worth a try. We will add a discussion refining and elaborating the above points to any future revision of the submission.

2. *Insertion sort has a simple resource discipline; can you do anything more complex*

   One of the features/constraints of LFPL is that the output always has smaller size than the input, where the size refers to the *iterable* size. It is possible to write a function of type $\mathrm{Nat} \multimap \mathrm{List}(1) \otimes \mathrm{List}(1)$ that generates two lists whose length is the size of the input, but these lists cannot themselves be iterated over.

   It is also possible to generate duplicate iterable lists by supplying $\Diamond$s: so a function of type $\mathrm{IList}(\Diamond \otimes A \otimes B) \multimap \mathrm{IList}(A) \otimes \mathrm{IList}(B)$ is implementable, by building the first output list with the $\Diamond$s from the input list and the second with the additional $\Diamond$ carried with each element.

   Programming in LFPL is nicer than programming in the Cons-free system, but is still not exactly like programming in normal functional languages. We will put more material on the necessary programming style for LFPL in any revision of this submission.

Comments from the *Detailed comments* section, omitting the typo corrections (thank you!)

1. > The type systems in section 2 seem to be affine (Γ, 𝑥 : 𝐴, Γ′ ⊢ 𝑥 : 𝐴), whereas the type system in section 3 is linear (0Γ, 𝑥:^𝜎 𝑆, 0Γ′ ⊢ 𝑥:^𝜎 𝑆). What is the motivation for this shift?

   The QTT instantiation is actually affine as well, due to the ordered semiring used, which has $1 \sqsubseteq 0$ so the $0$-annotated elements of the context in the var rule can be weakened to be $1$-annotated. We will alter the explanation of this point in Section 3.1.1. to make this clearer.

2. > What is the reference for the original Cons-free type system?

   We are not aware of a reference that exactly presents the same Cons-free system as in the submission.

   The underlying semantic model used is the similar to the one proposed by Dal Lago and Hofmann for Lafont's Soft Affine Logic (SAL), but the surface language differs in that it has an explicit natural number type, not a combination of impredicative polymorphism and a restricted exponential modality.

   Jones, 2001, described a polynomial-time system that also omits Cons, but is first order. Jones then goes on to add *unrestricted* (i.e. no linearity restriction) higher order functions which results in an exponential-time system. The system presented here restricts functions to be non-duplicable as well, which keeps the system to polynomial time.

   The combination of Cons-free-ness with non-duplicable functions seems like a step that someone else might have taken, so we didn't want to outright claim it as a contribution, but we are not aware of an exact match in the literature. We will include the above points in any future revision though.

3. > What about reducing pairs? That substitutes into both x and y at the same time.

   Each of `x` and `y` will appear at most once in the affine system, so the $\beta$-rule will still not have to duplicate the terms being substituted in and the term gets smaller by removing the $\beta$-redex.

4. > Why is it okay to duplicate natural numbers using the dupNat operator? Is this because you disallowed the constructors for Nat?

   The full answer to this lies in the semantic model discussed in Section 6. Intuitively, it is because while constructors allow arbitrary numbers to be generated from the input, duplication only allows the input to be iterated over again, possibly nested.

5. > line 261: "Rather the iterator I_k performs n choose k iterations". It actually performs n choose k - 1 operations.

   Good point. We will fix this.

6. > "Is the need for the n choose k iterations due to the absence of the duplicator operator?"

   Yes. The number of diamonds available for sub-iterations is equal to the *rest* of the natural number, so the recursion pattern ends up looking like a triangle, hence the use of binomial coefficients. We will elaborate on this connection further in any future revision.

7. > section 3: In the judgement for well-formed types, is it necessary to have the context of the form 0\Gamma? Couldn't we define well-formedness in \Gamma by just ignoring all annotations?"

   We think the system could be reformulated in this way, but the current presentation follows the original presentation of QTT.

8. > line 366: In the typing hypothesis for C, is the \pi annotation in the context necessary or can we just set it to be 0?

   The $\pi$ annotation is necessary because it is part of the type. Types with different annotations are different in QTT.

9. > line 456: Why is it necessary to add the dupNat construct to the 0 fragment, given that there we can just simulate it with the pair type?

   We need to add it to the $\sigma=0$ fragment so that the $0$-ing rule remains admissible directly. The definitional equality on line 463 then transforms it to a pair.

10. > How is the diamond different from the unit type in the 0 fragment?

    They are isomorphic in the $\sigma=0$ fragment, but they are not equal-as-types because they differ in their resource components.

11. > Does R^{-1}(R(M)) reduce definitionally to M?

    Yes. We will add in the equality rules for $\mathbf{R}$ in any future revision.

12. > 599: Sentence is incomplete.

    Please see the response to review B above.

### 2023-10-23 paper/paper.tex
[ paper ] Grant ack
[ nix ] A Makefile
[ paper ] Attempt to explain why programs are open

@@
-\documentclass[acmsmall,review,screen]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 \settopmatter{printfolios=true,printccs=true,printacmref=true}
@@
 accomplished within a certain complexity bound. Moreover, the system
-is usually proven to be complete for the relevant complexity class, by
+is usually proven to be complete for the relevant complexity class by
 constructing a simulation of some known representation. Such systems
@@
 
-Viewing logical systems though the lens of the Curry-Howard
-correspondence, the idea arises that one could define functional
-programming languages that characterise complexity classes such as
-polytime. SAL has been transformed into a programming language by
-\citet{BaillotM04}, and likewise for LLL by
-\citet{BaillotGM10}. \citet{hofmann99lfpl} proposed a new programming
-language, Linear Functional Programming Language (LFPL), that uses a
-novel ``payment'' system to track iteration. In this section, we
-review the use of linear types to capture polytime by presenting two
-systems, one based on ideas from SAL and the second more explicitly
-based on LFPL. With a view to extending to dependent types in
-\autoref{sec:qtt}, we take an approach slightly different to much of
-the polytime linear logic literature. We emphasise the use of explicit
-datatypes and eliminators, rather than using impredicative encodings
-via universal types. We are closer to Hofmann's original LFPL (though
-not a later presentation of it by \citet{dallago11realisability}) than
-BLL, SAL or LLL.
+Viewing logical systems though the Curry-Howard correspondence, the
+idea arises that one could define functional programming languages
+that characterise complexity classes such as polytime. SAL has been
+transformed into a programming language by \citet{BaillotM04}, and
+likewise for LLL by \citet{BaillotGM10}. \citet{hofmann99lfpl}
+proposed a new programming language, Linear Functional Programming
+Language (LFPL), that uses a novel ``payment'' system to track
+iteration.
+
+There are at least two ways that a functional programming language can
+be seen as representing polynomial time, differing in how the size of
+the input is measured. One approach is to define the input as a closed
+expression, including what we might think of as its input, and
+computation of the result is polynomial time in the combined size. A
+second approach is that the input is ``externally'' provided, so we
+consider open terms with a free variable representing the input. So a
+judgement $x : \tyNat \vdash M : A$ declares a program that computes
+results of type $A$ in time polynomial in the size of the natural
+number $x$. We take this latter approach in this paper.
+
+With a view to extending to dependent types in \autoref{sec:qtt}, we
+take an approach slightly different to much of the polytime linear
+logic literature. We use explicit datatypes and eliminators, rather
+than using impredicative encodings via universal types. We are closer
+to Hofmann's original LFPL (though not a later presentation of it by
+\citet{dallago11realisability}) than BLL, SAL or LLL.
+
+In this section, we review the use of linear types to capture polytime
+by presenting two systems, one based on ideas from SAL and the second
+more explicitly based on LFPL.
+
+\subsection{Affine Linear $\lambda$-Calculus}
 
 For this section, the affine linear $\lambda$-calculus we will use
-will have linear functions and $\otimes$-products:
+will have linear functions and $\otimes$-products:\bob{CHECK THAT THIS CHANGE TO THE VAR RULE IS OK, MENTION THAT EXCHANGE IS IMPLICIT}
 \begin{mathpar}
@@
   { }
-  {\Gamma, x : A, \Gamma' \vdash x : A}
+  {\Gamma, x : A \vdash x : A}
 
@@
 to the size of the term. This is because every $\beta$-redex
-substitutes into at most one variable, reducing the size of the term
-by one each time.
+substitutes each term into at most one variable, reducing the size of
+the term by one each time.
 
@@
 repeated multiplication). We therefore refer to this system as the
-\emph{Cons-Free} system.  The second prohibited feature is the ability
-to duplicate values of function type, even though we have allowed
-duplication of iterable naturals. If we were to allow this, then we
-would be able to sneak in a form of constructors for natural numbers
-by encoding them as eliminators that duplicate a function for every
-$\conSucc$ step.
+\emph{Cons-Free} system.  Because we cannot construct values of type
+$\tyNat$ within the system, complete programs in this system are open
+terms as we explained at the start of this section.
+
+The second prohibited feature is the ability to duplicate values of
+function type, even though we have allowed duplication of iterable
+naturals. If we were to allow this, then we would be able to sneak in
+a form of constructors for natural numbers by encoding them as
+eliminators that duplicate a function for every $\conSucc$ step.
 
@@
 
+\bob{Mention equality rules, and put them in an appendix; also
+  Conversion rule is missing!}
+
 \subsubsection{$\Pi$- and $\Sigma$-types}
@@
 
+\bob{$\beta\eta$-laws for $\Pi$-types}
+
 $\Sigma$-types are a little more involved, and demonstrate the
@@
 pattern-matching \cite{atkey18qtt}.
+
+\bob{$\beta\eta$-laws for $\Sigma$-types, with wrinkle in the $\sigma = 1$ fragment}
 
@@
 \subsubsection{Booleans} The boolean type for QTT was described in
-\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
-offer no possibility for iteration, but it is useful to see how the
-QTT rules extend the simply typed rules from
+\cite{atkey18qtt}. Booleans offer no possibility for iteration, but it
+is useful to see how the QTT rules extend the simply typed rules from
 \autoref{sec:linear-case-analysis} before moving to more complex
@@
 \end{mathpar}
+\bob{Mention the $\beta$ law}
 The introduction rules for booleans both use a $0$-d context,
@@
 \end{mathpar}
+\bob{$\beta$ laws!}
 In the $\sigma = 0$ fragment, however, we are free to iterate on lists
@@
 ensure that the resources captured by the number are not duplicated.
+
+\bob{$\beta$ laws in the $\sigma = 0$ fragment}
 
@@
 
+\bob{$\beta$ laws in both fragments}
+
 Just as for the Cons-free system iterator above, in the $\sigma = 0$
@@
 require to maintain the admissibility of the $0$-ing rule.
+
+\bob{Equality rules (see response)}
 
@@
 
+\bob{Check PP vs BPP. Also cite Dal Lago PP-time paper}
+
 By changing the computation effects supplied to a program, we can
@@
 \newcommand{\cstIf}{1}
-
-% \newcommand{\cstaccess}{c_{\mathit{access}}}
-% \newcommand{\cstmkclo}{c_{\mathit{mkclo}}}
-% \newcommand{\cstapp}{c_{\mathit{app}}}
-% \newcommand{\cstmkpair}{c_{\mathit{mkpair}}}
-% \newcommand{\cstmkunit}{c_{\mathit{mkunit}}}
-% \newcommand{\cstTrue}{c_{\mathit{mktrue}}}
-% \newcommand{\cstFalse}{c_{\mathit{mkfalse}}}
-% \newcommand{\cstLetpair}{c_{\mathit{letpair}}}
-% \newcommand{\cstSeq}{c_{\mathit{seq}}}
-% \newcommand{\cstIf}{c_{\mathit{if}}}
 
@@
 sketch a special case here to concentrate on the specifics of
-complexity bounded realisability.
+complexity bounded realisability. \bob{state clearly that Atkey 2018
+  has a mistake, and this construction here is more general}
 
@@
 $\oc_n$ has no effect on realising values. It only serves to alter the
-resource potentials.
+resource potentials.\bob{fix this to match the Agda implementation: special case $0$}
 \begin{displaymath}
@@
 \begin{acks}
-  FIXME
+  This work was funded by the \grantsponsor{EPSRC}{Engineering and
+    Physical Sciences Research
+    Council}{https://www.ukri.org/about-us/epsrc/}. Grant number
+  \grantnum{EPSRC}{EP/T026960/1}, \emph{AISEC: AI Secure and
+    Explainable by Construction}.
+  % FIXME: add Fred, when his comments have been included.
 \end{acks}

### 2023-10-24 paper/paper.tex
[ paper ] typo
[ paper ] equational theory throughout

@@
 steps. Then, using the iterator above we can iterate $f$ over a
-$\tyNat$ representing the size of the input:
+$\tyNat$ representing the size of the input:\bob{CHECK THIS}
 \begin{displaymath}
@@
 variables with annotations from a semiring. It sits in the general
-area of systems that use semiring annotatons to measure resource usage
+area of systems that use semiring annotations to measure resource usage
 \cite{BrunelGMZ14,GhicaS14,OrchardLE19}. The key feature of QTT, an
@@
 
-\bob{Mention equality rules, and put them in an appendix; also
-  Conversion rule is missing!}
+As usual, definitional equality of types impacts typing of terms via
+the conversion rule:
+\begin{displaymath}
+  \inferrule*
+  {\Gamma \vdash M : S \\ 0\Gamma \vdash S \equiv T~\istype}
+  {\Gamma \vdash M : T}
+\end{displaymath}
+Like type formation, definitional equality of types always takes place
+in $0$d contexts. We will describe the definitional equality rules for
+terms of each type as we introduce them. In QTT, it is possible for
+the definitional equality of terms to differ between the $\sigma = 0$
+and $\sigma = 1$ fragments, as we will see below.
 
@@
 we mean that the argument is to be used linearly:
-$(x \stackrel1: S) \to T$, where $x$ does not appear in $T$.
-
-\bob{$\beta\eta$-laws for $\Pi$-types}
+$(x \stackrel1: S) \to T$, where $x$ does not appear in
+$T$. $\Pi$-types support the usual $\beta\eta$ definitional
+equalities.
 
@@
 QTT also supports a unit type $I$ with constructor $*$ and
-pattern-matching \cite{atkey18qtt}.
-
-\bob{$\beta\eta$-laws for $\Sigma$-types, with wrinkle in the $\sigma = 1$ fragment}
+pattern-matching \cite{atkey18qtt}. $\Sigma$- and $I$ types support
+the usual $\beta\eta$ definitional equalities in the $\sigma = 0$
+fragment (e.g., $\mathrm{fst}(M,N) \equiv M$), but only $\beta$
+equalities (i.e.,
+$\mathrm{let}~(x,y) = (M_1,M_2)~\mathrm{in}~N \equiv N[M_1/x,M_2/y]$)
+in the $\sigma = 1$ fragment. It would also be sound to support
+commuting conversions \cite{Barber1996} for $\mathrm{let}$ in the
+$\sigma = 1$ fragment, but this would likely bring complications for
+implementation.
 
@@
 \subsubsection{Universe} QTT has universe types $\mathsf{U}$, as in
-standard type theory \cite{atkey18qtt}.
+standard type theory \cite{atkey18qtt}. For our examples below, we do
+not explicitly mark the use of terms of type $\mathsf{U}$ as types --
+i.e., we use a Russell-style presentation. Universes are where the
+definitional equality on terms affects the definitional equality on
+types.
 
@@
 \end{mathpar}
-\bob{Mention the $\beta$ law}
 The introduction rules for booleans both use a $0$-d context,
@@
 $\Gamma_2$. Since only one of the branches will be used, sharing
-resources between the branches is expected.
+resources between the branches is expected. Booleans and their
+eliminator obey the usual $\beta$ laws for definitional equality:
+$\If_{x.P}\: \cTrue \: \Then \: N_t \: \Else \: N_f \equiv N_t$, and
+similarly for $\cFalse$.
 
@@
 matching without recursion in the full calculus. Here is the rule for
-dependently typed case analysis on lists:
+dependently typed case analysis on lists, which also obeys the usual
+$\beta$-equalities for case analysis, analogous to the ones for
+Booleans:
 \begin{mathpar}
@@
 \end{mathpar}
-\bob{$\beta$ laws!}
 In the $\sigma = 0$ fragment, however, we are free to iterate on lists
@@
 \end{mathpar}
+This eliminator also obeys the usual $\beta$-equality laws for a list
+eliminator, using the resource freedom of the $\sigma = 0$ fragment to
+duplicate the $N_2$ term in the $\cCons$ case.
 
@@
 handed a list of an unknown length, we can only explore it to a fixed
-depth determined statically. To write programs that do work
-proportional to the size of their input, we need some form of iterable
-datatype. In both our Cons-free and LFPL-style QTT systems, we use a
-natural number datatype.
+depth, determined statically by the program. To write programs that do
+work proportional to the size of their input, we need some form of
+iterable datatype. In both our Cons-free and LFPL-style QTT systems,
+we use a natural number datatype.
 
@@
 \end{mathpar}
-Anyone who has reasoned about the metatheory of or implemented a type
-checker for dependent types will view this rule with unease as it
+Anyone who has reasoned about the metatheory of, or implemented a type
+checker for, dependent types will view this rule with unease as it
 appears to grant the ability to construct non-canonical values of pair
@@
 The eliminator for these natural numbers takes the following
-form. Disregarding the usage annotations, this has the same type
-structure as the normal dependently typed recursor for natural
-numbers:
+form. Disregarding the usage annotations, it has the same type
+structure as the normal dependently typed recursor for naturals:
 \begin{displaymath}
@@
 
-\bob{$\beta$ laws in the $\sigma = 0$ fragment}
+This eliminator cannot have a $\beta$-equality in the $\sigma = 1$
+fragment because there is no way to construct any natural numbers to
+iterate on in this fragment. In the $\sigma = 0$ fragment, this
+eliminator obeys the usual $\beta$-equality laws for a natural number
+recursor. This allows us to use it to compute and reason about
+operations on naturals in this fragment.
 
@@
 
-\bob{$\beta$ laws in both fragments}
+Unlike the natural number iterator in the Cons-free system, this
+iterator has $\beta$-equalities in both the $\sigma = 0$ and
+$\sigma = 1$ fragments. In the $\conSucc$ case, for example, we have:
+\begin{displaymath}
+  \begin{array}{cl}
+    & \tmRec\,(\conSucc(M_d,M_n)\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
+    \equiv & N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p]
+  \end{array}
+\end{displaymath}
+Note that the fact that the variable $n$ in the $N_s$ term is
+annotated $0$, which allows us to duplicate it even when we are in the
+$\sigma = 1$ fragment.
 
@@
 
-\bob{Equality rules (see response)}
+The equality rules for $\Rtype$ state that the two operations are
+mutually inverse: $\rIntro(\rElim(M)) \equiv M$, in both fragments,
+and $\rElim(\rIntro(M)) \equiv M$ in the $\sigma = 0$ fragment.
 

### 2023-10-24 paper/paper.tex
[ paper ] PP/BPP and semantic mistakes

@@
 For this section, the affine linear $\lambda$-calculus we will use
-will have linear functions and $\otimes$-products:\bob{CHECK THAT THIS CHANGE TO THE VAR RULE IS OK, MENTION THAT EXCHANGE IS IMPLICIT}
+will have linear functions and $\otimes$-products. Contexts are
+treated up to permutation, so uses of exchange are implicit.
 \begin{mathpar}
@@
 
-\subsubsection{Probabilistic Polynomial Time}
-
-\bob{Check PP vs BPP. Also cite Dal Lago PP-time paper}
+\subsubsection{Bounded-error Probabilistic Polynomial Time}
 
 By changing the computation effects supplied to a program, we can
-change the complexity class. To capture Probabilistic Polynomial time,
-we use a (non-iterable) data structure representing trees of
-probabilistic choices, where $\mathbb{Q}[0,1]$ is some type of
-(non-iterable) rationals in the closed interval $[0,1]$:
+change the complexity class. To capture the class BPP of Bounded-error
+Probabilistic Polynomial time \cite{aurora}, we use a (non-iterable)
+data structure representing trees of probabilistic choices, where
+$\mathbb{Q}[0,1]$ is some type of (non-iterable) rationals in the
+closed interval $[0,1]$:
 \begin{displaymath}
@@
 problems, where the decider is allowed to make probabilistic choices
-as long as it is correct with probability at least $\frac{1}{3}$:
-\begin{displaymath}
-  \mathrm{PP}(A,P) =
+as long as it is correct with probability at least $\frac{2}{3}$:
+\begin{displaymath}
+  \mathrm{BPP}(A,P) =
   \begin{array}[t]{l}
     (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
-    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{1}{3}) \Leftrightarrow P\,a
+    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{2}{3}) \Leftrightarrow P\,a
   \end{array}
 \end{displaymath}
-Again, problems in PP are easily seen to be closed under polytime
+Again, problems in BPP are easily seen to be closed under polytime
 reductions.
+
+Probabilistic Polynomial time has previously been considered in the
+setting of implicit computational complexity by
+\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
+cases, they must build probabilistic choice into the language, and
+have difficulty in directly capturing the class BPP due to its
+semantic nature \cite{aurora}, where the correctness of implementation
+is probabilistic. With a dependently-typed host language, adding
+probabilistic choice as an effect, and capturing the semantic
+constraint of BPP is straightforward.
 
@@
 \citet{atkey18qtt} described a general class of QTT models termed
-\emph{Quantitative Categories with Families} (QCwFs). This class of
-models fits well with the syntax of QTT, but to actually build models
-it is convenient to construct them from certain indexed linear
-preorders. The BCI-realisability model construction given by
-\citet{atkey18qtt} is a special case of this construction, so we only
-sketch a special case here to concentrate on the specifics of
-complexity bounded realisability. \bob{state clearly that Atkey 2018
-  has a mistake, and this construction here is more general}
+\emph{Quantitative Categories with Families}
+(QCwFs). \citet{atkey18qtt} constructs QCwFs from certain Linear
+Combinatory Algebras (LCA), where terms in the $\sigma = 1$ fragment
+are realised by elements of the LCA. However, there is a mistake in
+that paper where the interpretation of contexts is stated to be the
+category of \emph{assemblies} over the LCA, where it ought to be the
+category of sets paired with realisability relations, with no
+guarantee that all elements be realisable.
+
+Here, we fix the mistake of \citet{atkey18qtt}, and provide a more
+general construction of QCwFs in terms of indexed linear preorders. We
+construct indexed linear preorders specific to our polytime setting
+below. They could also be constructed from LCAs.
 
 \begin{definition}\label{def:indexed-linear-preorder}
-  A \emph{$\mathbb{N}$-linear preorder} is a preordered set
+  A \emph{$\mathbb{N}$-linear preorder}\footnote{We specialise to the semiring $\mathbb{N}$ here, but the same definition works for any suitable semiring $\mathcal{R}$.} is a preordered set
   $(L, \leq)$ with
@@
 $\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
-action of $(\mathbb{N}, \leq)$ on $M$ defined above. The modality
-$\oc_n$ has no effect on realising values. It only serves to alter the
-resource potentials.\bob{fix this to match the Agda implementation: special case $0$}
-\begin{displaymath}
-  \oc_n\,X = \{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
+action of $(\mathbb{N}, \leq)$ on $M$ defined above. When $n > 0$, the
+modality $\oc_n$ has no effect on realising values. It only serves to
+alter the resource potentials. In the $n = 0$, case the realising
+value must be $*$, in order to satisfy the $\oc_0 X \cong I$ condition
+in \defref{def:indexed-linear-preorder} 3(a):
+\begin{displaymath}
+  \begin{array}{lcl}
+    \oc_0\,X&=&\{(a, \alpha, *) \mid a \in A, \alpha \in M \} \\
+    \oc_n\,X&=&\{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
+  \end{array}
 \end{displaymath}

### 2023-10-24 paper/paper.tex
[ paper ] typos and a reference

@@
 in the list are themselves iterable natural numbers. This is needed to
-account for the
+account for the comparisons between elements.
 
@@
 this means that the two branches of this function can share resources
-(see the encoding of the additive product types in
-\cite{atkey18qtt}). Thus, each branch of this tree is polynomial time,
-not the whole tree itself.
+(see the encoding of the additive product types by
+\citet{atkey18qtt}). Thus, each branch of this tree is polynomial
+time, not the whole tree itself.
 
@@
 boolean-valued polynomial time function that has a path to returning
-$\cTrue$ exactly when the input is in predicate. Moreover, it is a
+$\cTrue$ exactly when the input satisfies the predicate. Moreover, it is a
 quick matter of programming to see that problems in NP are closed
@@
 change the complexity class. To capture the class BPP of Bounded-error
-Probabilistic Polynomial time \cite{aurora}, we use a (non-iterable)
-data structure representing trees of probabilistic choices, where
-$\mathbb{Q}[0,1]$ is some type of (non-iterable) rationals in the
-closed interval $[0,1]$:
+Probabilistic Polynomial time \cite{AroraBarak09}, we use a
+(non-iterable) data structure representing trees of probabilistic
+choices, where $\mathbb{Q}[0,1]$ is some type of (non-iterable)
+rationals in the closed interval $[0,1]$:
 \begin{displaymath}
@@
     \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
-    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy\ to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
+    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy \to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
   \end{array}
@@
 have difficulty in directly capturing the class BPP due to its
-semantic nature \cite{aurora}, where the correctness of implementation
-is probabilistic. With a dependently-typed host language, adding
-probabilistic choice as an effect, and capturing the semantic
-constraint of BPP is straightforward.
+semantic nature \cite{AroraBarak09}, where the correctness of
+implementation is probabilistic. With a dependently-typed host
+language, adding probabilistic choice as an effect, and capturing the
+semantic constraint of BPP is straightforward.
 

### 2023-10-24 paper/paper.tex
[ paper ] various small fixes

@@
   polytime computation with full dependent types, using Quantitative
-  Type Theory. The system is presented in \autoref{sec:qtt}. The
+  Type Theory. The systems are presented in \autoref{sec:qtt}. The
   linear typing discipline required for enforcing polytime is provided
@@
 specification purposes) and in terms (for computational purposes),
-syntactically yield multiple uses of the same variable.
+syntactically yielding multiple uses of the same variable.
 
@@
 regard it as usage $m$. This makes the system more like affine linear
-logic, since $m \leq 0$ for all $m$, matching the system in
-\autoref{sec:linear-intro}. We do \emph{not} have an unrestricted
-usage $\omega$, since this would allow the possibility of unrestricted
-duplication, and hence violate our polytime soundness properties.
+logic, since $m \leq 0$\footnote{Reverse ordering!} for all $m$,
+matching the system in \autoref{sec:linear-intro}. We do \emph{not}
+have an unrestricted usage $\omega$, since this would allow the
+possibility of unrestricted duplication, and hence violate our
+polytime soundness properties.
 
@@
   \inferrule*
-  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
-  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T}
+  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
+  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T~\istype}
 
@@
 \end{mathpar}
-As with $\Pi$-typed, the first component of a $\Sigma$ type is
+As with $\Pi$-types, the first component of a $\Sigma$-type is
 annotated with a usage for how many times it can be used, and this is
@@
 accounting its costs as described in the following sections. The
-$\sigma = 0$ recursor for lists has the following typing rule, which
-is the standard dependent eliminator for lists, with everything
-annotated as $0$ usage.
+$\sigma = 0$ fragment recursor for lists has the following typing
+rule, which is the standard dependent eliminator for lists, with
+everything annotated as $0$ usage.
 \begin{mathpar}
@@
   \inferrule*
-  {\Gamma \vdash}
+  {\Gamma~\isctxt}
   {\Gamma \vdash \conZero \stackrel0: \tyNat}
@@
   \inferrule*
-  {\Gamma \vdash}
+  {\Gamma~\isctxt}
   {0\Gamma \vdash \Diamond~\istype}
@@
   \inferrule*
-  {\Gamma \vdash}
+  {\Gamma~\isctxt}
   {0\Gamma \vdash * \stackrel0: \Diamond}
@@
 fragment, indicating that, in this fragment, it acts the same as a
-unit type. This allows us to use freely use diamonds in types, and to
+unit type. This allows us to freely use diamonds in types, and to
 not have to care about the identity of particular diamonds, since by
@@
 iterable inductive types, although in the \emph{Cons-free} system this
-is not particularly useful due to the prohibition of on
+is not particularly useful due to the prohibition of
 construction. However, sticking with just the natural numbers, we can
@@
 function, as we have no way of stating that the successor case must be
-arbitrarily duplicable. Lifting this restrict by means of some
+arbitrarily duplicable. Lifting this restriction by means of some
 modality is future work. The typing rule for the derived list iterator
@@
 Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
-decision problem $(A,P)$ is deciable in polytime. We make two notes
-about this definition: (i) this definition is intrinsic, in the sense
-that, whichever of the polytime systems is chosen, proving that a
-decision problem is solvable in polytime is a matter of programming,
+decision problem $(A,P)$ is deciable in polytime. We make three notes
+about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
+carried out in the $\sigma = 0$ fragment, where we have the full power
+of Type Theory to aid us; (ii) this definition is intrinsic, in the
+sense that, whichever of the polytime systems is chosen, proving that
+a decision problem is solvable in polytime is a matter of programming,
 without having to reason directly about machine models and step
-counting; and (ii) we have defined problems to have arbitrary types
+counting; and (iii) we have defined problems to have arbitrary types
 $A$ as domains, rather than bitstrings, and so the notion of size
@@
 \subsection{Polytime-based Complexity Classes}
+\label{sec:classes}
 
@@
     (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
-    \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
+    \quad \left((a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a\right)
   \end{array}
@@
 \paragraph{Agda Formalisation} The machine model is defined in the
-Agda module \AgdaModule{MachineModel}. We use an intrinsicaly
+Agda module \AgdaModule{MachineModel}. We use an intrinsically
 well-scoped syntax, which ensures that all variable look up operations
@@
   A \emph{$\mathbb{N}$-linear preorder}\footnote{We specialise to the semiring $\mathbb{N}$ here, but the same definition works for any suitable semiring $\mathcal{R}$.} is a preordered set
-  $(L, \leq)$ with
+  $(L, \leq)$:
   \begin{enumerate}
@@
   as $(A, X \otimes Y)$.
-\item For each set $A$, define the collection of semantics types
+\item For each set $A$, define the collection of semantic types
   $\Ty(A)$ as the collection of $B : A \to \Set$ and
@@
   terms $\Tm(A,(B,X))$ are functions $\Pi_{a \in A}.\,B(a)$. For each
-  context interpretation $(A, X)$ in $\cat{L}$ and type intepretation
+  context interpretation $(A, X)$ in $\cat{L}$ and type interpretation
   $(B,Y) \in \Ty(A)$, the $\sigma = 1$ fragment terms
@@
 \item Realisability reflection for a type $(B,X) \in \Ty(A)$ is
-  intepreted as the type
+  interpreted as the type
   $(\lambda a. \{ b \in B(a) \mid I \leq (\lambda a. (a,b))^*X \},
@@
 Thus, the boolean $\mathit{tt}$ is represented by the value $\synTrue$
-and $\mathit{ff}$ is represented by $\synFalse$. In both case, we
+and $\mathit{ff}$ is represented by $\synFalse$. In both cases, we
 allow arbitrary potential $\alpha$ to be attached.
@@
 set $B$. If we have $A : \Set$ and $B : A \to \Set$ and
-$X \in L(\Sigma a:A.\, B a)$, then we the resourced lists predicate
+$X \in L(\Sigma a:A.\, B a)$, then the resourced lists predicate
 $\mathrm{RList}(X) \in L(\Sigma a : A.\, \mathrm{List}(B a))$ must
@@
 The first property states that $\mathit{raise}$ is suitable as
-potential for whole programs2, meaning that it does not make any
+potential for whole programs, meaning that it does not make any
 requirements on the existence of iterable data. Note that we do
@@
 \autoref{sec:programming-polytime}. Idris 2 \cite{Brady21} is an
-implemenation of QTT, but cannot be used directly because its facility
+implementation of QTT, but cannot be used directly because its facility
 for defining datatypes is too liberal, not making a distinction
@@
 \paragraph{Other Complexity Classes, and Explicit Resource Tracking}
-It seems feasible to adapt the techniques presented here to other
-complexity classes such as LOGSPACE and ELEMENTARY, given the simply
-typed linear systems mentioned above. Also, our construction already
-includes soundness of a system with intrinsic but explicit resource
-tracking where $\Diamond$s are used to pay for every step of
-computation but never returned, via the natural number resource monoid
-defined in \autoref{sec:specific-monoids}. Investigation of such a
-system may yield a system that tracks costs intrinsically and
-precisely.
+We have been able to characterise the classes NP and BPP in terms of
+our underlying characterisation of P (\autoref{sec:classes}). It seems
+straighforward to extend this to related classes like coNP, RP,
+etc. It also seems feasible to adapt the techniques presented here to
+other complexity classes such as LOGSPACE and ELEMENTARY, given the
+simply typed linear systems mentioned above. Complexity classes based
+on circuits may be more challenging, but we do now have a way to
+characterise circuits that are generatable in polynomial
+time.
+
+Our construction already includes soundness of a system with intrinsic
+but explicit resource tracking where $\Diamond$s are used to pay for
+every step of computation but never returned, via the natural number
+resource monoid defined in
+\autoref{sec:specific-monoids}. Investigation of such a system may
+yield a system that tracks intrinsic costs precisely.
 
@@
 \begin{acks}
-  This work was funded by the \grantsponsor{EPSRC}{Engineering and
-    Physical Sciences Research
+  Thanks to Anton Lorenzen and Fredrik Nordvall Forsberg for their
+  comments on an earlier revision, and to the anonymous POPL reviewers
+  for their detailed and interesting reviews. This work was funded by
+  the \grantsponsor{EPSRC}{Engineering and Physical Sciences Research
     Council}{https://www.ukri.org/about-us/epsrc/}. Grant number

### 2023-10-24 paper/paper.tex
[ paper ] comment on proving 3-SAT is in NP

@@
 from standard Computational Complexity theory. Even though we can
-characterise the class NP, as we did in \autoref{sec:np-class}, we
-cannot prove the Cook-Levin theorem that 3-SAT is NP-complete. This is
+characterise the class NP, as we did in \autoref{sec:np-class}, and it
+is a ``matter of programming'' to show that 3-SAT is in NP, we cannot
+prove the Cook-Levin theorem that 3-SAT is NP-complete. This is
 because the proofs relies on obtaining the \emph{source code} of the

### 2023-10-24 paper/paper.tex
[ paper ] tidy up

@@
 
-Type theory is often claimed to be ideal for Computer Science,
+Type Theory is often claimed to be ideal for Computer Science,
 combining programming and proof in one unifying system, so a happy
 programmer can verify while they program, and program while they
-verify. From a broader Computer Science view, however, type theory
+verify. From a broader Computer Science view, however, Type Theory
 lacks the ability to talk about the very thing that makes Computer
@@
 cost of programs' execution. One can encode costs by embedding another
-programming language in type theory, for example \cite{GueneauCP18},
+programming language in Type Theory, for example \cite{GueneauCP18},
 or one can synthesise costs by treating resource counting as a
@@
 neither of these capture the intrinsic costs of the programs we write
-in type theory. These techniques deliver only \emph{conspicuous
+in Type Theory. These techniques deliver only \emph{conspicuous
   consumption}, not speaking of the real resources consumed.
 
-In this paper, we propose a method for extending dependent type theory
+In this paper, we propose a method for extending dependent Type Theory
 with a means for constraining the intrinsic computational complexity
@@
 which we hope will lead both to practical systems for reasoning about
-programs' resource usage, and to theoretical use as a form of
-\emph{synthetic computational complexity theory}. We discuss these
-possibilities further in \autoref{sec:conclusion}.
-
-% The goal of this paper is to unify implicit computational complexity
-% with full dependent types. ICC enables us to ensure that
-% \emph{programs} are within a certain complexity class, while dependent
-% types allow us to state and prove that these programs solve the
-% problems that we intend them to, and to construct reductions between
-% problems. We can marry complexity-controlled implementations with
-% problems, paving the way for a \emph{synthetic computational
-%   complexity theory}.
-
-% \item Section 5: Extensions
-%   \begin{enumerate}
-%   \item Explicit Complexity via Diamond Counting
-%   \item Poly-time univalence
-%   \item Towards P- and NP-completeness proofs. Issue here is the
-%     extensionality. Towards a synthetic (poly-time based)
-%     computational complexity theory.
-%   \end{enumerate}
-% \item Section 6: Related Work
-%   \begin{enumerate}
-%   \item ICC stuff
-%     \begin{enumerate}
-%     \item LFPL
-%     \item CONS-free
-%     \item Soft Affine Logic; which has the same underlying idea as
-%       CONS-free (is this a novel observation?)
-%     \item Other systems based on information flow.
-%     \end{enumerate}
-%   \item Explicit complexity: lambda-amor, and things by Deepak Garg
-%     and Marco Gaboardi.
-%   \item Recent CMU stuff on calf
-%   \end{enumerate}
+programs' resource usage as well as their extensional behaviour, and
+to theoretical use as a form of \emph{synthetic computational
+  complexity theory}. We discuss these possibilities further in
+\autoref{sec:conclusion}.
 
@@
   statements about polytime realisability to be reflected into
-  types. % We explain in \autoref{Sec:realisability-future} how this
-  % will have applications in QTT beyond the polytime systems presented
-  % here.
+  types.
 \item We demonstrate the utility of the combination of polytime and
@@
 
-% \subsection{Explicit Resource Tracking}
-
-% \begin{enumerate}
-% \item We still require diamonds to build zero and successor, but the
-%   iterator does not give them back.
-% \item In the simply typed case, we can annotate types with the static
-%   amount of potential they have per constructor.
-% \end{enumerate}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 under the type of polytime reductions given above.
-
-% \begin{displaymath}
-%   \begin{array}{l}
-%     \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A \\
-%     \mathrm{runWithOracle}\,(\mathsf{return}\,a)\,\mathit{bs} = \mathsf{just}\,a \\
-%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cTrue :: \mathit{bs}) = \mathrm{runWithOracle}\,x\,\mathit{bs} \\
-%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cFalse :: \mathit{bs}) = \mathrm{runWithOracle}\,y\,\mathit{bs} \\
-%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,\cNil = \mathsf{nothing}
-%   \end{array}
-% \end{displaymath}
 
@@
 The differencing operator of the natural number resource monoid can
-only supply as much fuel as is contained in the potential.
-% This makes
-% it suitable for amortised analysis, where we directly store potential
-% in data structures.
-For the two polynomial time systems, we need more sophisticated
-structures, both originally presented by Dal Lago and Hofmann. The
-fundamental idea with both is to represent potentials as pairs
-$(m,p)$, where $m$ is a natural number and $p$ is a polynomial. The
-$m$ tracks the ``size'' of data as it pertains to the number of times
-an operation will be repeated by iterating over it --- for example, an
-iterable natural number will have size equal to itself, but a
-non-iterable natural number may be assigned zero size. The polynomial
-$p$ tracks the complexity of a program as a function of the size of
-the input. This leads to a differencing operator that evaluates the
+only supply as much fuel as is contained in the potential.  For the
+two polynomial time systems, we need more sophisticated structures,
+both originally presented by Dal Lago and Hofmann. The fundamental
+idea with both is to represent potentials as pairs $(m,p)$, where $m$
+is a natural number and $p$ is a polynomial. The $m$ tracks the
+``size'' of data as it pertains to the number of times an operation
+will be repeated by iterating over it --- for example, an iterable
+natural number will have size equal to itself, but a non-iterable
+natural number may be assigned zero size. The polynomial $p$ tracks
+the complexity of a program as a function of the size of the
+input. This leads to a differencing operator that evaluates the
 polynomial with the size of the data:
@@
 $+$) used to compose sizes.
-
-% FIXME: difference between this definition and Hofmann-Dal Lago's: we
-% use polynomials with natural number coefficients instead of a function
-% that is bounded by a polynomial.
 
@@
 \AgdaModule{AmortisedModel.Bool} and \AgdaModule{AmortisedModel.List}.
-
-% \subsubsection{Realising QTT}
-
-% Summing up the previous sections, we have the following key theorem
-% paving the way to polytime soundness:
-% \begin{theorem}\label{thm:basic-qtt-realisability-model}
-%   For every resource monoid with a distinguished sub-monoid
-%   $(M, M_0)$, we obtain a realisability model of $\mathbb{N}$-QTT with
-%   $\Pi$ and $\Sigma$ types, a universe, extensional equality types,
-%   realisability reflection, and non-iterable datatypes.
-% \end{theorem}
 

### 2023-10-24 paper/paper.tex
[ paper ] fixes in related work

@@
 amortised complexity analysis arising from \cite{hofmann99lfpl}'s
-ideas, via the work of \cite{HofmannJ03}. More details are to be found
-in the survey paper of \citet{HoffmannJ22}. A approach is to track
-costs at the value level instead of the types. \citet{Danielsson08}
-describes a system that uses a ``tick'' effect to count steps of
-computation, which can be reasoned about via dependent
-types. \citet{NiuSGH22} take this idea further by employing a
-modality-based phase separation to ensure that tick counting never
+ideas, via the work of \citet{HofmannJ03}. More details are to be
+found in the survey paper of \citet{HoffmannJ22}. Another approach is
+to track costs at the value level instead of the
+types. \citet{Danielsson08} describes a system that uses a ``tick''
+effect to count steps of computation, which can be reasoned about via
+dependent types. \citet{NiuSGH22} take this idea further by employing
+a modality-based phase separation to ensure that tick counting never
 interferes with the functional business of
 programs. \citet{McCarthyFNFF16} is another tick effect based system
-in Coq. All of these techniques rely on the programmer correctly
-annotating the program with tick effects to count the resource usage
-they are interested in, in contrast to the intrinsic approach we have
-taken here.
+in Coq. All of these tick-counting techniques rely on the programmer
+correctly annotating the program with tick effects to count the
+resource usage they are interested in, in contrast to the intrinsic
+approach we have taken here.
 
@@
 systems. Other systems include systems such as those by
-\cite{CervesatoP02,KrishnaswamiPB15,Vakar14} which all use a strict
-separation between linear and non-linear variables. This strict
-separation would mean that we could not as easily move programs from
-the linear fragment into the types, as in
+\citet{CervesatoP02}, \citet{KrishnaswamiPB15}, and \citet{Vakar14}
+which all use a strict separation between linear and non-linear
+variables. This strict separation would mean that we could not as
+easily move programs from the linear fragment into the types, as in
 \autoref{sec:programming-polytime}. Systems that are more like QTT in
 that they do not have a strict separation of variables include those
-of \cite{MoonEO21,ChoudhuryEEW21,Abel18}. These systems differ from
-QTT in that they do not include a complete copy of unrestricted type
-theory as QTT does in its $\sigma = 0$ fragment, because they all
-track usage in types as well as terms, so it is not clear how to use
-them for unrestricted reasoning as we do with QTT. \citet{FuKS22}
-present a system that is closer to QTT but does not include a universe
-type.
+of \citet{MoonEO21}, \citet{ChoudhuryEEW21}, and \citet{Abel18}. These
+systems differ from QTT in that they do not include a complete copy of
+unrestricted type theory as QTT does in its $\sigma = 0$ fragment,
+because they all track usage in types as well as terms, so it is not
+clear how to use them for unrestricted reasoning as we do with
+QTT. \citet{FuKS22} present a system that is closer to QTT but does
+not include a universe type, which we used in \autoref{sec:classes} to
+be able to characterise complexity classes as predicates decidable in
+restricted complexity.
 

### 2023-10-24 paper/paper.tex
[ paper ] minor typo
[ paper ] more fixes

@@
 Like type formation, definitional equality of types always takes place
-in $0$d contexts. We will describe the definitional equality rules for
+in $0$-d contexts. We will describe the definitional equality rules for
 terms of each type as we introduce them. In QTT, it is possible for
@@
 $T$. $\Pi$-types support the usual $\beta\eta$ definitional
-equalities.
+equalities in both the $\sigma = 0$ and $\sigma = 1$ fragments.
 
@@
 mutually inverse: $\rIntro(\rElim(M)) \equiv M$, in both fragments,
-and $\rElim(\rIntro(M)) \equiv M$ in the $\sigma = 0$ fragment.
+and $\rElim(\rIntro(M)) \equiv M$ in the $\sigma = 0$ fragment. By
+congruence, the $\sigma = 1$ fragment's definitional equality affects
+the definitional equality of the $\sigma = 0$ fragment via the
+$\rIntro(-)$ constructor.
 
@@
 implementation is probabilistic. With a dependently-typed host
-language, adding probabilistic choice as an effect, and capturing the
+language, adding probabilistic choice as an effect and capturing the
 semantic constraint of BPP is straightforward.

### 2023-10-25 paper/paper.tex
[ paper ] more little fixes

@@
 be seen as representing polynomial time, differing in how the size of
-the input is measured. One approach is to define the input as a closed
-expression, including what we might think of as its input, and
+the problem to be computed is measured. One approach is to consider
+closed expressions, combining the program with its input, and
 computation of the result is polynomial time in the combined size. A
-second approach is that the input is ``externally'' provided, so we
+second approach is that the input is ``externally'' provided, where we
 consider open terms with a free variable representing the input. So a
@@
     Explainable by Construction}.
-  % FIXME: add Fred, when his comments have been included.
 \end{acks}

### 2023-10-25 paper/paper.tex
[ paper ] I was right
[ paper ] Fix acks

@@
 steps. Then, using the iterator above we can iterate $f$ over a
-$\tyNat$ representing the size of the input:\bob{CHECK THIS}
+$\tyNat$ representing the size of the input:
 \begin{displaymath}
@@
 Unlike in the \emph{Cons-free} system, this iterator does not raise
-the degree of the nested iterator directly. Rather the iterator $I_k$
-performs $\binom{n}{k}$ iterations. As observed by Aelhig and
-Schwictenberg, this is sufficient because the binomials form a basis
-for the vector space of all polynomials.
+the degree of the nested iterator directly. Rather, the iterator $I_k$
+on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
+Aelhig and Schwictenberg, this is sufficient because the binomials
+form a basis for the vector space of all polynomials.
 
@@
   comments on an earlier revision, and to the anonymous POPL reviewers
-  for their detailed and interesting reviews. This work was funded by
-  the \grantsponsor{EPSRC}{Engineering and Physical Sciences Research
-    Council}{https://www.ukri.org/about-us/epsrc/}. Grant number
+  for their detailed, perceptive, and interesting reviews. This work
+  was funded by the \grantsponsor{EPSRC}{Engineering and Physical
+    Sciences Research
+    Council}{https://www.ukri.org/about-us/epsrc/}: Grant number
   \grantnum{EPSRC}{EP/T026960/1}, \emph{AISEC: AI Secure and

### 2023-10-25 paper/paper.tex
[ paper ] complete typing rules

@@
   \inferrule*
-  {\Gamma \vdash M : S \\ \Gamma' \sqsubseteq \Gamma}
-  {\Gamma' \vdash M : S}
+  {\Gamma \vdash M \stackrel\sigma: S \\ \Gamma' \sqsubseteq \Gamma}
+  {\Gamma' \vdash M \stackrel\sigma: S}
 \end{mathpar}
@@
   \begin{array}{cl}
-    & \tmRec\,(\conSucc(M_d,M_n)\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
+    & \tmRec\,(\conSucc(M_d,M_n))\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
     \equiv & N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p]
@@
 
+\appendix
+
+\section{Typing Rules for QTT}
+
+The judgements of Quantitative Type Theory are as follows:
+\begin{mathpar}
+  \begin{array}{ll}
+    \Gamma~\isctxt&\textrm{contexts} \\
+%    \Gamma_1 \equiv \Gamma_2~\isctxt&\textrm{equal contexts} \\
+    \Gamma \vdash S~\istype&\textrm{types} \\
+    \Gamma \vdash S \equiv T~\istype&\textrm{equal types} \\
+  \end{array}
+
+  \begin{array}{ll}
+    \Gamma \vdash M \stackrel\sigma: S&\textrm{terms} \\
+    \Gamma \vdash M \equiv N \stackrel\sigma: S&\textrm{equal terms}
+  \end{array}
+\end{mathpar}
+In the term and term equality judgements, the usage $\sigma$ is either
+$0$ or $1$. We use $\rho$ and $\pi$ to range over arbitrary usages
+from the semiring $R$.
+
+\subsection{Context formation}
+\begin{mathpar}
+  \inferrule* [right=Emp]
+  { }
+  {\diamond~\isctxt}
+
+  \inferrule* [right=Ext]
+  {\Gamma~\isctxt \\ 0\Gamma \vdash S}
+  {\Gamma, x \stackrel\rho: S~\isctxt}
+\end{mathpar}
+
+\subsection{Type Equality}
+\begin{mathpar}
+  \inferrule* [right=Ty-Eq-Refl]
+  {0\Gamma \vdash S}
+  {0\Gamma \vdash S \equiv S}
+
+  \inferrule* [right=Ty-Eq-Symm]
+  {0\Gamma \vdash S \equiv T}
+  {0\Gamma \vdash T \equiv S}
+
+  \inferrule* [right=Ty-Eq-Tran]
+  {0\Gamma \vdash S \equiv T \\ 0\Gamma \vdash T \equiv U}
+  {0\Gamma \vdash S \equiv U}
+\end{mathpar}
+as well as congruence rules for each type formation rule (elided), and
+the universe eliminator.
+
+\subsection{Term Equality}
+\begin{mathpar}
+  \inferrule* [right=Tm-Eq-Refl]
+  {\Gamma \vdash M \stackrel\sigma: S}
+  {\Gamma \vdash M \equiv M \stackrel\sigma: S}
+
+  \inferrule* [right=Tm-Eq-Symm]
+  {\Gamma \vdash M \equiv N \stackrel\sigma: S}
+  {\Gamma \vdash N \equiv M \stackrel\sigma: S}
+
+  \inferrule* [right=Tm-Eq-Tran]
+  {\Gamma \vdash M \equiv N \stackrel\sigma: S \\
+    \Gamma \vdash N \equiv O \stackrel\sigma: S}
+  {\Gamma \vdash M \equiv O \stackrel\sigma: S}
+\end{mathpar}
+as well as congruence rules for each term formation rule (elided), and
+the specific $\beta\eta$-equalities for each type listed below.
+
+\subsection{Variables, conversion, sub-usaging}
+\begin{mathpar}
+  \inferrule* [right=Var]
+  {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
+  {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}
+
+  \inferrule* [right=Conv]
+  {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
+  {\Gamma \vdash M \stackrel\sigma: T}
+
+  \inferrule* [right=Sub]
+  {\Gamma \vdash M \stackrel\sigma: S \\ \Gamma' \sqsubseteq \Gamma}
+  {\Gamma' \vdash M \stackrel\sigma: S}
+\end{mathpar}
+
+\subsection{$\Pi$-types} Type formation:
+\begin{displaymath}
+  \inferrule* [right=Ty-Pi]
+  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
+  {0\Gamma \vdash (x \stackrel\pi: S) \to T}
+\end{displaymath}
+Introduction and elimination:
+\begin{mathpar}
+  \inferrule* [right=Tm-Lam]
+  {\Gamma, x \stackrel{\sigma\pi}: S \vdash M \stackrel\sigma: T}
+  {\Gamma \vdash \lambda x.M \stackrel\sigma: (x \stackrel\pi: S) \to T}
+
+  \inferrule* [right=Tm-App]
+  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: S) \to T \\
+    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
+    0\Gamma_1 = 0\Gamma_2 \\
+    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)
+  }
+  {\Gamma_1 + \pi\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
+\end{mathpar}
+$\beta\eta$-equalities (as well as congruences):
+\begin{mathpar}
+  \inferrule* [right=Tm-Eq-Pi$\beta$]
+  {\Gamma_1, x \stackrel{\sigma\pi}: S \vdash M \stackrel\sigma: T \\
+    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
+    (\sigma' = 0 \Leftrightarrow \pi = 0 \lor \sigma = 0)}
+  {\Gamma_1 + \pi\Gamma_2 \vdash (\lambda x.M)\, N \equiv M[N/x] \stackrel\sigma: T[N/x]}
+
+  \inferrule* [right=Tm-Eq-Pi$\eta$]
+  {\Gamma \vdash M \stackrel\sigma: (x \stackrel\pi: S) \to T}
+  {\Gamma \vdash \lambda x .M\,x \equiv M \stackrel\sigma: (x \stackrel\pi: S) \to T}
+\end{mathpar}
+
+\subsection{$\Sigma$-types}
+Type formation:
+\begin{mathpar}
+  \inferrule* [right=Ty-Tensor]
+  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
+  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T~\istype}
+
+  \inferrule* [right=Ty-Unit]
+  {0\Gamma~\isctxt}
+  {0\Gamma \vdash I~\istype}
+\end{mathpar}
+Introduction:
+\begin{mathpar}
+  \inferrule* [right=Tm-Pair]
+  {\Gamma_1 \vdash M \stackrel{\sigma'}: S \\
+    \Gamma_2 \vdash N \stackrel\sigma: T[M/x] \\
+    0\Gamma_1 = 0\Gamma_2 \\
+    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)}
+  {\pi\Gamma_1 + \Gamma_2 \vdash (M, N) \stackrel\sigma: (x \stackrel\pi: S) \otimes T}
+
+  \inferrule* [right=Tm-Unit]
+  {0\Gamma~\isctxt}
+  {0\Gamma \vdash * \stackrel\sigma: I}
+\end{mathpar}
+$\sigma = 0$ fragment eliminators:
+\begin{mathpar}
+  \inferrule* [right=Tm-Fst]
+  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
+  {\Gamma \vdash \mathrm{fst}(M) \stackrel0: S}
+
+  \inferrule* [right=Tm-Snd]
+  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
+  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
+\end{mathpar}
+Linear eliminators:
+\begin{mathpar}
+  \inferrule* [right=Tm-Let-Pair]
+  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C~\istype \\
+    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
+    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
+
+  \inferrule* [right=Tm-Let-Unit]
+  {0\Gamma_1, x \stackrel0: I \vdash C~\istype \\
+    \Gamma_1 \vdash M \stackrel\sigma: I \\
+    \Gamma_2 \vdash N \stackrel\sigma: C[*/x] \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1+\Gamma_2 \vdash \mathrm{let}~* = M~\mathrm{in}~N \stackrel\sigma: C[M/x]}
+\end{mathpar}
+$\beta$-equality for both fragments:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C~\istype \\
+    \Gamma_1 \vdash M_1 \stackrel{\sigma'}: A \\
+    \Gamma_2 \vdash M_2 \stackrel\sigma: B[M/x] \\
+    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0) \\
+    \Gamma_3, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
+    0\Gamma_1 = 0\Gamma_2 = 0\Gamma_3}
+  {\pi\Gamma_1 + \Gamma_2 + \Gamma_3 \vdash \mathrm{let}~(x,y) = (M_1,M_2)~\mathrm{in}~N \equiv N[M_1/x,M_2/y]\stackrel\sigma: C[(M_1,M_2)/z]}
+
+  \inferrule*
+  {0\Gamma, x \stackrel0: I \vdash C~\istype \\
+    \Gamma \vdash N \stackrel\sigma: C[*/x]}
+  {\Gamma \vdash \mathrm{let}~* = *~\mathrm{in}~N \equiv N[*/x] \stackrel\sigma: C[*/x]}
+\end{mathpar}
+$\beta\eta$-equalities for the $\sigma = 0$ fragment:
+\begin{mathpar}
+  \inferrule* [right=Eq-Pair-Fst]
+  {0\Gamma \vdash M \stackrel0: S \\
+    0\Gamma \vdash N \stackrel0: T[M/x]}
+  {0\Gamma \vdash \mathrm{fst}(M, N) \equiv M \stackrel0: S}
+
+  \inferrule* [right=Eq-Pair-Snd]
+  {0\Gamma \vdash M \stackrel0: S \\
+    0\Gamma \vdash N \stackrel0: T[M/x]}
+  {0\Gamma \vdash \mathrm{snd}(M, N) \equiv N \stackrel0: T[M/x]}
+
+  \inferrule* [right=Eq-Unit-$\eta$]
+  {0\Gamma \vdash M \stackrel0: I}
+  {0\Gamma \vdash M \equiv * \stackrel0: I}
+
+  \inferrule* [right=Eq-Pair-$\eta$]
+  {0\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
+  {0\Gamma \vdash (\mathrm{fst}(M), \mathrm{snd}(M)) \equiv M \stackrel0: (x \stackrel\pi: S) \otimes T}
+
+  \inferrule*
+  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
+    0\Gamma \vdash M \stackrel0: (x \stackrel\pi: A) \otimes B \\
+    0\Gamma, x \stackrel0: A, y \stackrel0: B \vdash N \stackrel0: C[(x,y)/z]}
+  {0\Gamma \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \equiv N[\mathrm{fst}(M)/x,\mathrm{snd}(M)/y] \stackrel0: C[M/z]}
+\end{mathpar}
+
+\subsection{Identity Type} Type formation, introduction, reflection
+and $\eta$-law:
+\begin{mathpar}
+  \inferrule* [right=Ty-Id]
+  {0\Gamma \vdash S~\istype \\
+    0\Gamma \vdash M \stackrel0: S \\
+    0\Gamma \vdash N \stackrel0: S}
+  {0\Gamma \vdash M =_S N~\istype}
+
+  \inferrule* [right=Id-Refl]
+  {\Gamma \vdash M \stackrel\sigma: S}
+  {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
+
+  \inferrule* [right=Id-Reflect]
+  {\Gamma \vdash N \stackrel0: M_1 =_S M_2}
+  {\Gamma \vdash M_1 \equiv M_2 \stackrel0: S}
+
+  \inferrule* [right=Id-Uniq]
+  {0\Gamma \vdash P \stackrel0: M =_A M}
+  {0\Gamma \vdash P \equiv \mathrm{refl}(M) : M=_A M}
+\end{mathpar}
+
+\subsection{Universe}
+
+\newcommand{\TySet}{\mathsf{U}}
+\newcommand{\TyEl}{\mathsf{El}}
+Type formation:
+\begin{displaymath}
+  \inferrule* [right=Ty-U]
+  {0\Gamma~\isctxt}
+  {0\Gamma \vdash \TySet~\istype}
+\end{displaymath}
+Introduction (also with introduction rules for all other type formers except $\TySet$):
+\begin{mathpar}
+  \inferrule* [right=Tm-U-Bool]
+  {\Gamma~\isctxt}
+  {\Gamma \vdash \BoolTy \stackrel\sigma: \TySet}
+
+  \inferrule* [right=Tm-U-Tensor]
+  {\Gamma \vdash M \stackrel\sigma: \TySet \\
+    \Gamma, x \stackrel\rho: \TyEl(M) \vdash N \stackrel\sigma: \TySet}
+  {\Gamma \vdash (x \stackrel\pi: M) \otimes N \stackrel\sigma: \TySet}
+
+  \ldots
+
+  \inferrule* [right=Tm-U-Eq]
+  {\Gamma \vdash M \stackrel\sigma: \TySet \\
+    \Gamma \vdash N_1 \stackrel\sigma: \TyEl(M) \\
+    \Gamma \vdash N_2 \stackrel\sigma: \TyEl(M)}
+  {\Gamma \vdash N_1 =_M N_2 \stackrel\sigma: \TySet}
+\end{mathpar}
+Elimination:
+\begin{displaymath}
+  \inferrule* [right=Ty-El]
+  {0\Gamma \vdash M \stackrel0: \TySet}
+  {0\Gamma \vdash \TyEl(M)~\istype}
+\end{displaymath}
+Equality:
+\begin{mathpar}
+  \inferrule* [right=Ty-Eq-El-Pi]
+  {0\Gamma \vdash M \stackrel0: \TySet \\
+    0\Gamma, x \stackrel0: \TyEl(M) \vdash N \stackrel0: \TySet}
+  {0\Gamma \vdash \TyEl((x \stackrel\rho: M) \to N) \equiv (x \stackrel\rho: \TyEl(M)) \to \TyEl(N)}
+
+  \inferrule* [right=Ty-Eq-El-Bool]
+  {0\Gamma~\isctxt}
+  {0\Gamma \vdash \TyEl(\BoolTy) \equiv \BoolTy}
+
+  \ldots
+
+  \inferrule* [right=Tm-Eq-Set-Pi-Cong]
+  {0\Gamma \vdash M \equiv M' \stackrel0: \TySet \\
+    0\Gamma, x \stackrel0: \TyEl(M) \vdash N \equiv N' \stackrel0: \TySet}
+  {0\Gamma \vdash (x \stackrel\pi: M) \to N \equiv (x \stackrel\pi: M') \to N' \stackrel0: \TySet}
+\end{mathpar}
+
+\subsection{Booleans}
+Formation, introduction, and elimination:
+\begin{mathpar}
+  \inferrule*
+  {\Gamma~\isctxt}
+  {\Gamma \vdash \BoolTy~\istype}
+
+  \inferrule*
+  {\Gamma~\isctxt}
+  {0\Gamma \vdash \cTrue, \cFalse \stackrel\sigma: \BoolTy}
+
+  \inferrule*
+  {0\Gamma_1, x \stackrel0: \BoolTy \vdash P~\istype \\
+    \Gamma_1 \vdash M \stackrel\sigma: \BoolTy \\
+    \Gamma_2 \vdash N_t \stackrel\sigma: P[\cTrue/x] \\
+    \Gamma_2 \vdash N_f \stackrel\sigma: P[\cFalse/x] \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \If_{x.P}\: M \: \Then \: N_t \: \Else \: N_f \stackrel\sigma: P[M/x]}
+\end{mathpar}
+$\beta$-equalities:
+\begin{mathpar}
+  \inferrule* [right=Tm-Eq-True$\beta$]
+  {0\Gamma, z \stackrel0: \BoolTy \vdash P \\
+    \Gamma \vdash M_t \stackrel\sigma: P[\mathrm{true}/z] \\
+    \Gamma \vdash M_f \stackrel\sigma: P[\mathrm{false}/z]}
+  {\Gamma \vdash \If_{x.P}\: \cTrue \: \Then \: N_t \: \Else \: N_f \equiv N_t \stackrel\sigma: P[\mathrm{true}/z]}
+
+  \inferrule* [right=Tm-Eq-False$\beta$]
+  {0\Gamma, z \stackrel0: \BoolTy \vdash P \\
+    \Gamma \vdash M_t \stackrel\sigma: P[\mathrm{true}/z] \\
+    \Gamma \vdash M_f \stackrel\sigma: P[\mathrm{false}/z]}
+  {\Gamma \vdash \If_{x.P}\: \cFalse \: \Then \: N_t \: \Else \: N_f \equiv N_f \stackrel\sigma: P[\mathrm{false}/z]}
+\end{mathpar}
+
+\subsection{Lists} Formation and introduction:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma \vdash T~\istype}
+  {0\Gamma \vdash \ListTy(T)~\istype}
+
+  \inferrule*
+  {\Gamma \vdash T~\istype}
+  {0\Gamma \vdash \cNil \stackrel\sigma: \ListTy(T)}
+
+  \inferrule*
+  {\Gamma_1 \vdash M \stackrel\sigma: T \\
+    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
+\end{mathpar}
+Case analysis:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
+    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
+    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
+    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
+\end{mathpar}
+$\beta$-equalities for case analysis:
+\begin{mathpar}
+  \inferrule* [right=Eq-List-Match-Nil]
+  {0\Gamma, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
+    \Gamma \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
+    \Gamma, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x]}
+  {\Gamma \vdash \Match_{x.P}\,\cNil\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \equiv N_1 \stackrel\sigma: P[\cNil/x]}
+
+  \inferrule* [right=Eq-List-Match-Cons]
+  {0\Gamma_1, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
+    \Gamma_1 \vdash M_1 \stackrel\sigma: T \\
+    \Gamma_2 \vdash M_2 \stackrel\sigma: \ListTy(T) \\
+    \Gamma_3 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
+    \Gamma_3, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
+    0\Gamma_1 = 0\Gamma_2 = 0\Gamma_3}
+  {\Gamma_1 + \Gamma_2 + \Gamma_3 \vdash {\begin{array}[m]{@{}l}
+                                            \Match_{x.P}\,(\cCons(M_1,M_2))\\
+                                            \qquad\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\}\\
+                                            \equiv N_2[M_1/h,M_2/t] \end{array}} \stackrel\sigma: P[\cCons(M_1,M_2)/x]}
+\end{mathpar}
+$\sigma = 0$ recursive eliminator:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
+    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
+    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
+    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
+  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
+\end{mathpar}
+$\beta$-equalities for recursion:
+\begin{mathpar}
+  \inferrule* [right=Eq-List-Rec-Nil]
+  {0\Gamma, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
+    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
+    0\Gamma, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
+  {0\Gamma \vdash \mathrm{recList}_{x.P}\,\cNil\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \equiv N_1 \stackrel0: P[\cNil/x]}
+
+  \inferrule* [right=Eq-List-Rec-Cons]
+  {0\Gamma, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
+    0\Gamma \vdash M_1 \stackrel0: T \\
+    0\Gamma \vdash M_2 \stackrel0: \ListTy(T) \\
+    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
+    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T) \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
+  {0\Gamma \vdash {\begin{array}[m]{@{}l}
+                     \mathrm{recList}_{x.P}\,(\cCons(M_1,M_2))\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\}\\
+                     \equiv N_2[\begin{array}[t]{@{}l} M_1/h,M_2/t,\\
+                                  \mathrm{recList}_{x,p}\,M_2\,\{\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\}/p]\end{array} \end{array}} \stackrel0: P[\cCons(M_1,M_2)/x]}
+\end{mathpar}
+
+\subsection{Cons-Free Naturals} Formation and introduction:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma~\isctxt}
+  {0\Gamma \vdash \tyNat~\istype}
+
+  \inferrule*
+  {\Gamma~\isctxt}
+  {\Gamma \vdash \conZero \stackrel0: \tyNat}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel0: \tyNat}
+  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
+\end{mathpar}
+Duplication:
+\begin{mathpar}
+  \inferrule*
+  {\Gamma \vdash M \stackrel\sigma: \tyNat}
+  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel0: \tyNat}
+  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
+\end{mathpar}
+Eliminator:
+\begin{displaymath}
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
+    0\Gamma \vdash N_z \stackrel\sigma: P[\conZero/x] \\\\
+    0\Gamma, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma: P[\conSucc(n)/x]}
+  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
+\end{displaymath}
+$\beta$-equalties (only available in the $\sigma = 0$ fragment):
+\begin{mathpar}
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    0\Gamma \vdash N_z \stackrel0: P[\conZero/x] \\\\
+    0\Gamma, n \stackrel0: \tyNat, p \stackrel0: P[n/x] \vdash N_s \stackrel0: P[\conSucc(n)/x]}
+  {0\Gamma \vdash \tmRec_{x.P}\,\conZero\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \equiv N_z \stackrel0: P[\conZero/x]}
+
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    0\Gamma \vdash M \stackrel0: \tyNat \\\\
+    0\Gamma \vdash N_z \stackrel0: P[\conZero/x] \\\\
+    0\Gamma, n \stackrel0: \tyNat, p \stackrel0: P[n/x] \vdash N_s \stackrel0: P[\conSucc(n)/x]}
+  {0\Gamma \vdash {\begin{array}{@{}l}
+                    \tmRec_{x.P}\,(\conSucc(M))\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \\
+                    \equiv N_s[M/n,\tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\}/p] \end{array}} \stackrel0: P[\conSucc(M)/x]}
+\end{mathpar}
+
+\subsection{LFPL Diamonds} Formation, introduction, and $\eta$-law:
+\begin{mathpar}
+  \inferrule*
+  {\Gamma~\isctxt}
+  {0\Gamma \vdash \Diamond~\istype}
+
+  \inferrule*
+  {\Gamma~\isctxt}
+  {0\Gamma \vdash * \stackrel0: \Diamond}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel0: \Diamond}
+  {\Gamma \vdash M \equiv * \stackrel0: \Diamond}
+\end{mathpar}
+
+\subsection{LFPL Naturals} Type formation and introduction:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma~\isctxt}
+  {0\Gamma \vdash \tyNat~\istype}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel\sigma: \Diamond}
+  {\Gamma \vdash \conZero(M) \stackrel\sigma: \tyNat}
+
+  \inferrule*
+  {\Gamma_1 \vdash M \stackrel\sigma: \Diamond \\
+    \Gamma_2 \vdash N \stackrel\sigma: \tyNat \\
+    0\Gamma_1 = 0\Gamma_2}
+  {\Gamma_1 + \Gamma_2 \vdash \conSucc(M,N) \stackrel\sigma: \tyNat}
+\end{mathpar}
+Elimination:
+\begin{mathpar}
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
+  {\Gamma \vdash \tmRec\,M\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
+\end{mathpar}
+$\beta$-equalities, available in both fragments:
+\begin{mathpar}
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    \Gamma \vdash M \stackrel\sigma: \Diamond \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
+  {\Gamma \vdash \tmRec\,\conZero(M)\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \equiv N_z[M/d] \stackrel\sigma: P[\conZero(*)/x]}
+
+  \mprset{flushleft}
+  \inferrule*
+  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
+    \Gamma_1 \vdash M_d \stackrel\sigma: \Diamond \\\\
+    \Gamma_2 \vdash M_n \stackrel\sigma: \tyNat \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
+    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x] \\\\
+    0\Gamma_1 = 0\Gamma_2 = 0\Gamma}
+  {\Gamma_1 + \Gamma_2 \vdash {\begin{array}{@{}l}
+                                 \tmRec\,(\conSucc(M_d,M_n))\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
+                                 \equiv N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p] \end{array}} \stackrel\sigma: P[\conSucc(*,M_n)/x]}
+\end{mathpar}
+
+\subsection{Realisability Reflection} Type formation, introduction and
+elimination:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma \vdash A~\istype}
+  {0\Gamma \vdash \Rtype(A)~\istype}
+
+  \inferrule*
+  {0\Gamma \vdash M \stackrel1: A}
+  {0\Gamma \vdash \rIntro(M) \stackrel\sigma: \Rtype(A)}
+
+  \inferrule*
+  {\Gamma \vdash M \stackrel\sigma: \Rtype(A)}
+  {\Gamma \vdash \rElim(M) \stackrel{\sigma'}: A}
+\end{mathpar}
+Equalities:
+\begin{mathpar}
+  \inferrule*
+  {0\Gamma \vdash M \stackrel1: A}
+  {0\Gamma \vdash \rElim(\rIntro(M)) \equiv M \stackrel\sigma: A}
+
+  \inferrule*
+  {0\Gamma \vdash M \stackrel\sigma: \Rtype(A)}
+  {0\Gamma \vdash \rIntro(\rElim(M)) \equiv M \stackrel\sigma: A}
+\end{mathpar}
+
 \end{document}

### 2023-10-25 paper/paper.tex
[ paper ] BPP, not PP

@@
 of polytime to further functionally characterise the classes of
-non-deterministic and probabilistic polynomial time.
+non-deterministic and bounded-error probabilistic polynomial time.
 
@@
   types to give monadic presentations of the complexity classes of
-  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
-  (PP) time. Since these classes rely on specific correctness
-  criteria, it is not possible to capture them in a simply typed
-  system for polytime.
+  Non-deterministic Polynomial (NP) time and Bounded-error
+  Probabilistic Polynomial (BPP) time. Since these classes rely on
+  specific correctness criteria, it is not possible to capture them in
+  a simply typed system for polytime.
 \item We prove the polytime soundness of our systems via a

### 2023-10-25 paper/paper.tex
[ paper ] typos and concepts

@@
 \setcopyright{acmcopyright}
-\copyrightyear{2020}
-\acmYear{2020}
+\copyrightyear{2024}
+\acmYear{2024}
 \acmDOI{10.1145/1122445.1122456}
@@
    <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+       <concept_id>10003752.10003790.10003801</concept_id>
+       <concept_desc>Theory of computation~Linear logic</concept_desc>
        <concept_significance>500</concept_significance>
@@
    <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003777.10003778</concept_id>
+       <concept_desc>Theory of computation~Complexity classes</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003777.10003787</concept_id>
+       <concept_desc>Theory of computation~Complexity theory and logic</concept_desc>
        <concept_significance>300</concept_significance>
@@
    <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
+       <concept_id>10003752.10010124.10010131.10010137</concept_id>
+       <concept_desc>Theory of computation~Categorical semantics</concept_desc>
        <concept_significance>100</concept_significance>
@@
 
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
+\ccsdesc[500]{Theory of computation~Linear logic}
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Theory of computation~Complexity classes}
+\ccsdesc[300]{Theory of computation~Complexity theory and logic}
+\ccsdesc[100]{Theory of computation~Categorical semantics}
 
@@
 
+In this section, we give an overview of the rules of QTT.
+
+The full rules are presented in \autoref{sec:rules}.
+
+The full rules are presented in the online appendix at
+
 \subsubsection{Natural-number usages}
@@
 regard it as usage $m$. This makes the system more like affine linear
-logic, since $m \leq 0$\footnote{Reverse ordering!} for all $m$,
+logic, since $m \sqsubseteq 0$\footnote{Reverse ordering!} for all $m$,
 matching the system in \autoref{sec:linear-intro}. We do \emph{not}
@@
 flexibility in QTT in allowing additional power in the $\sigma = 0$
-fragment, where we do not need to care about polytime
+fragment where we do not need to care about polytime
 realisability. Formation and introduction are given by the rules:
@@
 Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
-decision problem $(A,P)$ is deciable in polytime. We make three notes
+decision problem $(A,P)$ is decidable in polytime. We make three notes
 about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
@@
 $X \subseteq A \times M \times \ValSet$ and we define the ordering
-$X \leq Y$ to hold iff there exists realising expression
+$X \leq Y$ to hold iff there exists a realising expression
 $E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
@@
   a realising value for
-  $\llbracket M \rrbracket(n) \in \llbracket X \rrbracket(n)$.
+  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
 \end{theorem}
@@
   is a realising value for
-  $\llbracket M \rrbracket(n) \in \llbracket X \rrbracket(n)$.
+  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
 \end{theorem}
@@
 \autoref{sec:specific-monoids}. Investigation of such a system may
-yield a system that tracks intrinsic costs precisely.
+yield a system that tracks the intrinsic cost of programs precisely
+and explicitly.
 
@@
 realising expression $E$ that completes in polynomial time, and then
-writing polytime enodings into 3-SAT. We hope that the addition of
+writing polytime encodings into 3-SAT. We hope that the addition of
 such an axiom to our system would lead to an expressive machine-free
@@
 
-\section{Typing Rules for QTT}
+\section{Typing Rules for QTT/Cons-free and QTT/LFPL}
+\label{sec:rules}
 

### 2023-10-25 paper/paper.tex
[ paper ] more minor fixes, fix binomials, and tidy rules

@@
 
-Typical Type Theory only speaks of the public face that programs
+Typically, Type Theory only speaks of the public face that programs
 present to the world -- if you input things like this, you get things
@@
 will have linear functions and $\otimes$-products. Contexts are
-treated up to permutation, so uses of exchange are implicit.
+treated up to permutation of entries, so uses of exchange are
+implicit.
 \begin{mathpar}
@@
       \conZero(d)&\mapsto&(\conZero(d),s); \\
-      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
+      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
     \end{array}
@@
 
-In this section, we give an overview of the rules of QTT.
-
-The full rules are presented in \autoref{sec:rules}.
-
-The full rules are presented in the online appendix at
+In this section, we give an overview of the rules of QTT. The full
+rules are presented in \autoref{sec:rules}.
+
+\bob{The full rules are presented in the online appendix at}
 
@@
     (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
-    \quad \left((a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a\right)
+    \, \left((a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a\right)
   \end{array}
@@
   \ldots
-
-  \inferrule* [right=Tm-U-Eq]
-  {\Gamma \vdash M \stackrel\sigma: \TySet \\
-    \Gamma \vdash N_1 \stackrel\sigma: \TyEl(M) \\
-    \Gamma \vdash N_2 \stackrel\sigma: \TyEl(M)}
-  {\Gamma \vdash N_1 =_M N_2 \stackrel\sigma: \TySet}
 \end{mathpar}
@@
 \begin{mathpar}
-  \inferrule* [right=Ty-Eq-El-Pi]
-  {0\Gamma \vdash M \stackrel0: \TySet \\
-    0\Gamma, x \stackrel0: \TyEl(M) \vdash N \stackrel0: \TySet}
-  {0\Gamma \vdash \TyEl((x \stackrel\rho: M) \to N) \equiv (x \stackrel\rho: \TyEl(M)) \to \TyEl(N)}
-
-  \inferrule* [right=Ty-Eq-El-Bool]
-  {0\Gamma~\isctxt}
-  {0\Gamma \vdash \TyEl(\BoolTy) \equiv \BoolTy}
-
-  \ldots
-
-  \inferrule* [right=Tm-Eq-Set-Pi-Cong]
-  {0\Gamma \vdash M \equiv M' \stackrel0: \TySet \\
-    0\Gamma, x \stackrel0: \TyEl(M) \vdash N \equiv N' \stackrel0: \TySet}
-  {0\Gamma \vdash (x \stackrel\pi: M) \to N \equiv (x \stackrel\pi: M') \to N' \stackrel0: \TySet}
+  % \inferrule* [right=Ty-Eq-El-Pi]
+  % {0\Gamma \vdash M \stackrel0: \TySet \\
+  %   0\Gamma, x \stackrel0: \TyEl(M) \vdash N \stackrel0: \TySet}
+  % {0\Gamma \vdash \TyEl((x \stackrel\rho: M) \to N) \equiv (x \stackrel\rho: \TyEl(M)) \to \TyEl(N)~\istype}
+
+  % \inferrule* [right=Ty-Eq-El-Bool]
+  % {0\Gamma~\isctxt}
+  % {0\Gamma \vdash \TyEl(\BoolTy) \equiv \BoolTy~\istype}
+
+  % \ldots
+
+  \inferrule* [right=Ty-Eq-El-Cong]
+  {0\Gamma \vdash M \equiv N \stackrel0: \TySet}
+  {0\Gamma \vdash \TyEl(M) \equiv \TyEl(N)~\istype}
 \end{mathpar}

### 2023-10-26 paper/paper.tex
[ paper ] final changes and iteration count

@@
 In this section, we give an overview of the rules of QTT. The full
-rules are presented in \autoref{sec:rules}.
-
-\bob{The full rules are presented in the online appendix at}
+rules, including all equality rules, are presented in the online
+appendix for this paper \cite{atkey2023polydep-online}.
+
+%The full rules are presented in \autoref{sec:rules}.
 
@@
 
-\appendix
-
-\section{Typing Rules for QTT/Cons-free and QTT/LFPL}
-\label{sec:rules}
-
-The judgements of Quantitative Type Theory are as follows:
-\begin{mathpar}
-  \begin{array}{ll}
-    \Gamma~\isctxt&\textrm{contexts} \\
-%    \Gamma_1 \equiv \Gamma_2~\isctxt&\textrm{equal contexts} \\
-    \Gamma \vdash S~\istype&\textrm{types} \\
-    \Gamma \vdash S \equiv T~\istype&\textrm{equal types} \\
-  \end{array}
-
-  \begin{array}{ll}
-    \Gamma \vdash M \stackrel\sigma: S&\textrm{terms} \\
-    \Gamma \vdash M \equiv N \stackrel\sigma: S&\textrm{equal terms}
-  \end{array}
-\end{mathpar}
-In the term and term equality judgements, the usage $\sigma$ is either
-$0$ or $1$. We use $\rho$ and $\pi$ to range over arbitrary usages
-from the semiring $R$.
-
-\subsection{Context formation}
-\begin{mathpar}
-  \inferrule* [right=Emp]
-  { }
-  {\diamond~\isctxt}
-
-  \inferrule* [right=Ext]
-  {\Gamma~\isctxt \\ 0\Gamma \vdash S}
-  {\Gamma, x \stackrel\rho: S~\isctxt}
-\end{mathpar}
-
-\subsection{Type Equality}
-\begin{mathpar}
-  \inferrule* [right=Ty-Eq-Refl]
-  {0\Gamma \vdash S}
-  {0\Gamma \vdash S \equiv S}
-
-  \inferrule* [right=Ty-Eq-Symm]
-  {0\Gamma \vdash S \equiv T}
-  {0\Gamma \vdash T \equiv S}
-
-  \inferrule* [right=Ty-Eq-Tran]
-  {0\Gamma \vdash S \equiv T \\ 0\Gamma \vdash T \equiv U}
-  {0\Gamma \vdash S \equiv U}
-\end{mathpar}
-as well as congruence rules for each type formation rule (elided), and
-the universe eliminator.
-
-\subsection{Term Equality}
-\begin{mathpar}
-  \inferrule* [right=Tm-Eq-Refl]
-  {\Gamma \vdash M \stackrel\sigma: S}
-  {\Gamma \vdash M \equiv M \stackrel\sigma: S}
-
-  \inferrule* [right=Tm-Eq-Symm]
-  {\Gamma \vdash M \equiv N \stackrel\sigma: S}
-  {\Gamma \vdash N \equiv M \stackrel\sigma: S}
-
-  \inferrule* [right=Tm-Eq-Tran]
-  {\Gamma \vdash M \equiv N \stackrel\sigma: S \\
-    \Gamma \vdash N \equiv O \stackrel\sigma: S}
-  {\Gamma \vdash M \equiv O \stackrel\sigma: S}
-\end{mathpar}
-as well as congruence rules for each term formation rule (elided), and
-the specific $\beta\eta$-equalities for each type listed below.
-
-\subsection{Variables, conversion, sub-usaging}
-\begin{mathpar}
-  \inferrule* [right=Var]
-  {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
-  {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}
-
-  \inferrule* [right=Conv]
-  {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
-  {\Gamma \vdash M \stackrel\sigma: T}
-
-  \inferrule* [right=Sub]
-  {\Gamma \vdash M \stackrel\sigma: S \\ \Gamma' \sqsubseteq \Gamma}
-  {\Gamma' \vdash M \stackrel\sigma: S}
-\end{mathpar}
-
-\subsection{$\Pi$-types} Type formation:
-\begin{displaymath}
-  \inferrule* [right=Ty-Pi]
-  {0\Gamma \vdash S \\ 0\Gamma, x \stackrel0: S \vdash T}
-  {0\Gamma \vdash (x \stackrel\pi: S) \to T}
-\end{displaymath}
-Introduction and elimination:
-\begin{mathpar}
-  \inferrule* [right=Tm-Lam]
-  {\Gamma, x \stackrel{\sigma\pi}: S \vdash M \stackrel\sigma: T}
-  {\Gamma \vdash \lambda x.M \stackrel\sigma: (x \stackrel\pi: S) \to T}
-
-  \inferrule* [right=Tm-App]
-  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: S) \to T \\
-    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
-    0\Gamma_1 = 0\Gamma_2 \\
-    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)
-  }
-  {\Gamma_1 + \pi\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
-\end{mathpar}
-$\beta\eta$-equalities (as well as congruences):
-\begin{mathpar}
-  \inferrule* [right=Tm-Eq-Pi$\beta$]
-  {\Gamma_1, x \stackrel{\sigma\pi}: S \vdash M \stackrel\sigma: T \\
-    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
-    (\sigma' = 0 \Leftrightarrow \pi = 0 \lor \sigma = 0)}
-  {\Gamma_1 + \pi\Gamma_2 \vdash (\lambda x.M)\, N \equiv M[N/x] \stackrel\sigma: T[N/x]}
-
-  \inferrule* [right=Tm-Eq-Pi$\eta$]
-  {\Gamma \vdash M \stackrel\sigma: (x \stackrel\pi: S) \to T}
-  {\Gamma \vdash \lambda x .M\,x \equiv M \stackrel\sigma: (x \stackrel\pi: S) \to T}
-\end{mathpar}
-
-\subsection{$\Sigma$-types}
-Type formation:
-\begin{mathpar}
-  \inferrule* [right=Ty-Tensor]
-  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
-  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T~\istype}
-
-  \inferrule* [right=Ty-Unit]
-  {0\Gamma~\isctxt}
-  {0\Gamma \vdash I~\istype}
-\end{mathpar}
-Introduction:
-\begin{mathpar}
-  \inferrule* [right=Tm-Pair]
-  {\Gamma_1 \vdash M \stackrel{\sigma'}: S \\
-    \Gamma_2 \vdash N \stackrel\sigma: T[M/x] \\
-    0\Gamma_1 = 0\Gamma_2 \\
-    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0)}
-  {\pi\Gamma_1 + \Gamma_2 \vdash (M, N) \stackrel\sigma: (x \stackrel\pi: S) \otimes T}
-
-  \inferrule* [right=Tm-Unit]
-  {0\Gamma~\isctxt}
-  {0\Gamma \vdash * \stackrel\sigma: I}
-\end{mathpar}
-$\sigma = 0$ fragment eliminators:
-\begin{mathpar}
-  \inferrule* [right=Tm-Fst]
-  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
-  {\Gamma \vdash \mathrm{fst}(M) \stackrel0: S}
-
-  \inferrule* [right=Tm-Snd]
-  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
-  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
-\end{mathpar}
-Linear eliminators:
-\begin{mathpar}
-  \inferrule* [right=Tm-Let-Pair]
-  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C~\istype \\
-    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
-    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
-    0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
-
-  \inferrule* [right=Tm-Let-Unit]
-  {0\Gamma_1, x \stackrel0: I \vdash C~\istype \\
-    \Gamma_1 \vdash M \stackrel\sigma: I \\
-    \Gamma_2 \vdash N \stackrel\sigma: C[*/x] \\
-    0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1+\Gamma_2 \vdash \mathrm{let}~* = M~\mathrm{in}~N \stackrel\sigma: C[M/x]}
-\end{mathpar}
-$\beta$-equality for both fragments:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C~\istype \\
-    \Gamma_1 \vdash M_1 \stackrel{\sigma'}: A \\
-    \Gamma_2 \vdash M_2 \stackrel\sigma: B[M/x] \\
-    \sigma' = 0 \Leftrightarrow (\pi = 0 \lor \sigma = 0) \\
-    \Gamma_3, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
-    0\Gamma_1 = 0\Gamma_2 = 0\Gamma_3}
-  {\pi\Gamma_1 + \Gamma_2 + \Gamma_3 \vdash \mathrm{let}~(x,y) = (M_1,M_2)~\mathrm{in}~N \equiv N[M_1/x,M_2/y]\stackrel\sigma: C[(M_1,M_2)/z]}
-
-  \inferrule*
-  {0\Gamma, x \stackrel0: I \vdash C~\istype \\
-    \Gamma \vdash N \stackrel\sigma: C[*/x]}
-  {\Gamma \vdash \mathrm{let}~* = *~\mathrm{in}~N \equiv N[*/x] \stackrel\sigma: C[*/x]}
-\end{mathpar}
-$\beta\eta$-equalities for the $\sigma = 0$ fragment:
-\begin{mathpar}
-  \inferrule* [right=Eq-Pair-Fst]
-  {0\Gamma \vdash M \stackrel0: S \\
-    0\Gamma \vdash N \stackrel0: T[M/x]}
-  {0\Gamma \vdash \mathrm{fst}(M, N) \equiv M \stackrel0: S}
-
-  \inferrule* [right=Eq-Pair-Snd]
-  {0\Gamma \vdash M \stackrel0: S \\
-    0\Gamma \vdash N \stackrel0: T[M/x]}
-  {0\Gamma \vdash \mathrm{snd}(M, N) \equiv N \stackrel0: T[M/x]}
-
-  \inferrule* [right=Eq-Unit-$\eta$]
-  {0\Gamma \vdash M \stackrel0: I}
-  {0\Gamma \vdash M \equiv * \stackrel0: I}
-
-  \inferrule* [right=Eq-Pair-$\eta$]
-  {0\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
-  {0\Gamma \vdash (\mathrm{fst}(M), \mathrm{snd}(M)) \equiv M \stackrel0: (x \stackrel\pi: S) \otimes T}
-
-  \inferrule*
-  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
-    0\Gamma \vdash M \stackrel0: (x \stackrel\pi: A) \otimes B \\
-    0\Gamma, x \stackrel0: A, y \stackrel0: B \vdash N \stackrel0: C[(x,y)/z]}
-  {0\Gamma \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \equiv N[\mathrm{fst}(M)/x,\mathrm{snd}(M)/y] \stackrel0: C[M/z]}
-\end{mathpar}
-
-\subsection{Identity Type} Type formation, introduction, reflection
-and $\eta$-law:
-\begin{mathpar}
-  \inferrule* [right=Ty-Id]
-  {0\Gamma \vdash S~\istype \\
-    0\Gamma \vdash M \stackrel0: S \\
-    0\Gamma \vdash N \stackrel0: S}
-  {0\Gamma \vdash M =_S N~\istype}
-
-  \inferrule* [right=Id-Refl]
-  {\Gamma \vdash M \stackrel\sigma: S}
-  {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
-
-  \inferrule* [right=Id-Reflect]
-  {\Gamma \vdash N \stackrel0: M_1 =_S M_2}
-  {\Gamma \vdash M_1 \equiv M_2 \stackrel0: S}
-
-  \inferrule* [right=Id-Uniq]
-  {0\Gamma \vdash P \stackrel0: M =_A M}
-  {0\Gamma \vdash P \equiv \mathrm{refl}(M) : M=_A M}
-\end{mathpar}
-
-\subsection{Universe}
-
-\newcommand{\TySet}{\mathsf{U}}
-\newcommand{\TyEl}{\mathsf{El}}
-Type formation:
-\begin{displaymath}
-  \inferrule* [right=Ty-U]
-  {0\Gamma~\isctxt}
-  {0\Gamma \vdash \TySet~\istype}
-\end{displaymath}
-Introduction (also with introduction rules for all other type formers except $\TySet$):
-\begin{mathpar}
-  \inferrule* [right=Tm-U-Bool]
-  {\Gamma~\isctxt}
-  {\Gamma \vdash \BoolTy \stackrel\sigma: \TySet}
-
-  \inferrule* [right=Tm-U-Tensor]
-  {\Gamma \vdash M \stackrel\sigma: \TySet \\
-    \Gamma, x \stackrel\rho: \TyEl(M) \vdash N \stackrel\sigma: \TySet}
-  {\Gamma \vdash (x \stackrel\pi: M) \otimes N \stackrel\sigma: \TySet}
-
-  \ldots
-\end{mathpar}
-Elimination:
-\begin{displaymath}
-  \inferrule* [right=Ty-El]
-  {0\Gamma \vdash M \stackrel0: \TySet}
-  {0\Gamma \vdash \TyEl(M)~\istype}
-\end{displaymath}
-Equality:
-\begin{mathpar}
-  % \inferrule* [right=Ty-Eq-El-Pi]
-  % {0\Gamma \vdash M \stackrel0: \TySet \\
-  %   0\Gamma, x \stackrel0: \TyEl(M) \vdash N \stackrel0: \TySet}
-  % {0\Gamma \vdash \TyEl((x \stackrel\rho: M) \to N) \equiv (x \stackrel\rho: \TyEl(M)) \to \TyEl(N)~\istype}
-
-  % \inferrule* [right=Ty-Eq-El-Bool]
-  % {0\Gamma~\isctxt}
-  % {0\Gamma \vdash \TyEl(\BoolTy) \equiv \BoolTy~\istype}
-
-  % \ldots
-
-  \inferrule* [right=Ty-Eq-El-Cong]
-  {0\Gamma \vdash M \equiv N \stackrel0: \TySet}
-  {0\Gamma \vdash \TyEl(M) \equiv \TyEl(N)~\istype}
-\end{mathpar}
-
-\subsection{Booleans}
-Formation, introduction, and elimination:
-\begin{mathpar}
-  \inferrule*
-  {\Gamma~\isctxt}
-  {\Gamma \vdash \BoolTy~\istype}
-
-  \inferrule*
-  {\Gamma~\isctxt}
-  {0\Gamma \vdash \cTrue, \cFalse \stackrel\sigma: \BoolTy}
-
-  \inferrule*
-  {0\Gamma_1, x \stackrel0: \BoolTy \vdash P~\istype \\
-    \Gamma_1 \vdash M \stackrel\sigma: \BoolTy \\
-    \Gamma_2 \vdash N_t \stackrel\sigma: P[\cTrue/x] \\
-    \Gamma_2 \vdash N_f \stackrel\sigma: P[\cFalse/x] \\
-    0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1 + \Gamma_2 \vdash \If_{x.P}\: M \: \Then \: N_t \: \Else \: N_f \stackrel\sigma: P[M/x]}
-\end{mathpar}
-$\beta$-equalities:
-\begin{mathpar}
-  \inferrule* [right=Tm-Eq-True$\beta$]
-  {0\Gamma, z \stackrel0: \BoolTy \vdash P \\
-    \Gamma \vdash M_t \stackrel\sigma: P[\mathrm{true}/z] \\
-    \Gamma \vdash M_f \stackrel\sigma: P[\mathrm{false}/z]}
-  {\Gamma \vdash \If_{x.P}\: \cTrue \: \Then \: N_t \: \Else \: N_f \equiv N_t \stackrel\sigma: P[\mathrm{true}/z]}
-
-  \inferrule* [right=Tm-Eq-False$\beta$]
-  {0\Gamma, z \stackrel0: \BoolTy \vdash P \\
-    \Gamma \vdash M_t \stackrel\sigma: P[\mathrm{true}/z] \\
-    \Gamma \vdash M_f \stackrel\sigma: P[\mathrm{false}/z]}
-  {\Gamma \vdash \If_{x.P}\: \cFalse \: \Then \: N_t \: \Else \: N_f \equiv N_f \stackrel\sigma: P[\mathrm{false}/z]}
-\end{mathpar}
-
-\subsection{Lists} Formation and introduction:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma \vdash T~\istype}
-  {0\Gamma \vdash \ListTy(T)~\istype}
-
-  \inferrule*
-  {\Gamma \vdash T~\istype}
-  {0\Gamma \vdash \cNil \stackrel\sigma: \ListTy(T)}
-
-  \inferrule*
-  {\Gamma_1 \vdash M \stackrel\sigma: T \\
-    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
-    0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
-\end{mathpar}
-Case analysis:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
-    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
-    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
-    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
-    0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
-\end{mathpar}
-$\beta$-equalities for case analysis:
-\begin{mathpar}
-  \inferrule* [right=Eq-List-Match-Nil]
-  {0\Gamma, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
-    \Gamma \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
-    \Gamma, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x]}
-  {\Gamma \vdash \Match_{x.P}\,\cNil\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \equiv N_1 \stackrel\sigma: P[\cNil/x]}
-
-  \inferrule* [right=Eq-List-Match-Cons]
-  {0\Gamma_1, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
-    \Gamma_1 \vdash M_1 \stackrel\sigma: T \\
-    \Gamma_2 \vdash M_2 \stackrel\sigma: \ListTy(T) \\
-    \Gamma_3 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
-    \Gamma_3, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
-    0\Gamma_1 = 0\Gamma_2 = 0\Gamma_3}
-  {\Gamma_1 + \Gamma_2 + \Gamma_3 \vdash {\begin{array}[m]{@{}l}
-                                            \Match_{x.P}\,(\cCons(M_1,M_2))\\
-                                            \qquad\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\}\\
-                                            \equiv N_2[M_1/h,M_2/t] \end{array}} \stackrel\sigma: P[\cCons(M_1,M_2)/x]}
-\end{mathpar}
-$\sigma = 0$ recursive eliminator:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
-    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
-    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
-    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
-  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
-\end{mathpar}
-$\beta$-equalities for recursion:
-\begin{mathpar}
-  \inferrule* [right=Eq-List-Rec-Nil]
-  {0\Gamma, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
-    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
-    0\Gamma, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
-  {0\Gamma \vdash \mathrm{recList}_{x.P}\,\cNil\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \equiv N_1 \stackrel0: P[\cNil/x]}
-
-  \inferrule* [right=Eq-List-Rec-Cons]
-  {0\Gamma, x : \stackrel0: \ListTy(T) \vdash P~\istype \\
-    0\Gamma \vdash M_1 \stackrel0: T \\
-    0\Gamma \vdash M_2 \stackrel0: \ListTy(T) \\
-    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
-    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T) \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
-  {0\Gamma \vdash {\begin{array}[m]{@{}l}
-                     \mathrm{recList}_{x.P}\,(\cCons(M_1,M_2))\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\}\\
-                     \equiv N_2[\begin{array}[t]{@{}l} M_1/h,M_2/t,\\
-                                  \mathrm{recList}_{x,p}\,M_2\,\{\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\}/p]\end{array} \end{array}} \stackrel0: P[\cCons(M_1,M_2)/x]}
-\end{mathpar}
-
-\subsection{Cons-Free Naturals} Formation and introduction:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma~\isctxt}
-  {0\Gamma \vdash \tyNat~\istype}
-
-  \inferrule*
-  {\Gamma~\isctxt}
-  {\Gamma \vdash \conZero \stackrel0: \tyNat}
-
-  \inferrule*
-  {\Gamma \vdash M \stackrel0: \tyNat}
-  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
-\end{mathpar}
-Duplication:
-\begin{mathpar}
-  \inferrule*
-  {\Gamma \vdash M \stackrel\sigma: \tyNat}
-  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
-
-  \inferrule*
-  {\Gamma \vdash M \stackrel0: \tyNat}
-  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
-\end{mathpar}
-Eliminator:
-\begin{displaymath}
-  \mprset{flushleft}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
-    0\Gamma \vdash N_z \stackrel\sigma: P[\conZero/x] \\\\
-    0\Gamma, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma: P[\conSucc(n)/x]}
-  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
-\end{displaymath}
-$\beta$-equalties (only available in the $\sigma = 0$ fragment):
-\begin{mathpar}
-  \mprset{flushleft}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    0\Gamma \vdash N_z \stackrel0: P[\conZero/x] \\\\
-    0\Gamma, n \stackrel0: \tyNat, p \stackrel0: P[n/x] \vdash N_s \stackrel0: P[\conSucc(n)/x]}
-  {0\Gamma \vdash \tmRec_{x.P}\,\conZero\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \equiv N_z \stackrel0: P[\conZero/x]}
-
-  \mprset{flushleft}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    0\Gamma \vdash M \stackrel0: \tyNat \\\\
-    0\Gamma \vdash N_z \stackrel0: P[\conZero/x] \\\\
-    0\Gamma, n \stackrel0: \tyNat, p \stackrel0: P[n/x] \vdash N_s \stackrel0: P[\conSucc(n)/x]}
-  {0\Gamma \vdash {\begin{array}{@{}l}
-                    \tmRec_{x.P}\,(\conSucc(M))\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \\
-                    \equiv N_s[M/n,\tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\}/p] \end{array}} \stackrel0: P[\conSucc(M)/x]}
-\end{mathpar}
-
-\subsection{LFPL Diamonds} Formation, introduction, and $\eta$-law:
-\begin{mathpar}
-  \inferrule*
-  {\Gamma~\isctxt}
-  {0\Gamma \vdash \Diamond~\istype}
-
-  \inferrule*
-  {\Gamma~\isctxt}
-  {0\Gamma \vdash * \stackrel0: \Diamond}
-
-  \inferrule*
-  {\Gamma \vdash M \stackrel0: \Diamond}
-  {\Gamma \vdash M \equiv * \stackrel0: \Diamond}
-\end{mathpar}
-
-\subsection{LFPL Naturals} Type formation and introduction:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma~\isctxt}
-  {0\Gamma \vdash \tyNat~\istype}
-
-  \inferrule*
-  {\Gamma \vdash M \stackrel\sigma: \Diamond}
-  {\Gamma \vdash \conZero(M) \stackrel\sigma: \tyNat}
-
-  \inferrule*
-  {\Gamma_1 \vdash M \stackrel\sigma: \Diamond \\
-    \Gamma_2 \vdash N \stackrel\sigma: \tyNat \\
-    0\Gamma_1 = 0\Gamma_2}
-  {\Gamma_1 + \Gamma_2 \vdash \conSucc(M,N) \stackrel\sigma: \tyNat}
-\end{mathpar}
-Elimination:
-\begin{mathpar}
-  \mprset{flushleft}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
-    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
-    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
-  {\Gamma \vdash \tmRec\,M\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
-\end{mathpar}
-$\beta$-equalities, available in both fragments:
-\begin{mathpar}
-  \mprset{flushleft}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    \Gamma \vdash M \stackrel\sigma: \Diamond \\\\
-    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
-    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
-  {\Gamma \vdash \tmRec\,\conZero(M)\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \equiv N_z[M/d] \stackrel\sigma: P[\conZero(*)/x]}
-
-  \mprset{flushleft}
-  \inferrule*
-  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
-    \Gamma_1 \vdash M_d \stackrel\sigma: \Diamond \\\\
-    \Gamma_2 \vdash M_n \stackrel\sigma: \tyNat \\\\
-    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
-    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x] \\\\
-    0\Gamma_1 = 0\Gamma_2 = 0\Gamma}
-  {\Gamma_1 + \Gamma_2 \vdash {\begin{array}{@{}l}
-                                 \tmRec\,(\conSucc(M_d,M_n))\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
-                                 \equiv N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p] \end{array}} \stackrel\sigma: P[\conSucc(*,M_n)/x]}
-\end{mathpar}
-
-\subsection{Realisability Reflection} Type formation, introduction and
-elimination:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma \vdash A~\istype}
-  {0\Gamma \vdash \Rtype(A)~\istype}
-
-  \inferrule*
-  {0\Gamma \vdash M \stackrel1: A}
-  {0\Gamma \vdash \rIntro(M) \stackrel\sigma: \Rtype(A)}
-
-  \inferrule*
-  {\Gamma \vdash M \stackrel\sigma: \Rtype(A)}
-  {\Gamma \vdash \rElim(M) \stackrel{\sigma'}: A}
-\end{mathpar}
-Equalities:
-\begin{mathpar}
-  \inferrule*
-  {0\Gamma \vdash M \stackrel1: A}
-  {0\Gamma \vdash \rElim(\rIntro(M)) \equiv M \stackrel\sigma: A}
-
-  \inferrule*
-  {0\Gamma \vdash M \stackrel\sigma: \Rtype(A)}
-  {0\Gamma \vdash \rIntro(\rElim(M)) \equiv M \stackrel\sigma: A}
-\end{mathpar}
-
 \end{document}

### 2023-10-26 paper/LFPLIter.hs

@@ -1,0 +1,9 @@
+module LFPLIter where
+
+iter :: Integer -> Integer -> Integer
+iter 0 _ = 1
+iter k 0 = 0
+iter k n = iter (k-1) (n-1) + iter k (n-1)
+
+binom :: Integer -> Integer -> Integer
+binom k n = product [1..n] `div` (product [1..k] * product [1..n-k])

### 2023-11-14 paper/paper-extended.tex
[ paper ] extended and diff versions, final changes
[ paper ] ACM publishing stuff

--- 2023-10-26 paper/paper.tex
@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
+%\setcopyright{ACMUNKNOWN}
+
+\setcopyright{rightsretained}
+\acmDOI{10.1145/3632918}
+\acmYear{2024}
 \copyrightyear{2024}
-\acmYear{2024}
-\acmDOI{10.1145/1122445.1122456}
-
-%%
-%% These commands are for a JOURNAL article.
+\acmSubmissionID{popl24main-p453-p}
 \acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF}
-\acmArticle{1}
-\acmMonth{8}
+\acmVolume{8}
+\acmNumber{POPL}
+\acmArticle{76}
+\acmMonth{1}
+\received{2023-07-11}
+\received[accepted]{2023-11-07}
 
@@
 \title{Polynomial Time and Dependent Types}
+\subtitle{Extended Version}
 
@@
 
-\subsection{The Cons-free system}
+\subsection{The Cons-free System}
 \label{sec:cons-free-intro}
@@
 In this section, we give an overview of the rules of QTT. The full
-rules, including all equality rules, are presented in the online
-appendix for this paper \cite{atkey2023polydep-online}.
-
-%The full rules are presented in \autoref{sec:rules}.
-
-\subsubsection{Natural-number usages}
+rules, including all equality rules, are presented in
+\autoref{sec:rules}.
+
+\subsubsection{Natural-number Usages}
 
@@
 
-\subsubsection{Contexts, variables, and conversion}
+\subsubsection{Contexts, Variables, and Conversion}
 
@@
 
-\subsubsection{Datatypes} QTT, as we have presented it so far, has no
+\subsubsection{Data Types} QTT, as we have presented it so far, has no
 interesting base types to perform computation on. Following our
@@
 
-\subsection{Non-iterable Datatypes}
+\subsection{Non-iterable Data Types}
 \label{sec:noniter-qtt}
@@
 
-\subsection{Building Datatypes}
+\subsection{Building Data Types}
 \label{sec:prog-datatypes}
@@
     (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
-    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{2}{3}) \Leftrightarrow P\,a
+    \left((a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{2}{3}) \Leftrightarrow P\,a\right)
   \end{array}
@@
 (QCwFs). \citet{atkey18qtt} constructs QCwFs from certain Linear
-Combinatory Algebras (LCA), where terms in the $\sigma = 1$ fragment
+Combinatory Algebras (LCAs), where terms in the $\sigma = 1$ fragment
 are realised by elements of the LCA. However, there is a mistake in
@@
 
-Here, we fix the mistake of \citet{atkey18qtt}, and provide a more
+Here, we fix the mistake of \citet{atkey18qtt} and provide a more
 general construction of QCwFs in terms of indexed linear preorders. We
@@
 
+\input{appendix-rules.tex}
+
 \end{document}

### 2023-11-14 paper/paper.tex
[ paper ] sync short and extended versions

@@
 \title{Polynomial Time and Dependent Types}
+% \subtitle{Extended Version}
 
@@
 In this section, we give an overview of the rules of QTT. The full
-rules, including all equality rules, are presented in the online
-appendix for this paper \cite{atkey2023polydep-online}.
-
-%The full rules are presented in \autoref{sec:rules}.
-
-\subsubsection{Natural-number usages}
+rules, including all equality rules, are presented in
+%\autoref{sec:rules}.
+the online appendix \cite{atkey2023polydep-online}.
+
+\subsubsection{Natural-number Usages}
 
@@
 
+%\input{appendix-rules.tex}
+
 \end{document}

### 2023-11-14 paper/paper-extended.tex
[ paper ] sync short and extended versions
[ paper ] edits

@@
   can also prove properties of these programs. A simple example is
-  proving that a sorting program actually sorts. Using QTT reflection,
+  proving that a polytime sorting program actually sorts. Using QTT reflection,
   we can go further and represent the class of polytime problems, with
@@
   Probabilistic Polynomial (BPP) time. Since these classes rely on
-  specific correctness criteria, it is not possible to capture them in
+  specific semantic correctness criteria, it is not possible to capture them in
   a simply typed system for polytime.
 \item We prove the polytime soundness of our systems via a
-  realisability argument in \autoref{sec:soundness}. Our construction
-  is an extension of the amortised complexity realisability
-  constructions of \citet{dallago11realisability}. We extend their
-  work to our dependently typed setting, and also give a realisability
+  realisability argument in \autoref{sec:soundness} and
+  \autoref{sec:realising-iteration}. Our construction is an extension
+  of the amortised complexity realisability constructions of
+  \citet{dallago11realisability}. We extend their work to our
+  dependently typed setting, and also give a realisability
   interpretation of datatypes directly, instead of via second-order
-  impredicative encodings. The technical content of this section has
+  impredicative encodings. The technical content of these sections has
   been formalised in the Agda proof assistant
-  \cite{norell2008dependently}, and is included in the supplementary
-  material.
+  \cite{norell2008dependently}, and is included in the associated
+  artefact \cite{atkey2023polydep-artefact}.
 \end{enumerate}
@@
 It is not too difficult to see that reduction of linear
-$\lambda$-terms always takes a number of steps linearly proprotional
+$\lambda$-terms always takes a number of steps linearly proportional
 to the size of the term. This is because every $\beta$-redex
@@
 complexity, of the system by adding datatypes that do not allow
-iteration. These can be used for representation, but not for driving
-computation. We include the rules here, to show how linearity must be
-preserved in these rules, and to foreshadow their dependently typed
+iteration. These can be used for representation but not for driving
+computation. We include the rules here to show how linearity must be
+preserved in these rules and to foreshadow their dependently typed
 counterparts in \autoref{sec:noniter-qtt}. The first type is the
-booleans, which are anyway non-recursive and so would not allow
+booleans, which are non-recursive and so would not allow
 iteration anyway:
@@
 on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
-Aelhig and Schwictenberg, this is sufficient because the binomials
+Aehlig and Schwictenberg, this is sufficient because the binomials
 form a basis for the vector space of all polynomials.
@@
 \autoref{sec:rules}.
+%the online appendix \cite{atkey2023polydep-online}.
 
@@
 that they do not have a strict separation of variables include those
-of \citet{MoonEO21}, \citet{ChoudhuryEEW21}, and \citet{Abel18}. These
+of \citet{MoonEO21}, \citet{ChoudhuryEEW21}, and \citet{Abel23}. These
 systems differ from QTT in that they do not include a complete copy of
@@
 
+\section*{Data Availability Statement}
+
+The Agda source files and rendered HTML for this paper is available from Zenodo \cite{atkey2023polydep-artefact}.
+The source files are also available online at GitHub: \url{https://github.com/bobatkey/qtt-models}, commit SHA1
+
 %%

### 2023-11-14 paper/paper-extended.tex
[ paper ] final edits

@@
 \newcommand{\op}{\mathsf{op}}
+
+\newcommand{\ConsFree}{\emph{Cons-free}}
 
@@
 repeated multiplication). We therefore refer to this system as the
-\emph{Cons-Free} system.  Because we cannot construct values of type
+\ConsFree{} system.  Because we cannot construct values of type
 $\tyNat$ within the system, complete programs in this system are open
@@
 polynomials, the function $f$ can now be iterated $p(n)$ many times,
-where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
+where $n$ is the input $\tyNat$. Thus, the \ConsFree{} system can
 represent all polytime computations.
@@
 
-The Cons-free system is sound and complete for polytime, but is quite
+The \ConsFree{} system is sound and complete for polytime, but is quite
 awkward from the point of view of functional programming. It allows us
@@
 iteration. Diamonds cannot be created by a program itself, for the
-same reason that constructors were prohibited in the \emph{Cons-free}
+same reason that constructors were prohibited in the \ConsFree{}
 system, but they are released from iterable data during iteration. The
@@
 \end{displaymath}
-The difference with the \emph{Cons-free} iterator above is that the
+The difference with the \ConsFree{} iterator above is that the
 $\conZero$ and $\conSucc$ cases now both have an additional binding of
@@
 \end{displaymath}
-Unlike in the \emph{Cons-free} system, this iterator does not raise
+Unlike in the \ConsFree{} system, this iterator does not raise
 the degree of the nested iterator directly. Rather, the iterator $I_k$
@@
 
-We have now seen the \emph{Cons-free} and \emph{LFPL} systems for
+We have now seen the \ConsFree{} and LFPL systems for
 capturing polytime by means of linear typing and restricted
@@
 datatypes that allow construction and case analysis, but no
-recursion. In \autoref{sec:cons-free-qtt} we describe how to extend
-QTT to be a dependently typed adaptation of the Cons-free system of
-\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
-same treatment to the LFPL-style system.
+recursion. Then, in \autoref{sec:cons-free-qtt} we describe how to
+extend QTT to be a dependently typed adaptation of the \ConsFree{}
+system of \autoref{sec:cons-free-intro} by adding a type of iterable
+naturals. In \autoref{sec:lfpl-qtt} we apply the same treatment to
+the LFPL-style system.
 
@@
 Lists do have the potential for iteration by their recursive nature,
-but in order to ensure the complexity guarantees we only permit
-matching without recursion in the full calculus. Here is the rule for
+but in order to ensure the polytime complexity guarantees we only permit
+matching without recursion in the $\sigma = 1$ fragment. Here is the rule for
 dependently typed case analysis on lists, which also obeys the usual
 $\beta$-equalities for case analysis, analogous to the ones for
-Booleans:
+booleans:
 \begin{mathpar}
@@
 $\sigma = 0$ fragment recursor for lists has the following typing
-rule, which is the standard dependent eliminator for lists, with
+rule, which is the standard dependent eliminator for lists except that
 everything annotated as $0$ usage.
@@
 work proportional to the size of their input, we need some form of
-iterable datatype. In both our Cons-free and LFPL-style QTT systems,
+iterable datatype. In both our \ConsFree{} and LFPL-style QTT systems,
 we use a natural number datatype.
 
-The cons-free system cannot allow the programmer to construct natural
+The \ConsFree{} system cannot allow the programmer to construct natural
 numbers in the $\sigma = 1$ fragment, as this would violate the
@@
 adding an equational rule to the $\sigma = 0$ fragment, ensuring
-\emph{definitionally} that $\mathrm{dupNat}$ acts as its name implies:
+definitionally that $\mathrm{dupNat}$ acts as its name implies:
 \begin{mathpar}
@@
 As explained in \autoref{sec:lfpl-intro}, the LFPL system differs from
-the Cons-free system in that it is possible to construct natural
+the \ConsFree{} system in that it is possible to construct natural
 numbers (and other iterable datatypes), provided one has the necessary
 diamonds to pay for the construction. As with the natural number type
-in the Cons-free system, it ought not be possible to construct
+in the \ConsFree{} system, it ought not be possible to construct
 diamonds in the $\sigma = 1$ fragment, as this would amount to the
@@
 and so construct natural numbers freely as well just as we did for the
-Cons-free system above.
+\ConsFree{} system above.
 
@@
 
-Unlike the natural number iterator in the Cons-free system, this
+Unlike the natural number iterator in the \ConsFree{} system, this
 iterator has $\beta$-equalities in both the $\sigma = 0$ and
@@
 Note that the fact that the variable $n$ in the $N_s$ term is
-annotated $0$, which allows us to duplicate it even when we are in the
-$\sigma = 1$ fragment.
-
-Just as for the Cons-free system iterator above, in the $\sigma = 0$
+annotated $0$, which allows us to use $M_n$ twice even when we are in
+the $\sigma = 1$ fragment.
+
+Just as for the \ConsFree{} system iterator above, in the $\sigma = 0$
 fragment this rule is identical to the usual dependently typed
@@
 our systems above. We could extend both systems to include further
-iterable inductive types, although in the \emph{Cons-free} system this
+iterable inductive types, although in the \ConsFree{} system this
 is not particularly useful due to the prohibition of
@@
   \mathrm{PTIME}(A,P) =
-    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
+    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes \left((a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a\right)
 \end{displaymath}
@@
 \begin{displaymath}
-  (A,P) \stackrel{\textsc{Poly}}\Rightarrow (B,Q) = (f \stackrel1: \mathbf{R}(A \to B)) \otimes (a \stackrel1: A) \to Q(\mathbf{R}^{-1}(f)\,a) \Leftrightarrow P\,a
+  (A,P) \stackrel{\textsc{Poly}}\Rightarrow (B,Q) = (f \stackrel1: \mathbf{R}(A \to B)) \otimes \left((a \stackrel1: A) \to Q(\mathbf{R}^{-1}(f)\,a) \Leftrightarrow P\,a\right)
 \end{displaymath}
@@
 well known in dependent type theory \cite{AbbottAG05}, and speaks to a
-general conception of containers as ``problem/solution'' pairings.
+general conception of containers as ``problem/solution'' pairings and
+container morphisms as problem reductions.
 
@@
 polytime. We do this by augmenting our polytime functions with
-additional power in the form of computation effects.
+additional power in the form of computational effects.
 
@@
 to reason about equality of these non-deterministic programs, so we
-can represent non-deterministic choices as \emph{additive} binary
-trees. We suppose a non-iterable datatype defined like so:
+can represent non-deterministic choices as binary trees. We suppose a
+non-iterable datatype defined like so:
 \begin{displaymath}
@@
 (see the encoding of the additive product types by
-\citet{atkey18qtt}). Thus, each branch of this tree is polynomial
-time, not the whole tree itself.
+\citet{atkey18qtt}). Thus, each branch of this tree can be explored in
+polynomial time, but not the whole tree itself.
 
@@
 have difficulty in directly capturing the class BPP due to its
-semantic nature \cite{AroraBarak09}, where the correctness of
-implementation is probabilistic. With a dependently-typed host
-language, adding probabilistic choice as an effect and capturing the
-semantic constraint of BPP is straightforward.
+semantic nature, where the correctness of implementation is
+probabilistic. With a dependently-typed host language, adding
+probabilistic choice as an effect and capturing the semantic
+constraint of BPP is straightforward.
 
@@
 polymorphic encodings and special $\oc$-style modalities. These
-explicit constructions are essential to construct a model of dependent
-types.
+explicit constructions are essential to construct models of our systems.
 
@@
 \cite{norell2008dependently}. The Agda formalisation can be found in
-the supplementary material. After each definition and result we
-provide a pointer to the Agda modules where the corresponding
-formalisation can be found, and note interesting features of the
-mechanisation.
+the associated artefact \cite{atkey2023polydep-artefact}. After each
+definition and result we provide a pointer to the Agda modules where
+the corresponding formalisation can be found, and note interesting
+features of the mechanisation.
 
@@
 remaining potential yields $p(m)$ available steps. The $\MaxPoly$ and
-$\PlusPoly$ resource monoids will be used for the \emph{Cons-free} and
+$\PlusPoly$ resource monoids will be used for the \ConsFree{} and
 LFPL-style systems respectively, as we explain in
@@
 reformulation of \citet{dallago11realisability}'s realisability models
-suitable for modelling dependent types. For a set $A$, the carrier of
+to make it suitable for dependent types. For a set $A$, the carrier of
 $L(A)$ is the set of ternary relations
@@
 and \autoref{sec:noniter-qtt}. Booleans are the simplest case, with
-only two cases and no chance of iteration. Lists are more complex; we
+only two cases and no chance of iteration. Lists are more complex: we
 can have non-iterable lists containing iterable data.
@@
 constant-time programs to be realised. To interpret the iterators of
-the \emph{Cons-free} and LFPL-style systems, we need to use the
+the \ConsFree{} and LFPL-style systems, we need to use the
 $\MaxPoly$ and $\PlusPoly$ resource monoids. We do this in this
@@
 of LFPL), we need to construct specific realisers for the
-\emph{Cons-free} and LFPL systems.
+\ConsFree{} and LFPL systems.
 
@@
 
-The \emph{Cons-free} system uses the $\MaxPoly$ resource monoid, with
+The \ConsFree{} system uses the $\MaxPoly$ resource monoid, with
 the distinguished sub-monoid being those elements that are $0$ in the
-size component. We enumerate the features of the \emph{Cons-free}
+size component. We enumerate the features of the \ConsFree{}
 system and justify their realisability with the $\MaxPoly$ resource
@@
 \end{enumerate}
-Together, we have a soundness result for the \emph{Cons-free} system,
+Together, we have a soundness result for the \ConsFree{} system,
 that ensures that every term in the $\sigma = 1$ fragment is
@@
 time for all inputs:
-\begin{theorem}[Soundness for the Cons-free System]
+\begin{theorem}[Soundness for the \ConsFree{} System]
   \label{thm:cons-free-soundness}
-  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
+  If we have a term $n \stackrel1: \tyNat \vdash M \stackrel1: T(n)$ then there
   exists a realising expression $E$ and polynomial $p$ such that for
@@
 \paragraph{Agda Formalisation} The realisability of the
-\emph{Cons-free} system iterator and the soundness property of the
+\ConsFree{} system iterator and the soundness property of the
 whole system are formalised in the Agda modules \AgdaModule{ConsFree}
@@
   $n+2$ for a new number. Note that, even if we add a $\Diamond$ type
-  to the \emph{Cons-free} system, it would still not be possible to
+  to the \ConsFree{} system, it would still not be possible to
   realise the constructors, because we would only have
@@
 \item The construction of the recursor follows a very similar proof to
-  the realisability of \emph{Cons-free} iterator, up to some
+  the realisability of \ConsFree{} iterator, up to some
   additional work to make sure that the dummy $*$ values representing
@@
 
-Soundness for the LFPL system is similar to the \emph{Cons-free}
+Soundness for the LFPL system is similar to the \ConsFree{}
 system, except for a $+1$ to the input to the polynomial, to account
@@
   \label{thm:lfpl-soundness}
-  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
+  If we have a term $n \stackrel1: \tyNat \vdash M \stackrel1: T(n)$ then there
   exists a realising expression $E$ and polynomial $p$ such that for
@@
 The proofs of well-accounted realisability for the LFPL iterator, and
-the \emph{Cons-free} iterator, could be adapted to any other
+the \ConsFree{} iterator, could be adapted to any other
 inductively defined type that is finitely branching. This is not
@@
 We have presented two extensions of Quantitative Type Theory that
-soundly and completely capture polynomial time. We have seen that this
-allows for an expressive combination of verification and complexity
-constrained computation. We now discuss some related work in implicit
-computational complexity and linear dependent types, and take a look
-at where the combination of resource conciousness and dependent types
-could take us.
+soundly and completely capture polynomial time. This allows for an
+expressive combination of verification and complexity constrained
+computation, including characterisations of the classes P, NP, and
+BPP. We now discuss related work, and take a look at where the
+combination of polytime and dependency could take us.
 
@@
 functional programs without constructors. Thus it shares a method with
-our \emph{Cons-free} system, but we use linear typing to permit
+our \ConsFree{} system, but we use linear typing to permit
 controlled use of higher-order functions. Other approaches to
@@
 \citet{RajaniG0021}, both of which are based on ideas of type-based
-amortised complexity analysis arising from \cite{hofmann99lfpl}'s
+amortised complexity analysis arising from \citet{hofmann99lfpl}'s
 ideas, via the work of \citet{HofmannJ03}. More details are to be
@@
 correctly annotating the program with tick effects to count the
-resource usage they are interested in, in contrast to the intrinsic
-approach we have taken here.
+resource usage they are interested in, in contrast our intrinsic
+approach.
 
@@
 
-\paragraph{Other Complexity Classes, and Explicit Resource Tracking}
+\paragraph{Other Complexity Classes}
 We have been able to characterise the classes NP and BPP in terms of
@@
 
+\paragraph{Explicit Resource Tracking}
 Our construction already includes soundness of a system with intrinsic
@@
 
-\paragraph{Towards Synthetic Computational Complexity Theory?}
+\paragraph{Towards a Synthetic Computational Complexity Theory?}
 
@@
 prove the Cook-Levin theorem that 3-SAT is NP-complete. This is
-because the proofs relies on obtaining the \emph{source code} of the
+because the proof relies on obtaining the \emph{source code} of the
 program solving an NP problem and then encoding that program in
@@
 \emph{Synthetic Computational Complexity Theory}, analogous to the
-Church-Turing axiom for Synthetic Computability Theory \cite{Bauer06}.
+Church-Turing axiom for Synthetic Computability Theory as described by
+\citet{Bauer06}.
 
@@
 
-The Agda source files and rendered HTML for this paper is available from Zenodo \cite{atkey2023polydep-artefact}.
-The source files are also available online at GitHub: \url{https://github.com/bobatkey/qtt-models}, commit SHA1
+The Agda source files and rendered HTML for this paper is available
+from Zenodo \cite{atkey2023polydep-artefact}. The source files are
+also available online at GitHub:
+\url{https://github.com/bobatkey/qtt-models}.
 

### 2023-11-14 paper/paper.tex
[ paper ] final edits

@@
 \title{Polynomial Time and Dependent Types}
-% \subtitle{Extended Version}
+%\subtitle{Extended Version}
 
@@
 \end{displaymath}
-Unlike in the \emph{Cons-free} system, this iterator does not raise
+Unlike in the \ConsFree{} system, this iterator does not raise
 the degree of the nested iterator directly. Rather, the iterator $I_k$
 on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
-Aelhig and Schwictenberg, this is sufficient because the binomials
+Aehlig and Schwictenberg, this is sufficient because the binomials
 form a basis for the vector space of all polynomials.
@@
 %\autoref{sec:rules}.
-the online appendix \cite{atkey2023polydep-online}.
+the exteneded version of this paper \cite{atkey2023polydep-online}.
 
@@
 
+\section*{Data Availability Statement}
+
+The Agda source files and rendered HTML for this paper is available
+from Zenodo \cite{atkey2023polydep-artefact}. The source files are
+also available online at GitHub:
+\url{https://github.com/bobatkey/qtt-models}.
+
 %%

### 2023-12-14 paper/paper.tex
[ paper ] Conference Publishing's demands

@@
 
-\subsection{Contributions and Content}
+\paragraph{Contributions and Content}
 
@@
 
-\subsection{The Cons-free System}
+\subsection{The Cons-Free System}
 \label{sec:cons-free-intro}
@@
 
-\subsubsection{Natural-number Usages}
+\subsubsection{Natural-Number Usages}
 
@@
 
-\subsection{Non-iterable Data Types}
+\subsection{Non-Iterable Data Types}
 \label{sec:noniter-qtt}
@@
 
-\subsection{Cons-free Natural Numbers and their Recursor}
+\subsection{Cons-Free Natural Numbers and their Recursor}
 \label{sec:cons-free-qtt}
@@
 
-\subsection{LFPL-style Diamonds, Natural Numbers, and a Recursor that Gives Back}
+\subsection{LFPL-Style Diamonds, Natural Numbers, and a Recursor that Gives Back}
 \label{sec:lfpl-qtt}
@@
 
-\subsection{Polytime-based Complexity Classes}
+\subsection{Polytime-Based Complexity Classes}
 \label{sec:classes}
@@
 
-\subsubsection{Bounded-error Probabilistic Polynomial Time}
+\subsubsection{Bounded-Error Probabilistic Polynomial Time}
 
@@
 polymorphic encodings and special $\oc$-style modalities. These
-explicit constructions are essential to construct models of our systems.
+explicit constructions are essential for constructing models of our systems.
 
@@
 the number of steps. For simplicity, all operations cost $1$ unit,
-though this could be generalised to allow for different operations to
-have different costs. We use $\eta[i]$ to access the $i$th variable in
-the environment, counting from the right. The evaluation rules are
-mostly as one would expect, except for the application rule which
-includes a self reference to the closure being invoked, in order to
+though this could be generalised. We use $\eta[i]$ to access the $i$th
+variable in the environment, counting from the right. The evaluation
+rules are mostly as one would expect, except that the application rule
+includes a self reference to the closure being invoked in order to
 allow recursive definitions.
@@
 Agda module \AgdaModule{MachineModel}. We use an intrinsically
-well-scoped syntax, which ensures that all variable look up operations
-are always well defined.
+well-scoped syntax, which ensures that all variable accesses are well
+defined.
 
@@
 
-\subsubsection{Resource sub-monoids}
+\subsubsection{Resource Sub-Monoids}
 
@@
 
-\subsubsection{Non-iterable Data Types}
+\subsubsection{Non-Iterable Data Types}
 \label{sec:noniter-model}
